### 2009-09-28 papers/esop2010/main.tex
esop paper

\documentclass[a4paper]{llncs}

\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc} 
\input{utf8symbols}
\usepackage{stmaryrd}
\usepackage{amsmath,amssymb}
\usepackage{mathpartir}

\newcommand{\wpre}{ {\bf wp}}
\newcommand{\sep}{ ~|~ }
\newcommand{\letml}{{\bf let}}
\newcommand{\inml}{{\bf in}}
\newcommand{\ifml}{{\bf if}}
\newcommand{\thenml}{{\bf then}}
\newcommand{\elseml}{{\bf else}}
\newcommand{\refml}{{\bf ref}}
\newcommand{\propml}{{\bf prop}}
\newcommand{\boolml}{{\bf bool}}
\newcommand{\recml}{{\bf rec}}
\newcommand{\efft}[1]{ \langle #1 \rangle }
\newcommand{\alist}[1]{ \overline{#1} }

\begin{document}
\title{\mbox{Specifying Imperative Higher Order Programs}}

\author{Johannes Kanig \and Jean-Christophe Fill\^atre}
\institute{
Laboratoire de Recherche en Informatique, Univ Paris-Sud, CNRS, Orsay F-91405\\
INRIA Futurs, ProVal, Parc Orsay Universit\'e, F-91893
}
%TODO affiliation

\maketitle

\begin{abstract}
  %TODO
\end{abstract}

\section{Introduction.}
%TODO Hoare logic = sharing between logic and prog language + alias free

\section{A programming language for specification.}

\subsection{Syntax}
\begin{figure}[htpb]
\begin{eqnarray*}
  x & & \text{Program Variables}\\
  α & & \text{Type Variables}\\
  e & & \text{Effect Variables}\\
  ρ & & \text{Region Variables}\\
  χ &::=& α \sep e \sep ρ \\
  t,p,q &::=& x~[\alist{κ}] \sep λx:τ.t \sep \recml~f~(x:τ).\{p\}t\{q\} \sep t~t\sep \\ 
  & & \letml~ x~[\alist{χ}] = t~\inml~t \sep \ifml~t~\thenml~t~\elseml~t\\
  τ &::=& α \sep τ -> τ \sep τ ->^ε τ \sep τ \times τ \sep τ~\refml_ρ \sep
  \efft{ε} \\
  ε &::=& (ρ \sep e)*\\
  κ &::=& τ \sep ε \sep ρ
\end{eqnarray*}
  \caption{Syntax}
  \label{fig:syntax}
\end{figure}

Our language is basically λ-calculus with \letml-bindings. For technical
reasons, the presentation is different from its usual form.
Fig.\ref{fig:syntax} describes the whole language.

Let us first look at the {\em types}. Type variables are of course types, as
well as tuples ($τ\times τ$) and pure functions ($τ -> τ$). There is a second
function type $τ->^ε τ$ for impure functions, annotated with an {\em effect}
$ε$. Effects are simply lists of {\em regions} $ρ$ and {\em effect variables}
$e$. The type expression $τ~\refml_ρ$ describes the type of references in
region $ρ$ containing objects of type $τ$. The type $\efft{ε}$ describes the
portion of the store represented by the effect expression $ε$.

The syntax of terms contains some differences to standard λ-calculus. The
construct $λx:τ.t$ serves to construct a pure function, while the construct
$\recml~f~(x:τ).\{p\}t\{q\}$ can be used to build impure, possibly recursive
functions. The terms $p$ and $q$ are the {\em precondition} and the {\em
postcondition} of the so defined anonymous function. This is of course
reminiscent of the syntax of Hoare Logic.

Type, region and effect variables must be explicitly generalized and
instantiated. For this purpose, we have defined the metavariable $χ$,
representing variables, and $κ$, representing the corresponding instantiation
objects types, regions and effects\footnote{We do not distinguish between
region variables and region constants; a region constant is simply a region
variable whose scope is the entire program.}. Now, it can be seen that
generalization takes place at \letml-bindings and instantiation at variable
occurrences.
%TODO describe the overline

%TODO : get rid of instantiations ??
\subsection{Typing}

\begin{figure}[htpb]
\begin{mathpar}
  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~\alist{κ} :
  τ[\alist{χ}|->\alist{κ}]} } \and
  { \inferrule*[Left=PureFun] {Γ,x : τ' |-_v t : τ} {Γ|-_v λx : τ'.t : τ' ->
  τ} } \and
  { \inferrule*[Left=Rec] 
  { Γ' = Γ,~f : τ' ->^ε τ,~ x : τ' \\ Γ' |- t : τ,ε \\ Γ', cur : \efft{ε} |- p : \propml \\ 
  Γ', old, cur : \efft{ε} |- q : τ -> \propml } 
      {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
  } \and
  { \inferrule*[Left=PureApp]
    {Γ |-_v : t1 : τ' -> τ \\ Γ|-_v : t2 : τ'}
    {Γ |-_v : t1 ~ t2 : τ}
  } \and
  { \inferrule*[Left=App]
    {Γ |-_v : t1 : τ' ->^ε τ \\ Γ|-_v : t2 : τ'}
    {Γ |- : t1 ~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=ite]
    { Γ |- _v : t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |-t3 : τ, ε2 }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Letv]
    { Γ, \alist{χ} |-_v : t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
    { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=Let]
    { Γ |- : t1 : τ', ε1 \\ Γ, x : τ' |- t2 : τ, ε2 }
    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε1  ε2 }
  }
\end{mathpar}
  \caption{The Typing relation}
  \label{fig:typing}
\end{figure}

Again in order to simplify future definitions, the definition typing relation
differs somewhat from the standard way (see Fig.\ref{fig:typing}). First of
all, we present the language in {\em A-normal form} (ANF), where all values
obtained by intermediate computations are named. However, our definition of
values is not syntactical: our ``semantic'' values are objects which are
either syntactic values (abstractions and variables) or are obtained by {\em
pure application}, the application which involves the pure arrow $->$.
Furthermore, the system is a type and effect system. There are in fact two
typing relations, the first one is of the form $Γ|-_v : t : τ$. In this case,
$t$ is a semantic value of type $τ$ and cannot have any effect. In the other
case, we have $Γ|- : t : τ, ε$, where $t$ is an expression of type $τ$ and
effect $ε$.

Effects can be produced only by function application.

predefined : \propml \boolml

unusual : 
* ANF 
* values defined by types
* two different types of arrows
* distinction bool and prop
* notion of substitution - correspondence χ κ
* TODO logic typing

missing : 
* value injection
* subeffecting

\section{A Simple Weakest Preconditions Calculus.}

\section{Extensions.}

subtyping: effects

\section{Implementation.}

\section{Examples.}

\section{Related Work.}

\section{Conclusion.}

\end{document}

### 2009-09-28 papers/esop2010/main.tex
little change

@@
 either syntactic values (abstractions and variables) or are obtained by {\em
-pure application}, the application which involves the pure arrow $->$.
+pure application}, the application which involves the pure arrow $τ->τ$.
 Furthermore, the system is a type and effect system. There are in fact two
@@
 
-Effects can be produced only by function application.
+Effects can be produced only by function application, where the impure arrow
+$τ->^ετ$ is involved.
 

### 2009-09-30 papers/esop2010/main.tex
go

@@
 \newcommand{\efft}[1]{ \langle #1 \rangle }
+\newcommand{\lift}[1]{ \langle #1 \rangle }
 \newcommand{\alist}[1]{ \overline{#1} }
+\newcommand{\Trueml}{{\bf True}}
+\newcommand{\Falseml}{{\bf False}}
+\newcommand{\trueml}{{\bf true}}
+\newcommand{\falseml}{{\bf false}}
+\newcommand{\coq}{Coq}
+
+\newcommand{\ceil}[1]{\lceil #1 \rceil}
 
@@
   { \inferrule*[Left=Rec] 
-  { Γ' = Γ,~f : τ' ->^ε τ,~ x : τ' \\ Γ' |- t : τ,ε \\ Γ', cur : \efft{ε} |- p : \propml \\ 
-  Γ', old, cur : \efft{ε} |- q : τ -> \propml } 
+  { Γ' = Γ,~f : τ' ->^ε τ,~ x : τ' \\ Γ' |- t : τ,ε \\ Γ', cur : \efft{ε} |-_l p : \propml \\ 
+  Γ', old, cur : \efft{ε} |-_l q : τ -> \propml } 
       {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
@@
   { \inferrule*[Left=PureApp]
-    {Γ |-_v : t1 : τ' -> τ \\ Γ|-_v : t2 : τ'}
-    {Γ |-_v : t1 ~ t2 : τ}
+    {Γ |-_v t1 : τ' -> τ \\ Γ|-_v t2 : τ'}
+    {Γ |-_v  t1 ~ t2 : τ}
   } \and
   { \inferrule*[Left=App]
-    {Γ |-_v : t1 : τ' ->^ε τ \\ Γ|-_v : t2 : τ'}
-    {Γ |- : t1 ~ t2 : τ, ε}
+    {Γ |-_v t1 : τ' ->^ε τ \\ Γ|-_v t2 : τ'}
+    {Γ |- t1 ~ t2 : τ, ε}
   } \and
   { \inferrule*[Left=ite]
-    { Γ |- _v : t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |-t3 : τ, ε2 }
+    { Γ |- _v t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |- t3 : τ, ε2 }
     { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
@@
   { \inferrule*[Left=Letv]
-    { Γ, \alist{χ} |-_v : t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
+    { Γ, \alist{χ} |-_v t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
     { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
@@
   { \inferrule*[Left=Let]
-    { Γ |- : t1 : τ', ε1 \\ Γ, x : τ' |- t2 : τ, ε2 }
+    { Γ |- t1 : τ', ε1 \\ Γ, x : τ' |- t2 : τ, ε2 }
     { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε1  ε2 }
+  } \and
+  { \inferrule*[Left=Value]
+  { Γ |-_v t : τ  }
+  { Γ |- t : τ,\emptyset  }
+  } \and
+  { \inferrule*[Left=Logic]
+    { Γ |-_v t : τ  }
+    { Γ |-_l t : \ceil{τ} }
   }
@@
   \label{fig:typing}
+\end{figure}
+
+\begin{figure}[htpb]
+  \begin{eqnarray*}
+    \ceil{α} &=& α\\
+    \ceil{τ\ref_ρ} &=& \ceil{τ}\ref_ρ\\
+    \ceil{τ -> τ'} &=& \ceil{τ} -> \ceil{τ'}\\
+    \ceil{τ \times τ'} &=& \ceil{τ} \times-> \ceil{τ'}\\
+    \ceil{\efft{ε}} &=& \efft{ε}\\
+    \ceil{τ ->^ε τ'} &=& (\ceil{τ} -> \efft{ε} -> \propml)\times (\ceil{τ} ->
+    \efft{ε} -> \efft{ε} -> \ceil{τ'} -> \propml)()
+  \end{eqnarray*}
+  \caption{Lifting types}
+  \label{fig:typelift}
 \end{figure}
@@
 Furthermore, the system is a type and effect system. There are in fact two
-typing relations, the first one is of the form $Γ|-_v : t : τ$. In this case,
-$t$ is a semantic value of type $τ$ and cannot have any effect. In the other
-case, we have $Γ|- : t : τ, ε$, where $t$ is an expression of type $τ$ and
-effect $ε$.
+typing relations, the first one is of the form $Γ|-_v  t : τ$. In this case,
+$t$ is a semantic value of type $τ$ and cannot have any effect. It can be seen
+that the set of terms described by $|-_v$ is the precisely the one obtained by
+combining syntactic values and pure application in a well-typed way.
 
-Effects can be produced only by function application, where the impure arrow
-$τ->^ετ$ is involved.
+In the other case, we have $Γ|- t : τ, ε$, where $t$ is an expression of
+type $τ$ and effect $ε$. Impure applications, \letml-expressions and
+\ifml-expressions can only be typed in this way. Effects can be produced only
+by function application, where the impure arrow $τ'->^ετ$ is involved. In this
+case, the effect of the application is precisely the effect $ε$ of the arrow,
+because the two subterms are semantic values and don't have any effect.
 
-predefined : \propml \boolml
+The typing relies on two predefined types, \propml\ and \boolml. \propml\ is
+the logical truth and contains the values \Trueml\ and \Falseml, while
+\boolml\ describes boolean (decidable) properties and contains \trueml\ and
+\falseml. The distinction between \propml\ and \boolml\ is the same one as in
+\coq, for example. Properties expressed in \propml\ are potentially
+undecidable and one cannot take decisions by testing their truth value. On the
+other hand, \boolml\ talks about decidable properties and so it makes sense to
+expect the predicate of the \ifml-statement to be of type \boolml.
 
@@
 * notion of substitution - correspondence χ κ
-* TODO logic typing
 
@@
 * subeffecting
+* logic typing
+
+%TODO is the LOGIC rule correct ??
 

### 2009-09-30 papers/esop2010/main.tex
go

@@
 \newcommand{\falseml}{{\bf false}}
+\newcommand{\unitml}{{\bf unit}}
 \newcommand{\coq}{Coq}
@@
 
+also one or two examples in the described language
+
+characteristics : 
+\begin{itemize}
+  \item Hoare logic
+  \item logic and program share values  
+  \item Type and effect system
+  \item Region typing to track references
+  \item the wp needs absence of aliasing to produce good POs
+\end{itemize}
+
 \section{A programming language for specification.}
+
+technical presentation:
+\begin{itemize}
+  \item ANF
+  \item values defined by typing 
+  \item  two different types of arrows
+  \item  distinction bool and prop
+  \item  notion of substitution - correspondence χ κ
+  \item semantics is standard (goes to appendix )
+\end{itemize}
 
@@
     \ceil{α} &=& α\\
-    \ceil{τ\ref_ρ} &=& \ceil{τ}\ref_ρ\\
+    \ceil{τ~\refml_ρ} &=& \ceil{τ}\refml_ρ\\
     \ceil{τ -> τ'} &=& \ceil{τ} -> \ceil{τ'}\\
@@
 Again in order to simplify future definitions, the definition typing relation
-differs somewhat from the standard way (see Fig.\ref{fig:typing}). First of
+differs somewhat from the standard way (see Fig.~\ref{fig:typing}). First of
 all, we present the language in {\em A-normal form} (ANF), where all values
@@
 unusual : 
-* ANF 
-* values defined by types
-* two different types of arrows
-* distinction bool and prop
-* notion of substitution - correspondence χ κ
-
 missing : 
@@
 
+\subsection{Sharing between logic and programs}
+
+To obtain sharing between
+\subsection{References}
+
+We haven't mentioned references yet, but all we need to support them is to add
+the three usual functions $ref, :=, !$ for reference creation, assignment and
+lookup, respectively, to the environment. In our system, these functions have
+the following types:
+\begin{eqnarray*}
+  ref &:& ∀αρ.α ->^ρ α~\refml_ρ\\
+  := &:& ∀αρ.α~\refml_ρ ->^\emptyset α ->^ρ \unitml\\
+  ! &:& ∀αρ. α~\refml_ρ ->^ρ α
+\end{eqnarray*}
+
+\subsection{Alias restriction}
+
+\begin{figure}[htpb]
+  \begin{mathpar}
+    {\inferrule*[Left=Var]
+      {Γ(x) = ∀\alist{χ}.τ \\ σ = [\alist{χ}|->\alist{κ}] \\ σ \sim τ }
+      {Γ |- x : τσ }
+    } \\
+    { \inferrule*[Left=LetRef, Right={ρ \textup{fresh}, $ρ\notin τ$}]
+      {Γ |-_v t1 : τ' \\ Γ, x : τ'~\refml_ρ |- t2 : τ }
+      {Γ |- \letml~x = ref~ t1 ~\inml~ t2 : τ}
+    }
+  \end{mathpar}
+  \caption{A modified and a new rule for aliasing}
+  \label{fig:aliasing}
+\end{figure}
+
 \section{A Simple Weakest Preconditions Calculus.}
+
+In the following, we use the metasymbol $v$ to denote well-typed semantic
+values, i.e. terms with a typing derivation $Γ|-_v v : τ$.
 

### 2009-10-01 papers/esop2010/main.tex
biblio, and a lot of text (7 pages)

@@
 \newcommand{\coq}{Coq}
+\newcommand{\who}{Who}
+\newcommand{\correct}[1]{ c(#1) }
 
@@
 %TODO affiliation
+%TODO find notions for:
+%  * ``semantic values'' v
+%  * ``store types'' <ε>
+%
 
@@
 
-\section{Introduction.}
+\section{Introduction}
 %TODO Hoare logic = sharing between logic and prog language + alias free
@@
 
-\section{A programming language for specification.}
+\section{A programming language for specification}
 
@@
 
-Our language is basically λ-calculus with \letml-bindings. For technical
-reasons, the presentation is different from its usual form.
-Fig.\ref{fig:syntax} describes the whole language.
+Our programming language (Fig.~\ref{fig:syntax}) is basically the λ-calculus
+with \letml-bindings, where there are two different λ-bindings. The first one
+($λx:τ.t$) describes a pure, non-recursive binding, and the second one
+($\recml~f~(x:τ).\{p\}t\{q\}$) describes recursive, potentially impure
+functions; $p$ and $q$ in this term are {\em specifications}, which are
+of the same syntax as programs, but are differently typed (typing will be
+discussed in the next section). The term $p$ corresponds to the {\em
+precondition}, $q$ to the {\em postcondition} of the function. The analogy to
+Hoare triples is of course obvious. As usual, there are also variables,
+\letml-bindings and an \ifml-\thenml-\elseml-construct. Variables on the type
+level can be generalized at \letml-boundaries and have to be instantiated when
+the corresponding program variable is used.
 
-Let us first look at the {\em types}. Type variables are of course types, as
-well as tuples ($τ\times τ$) and pure functions ($τ -> τ$). There is a second
-function type $τ->^ε τ$ for impure functions, annotated with an {\em effect}
-$ε$. Effects are simply lists of {\em regions} $ρ$ and {\em effect variables}
-$e$. The type expression $τ~\refml_ρ$ describes the type of references in
-region $ρ$ containing objects of type $τ$. The type $\efft{ε}$ describes the
-portion of the store represented by the effect expression $ε$.
-
-The syntax of terms contains some differences to standard λ-calculus. The
-construct $λx:τ.t$ serves to construct a pure function, while the construct
-$\recml~f~(x:τ).\{p\}t\{q\}$ can be used to build impure, possibly recursive
-functions. The terms $p$ and $q$ are the {\em precondition} and the {\em
-postcondition} of the so defined anonymous function. This is of course
-reminiscent of the syntax of Hoare Logic.
+Before we describe the syntax of types, let us first explain {\em regions} and
+{\em effects}. Regions are simply type level names for memory locations. An
+effect is a set of memory locations and {\em effect variables}. An effect
+variable stands for an unknown effect. Effects are used to specify
+of memory locations an expression can write. They are also used in function
+types to describe the memory locations which may be written if the function is
+called.
 
@@
 generalization takes place at \letml-bindings and instantiation at variable
-occurrences.
-%TODO describe the overline
+occurrences. We use a horizontal bar over terms (like in ${\overline{χ}}$) to
+denote lists.
 
-%TODO : get rid of instantiations ??
 \subsection{Typing}
 
-\begin{figure}[htpb]
+\begin{figure}[htbp]
   \begin{mathpar}
@@
   { \inferrule*[Left=Rec] 
-  { Γ' = Γ,~f : τ' ->^ε τ,~ x : τ' \\ Γ' |- t : τ,ε \\ Γ', cur : \efft{ε} |-_l p : \propml \\ 
+  { Γ' = Γ, x : τ' \\ Γ'' = Γ',~f : τ' ->^ε τ \\ Γ'' |- t : τ,ε \\ Γ', cur : \efft{ε} |-_l p : \propml \\ 
   Γ', old, cur : \efft{ε} |-_l q : τ -> \propml } 
@@
     {Γ |-_v  t1 ~ t2 : τ}
+  }
+  \end{mathpar}
+  \caption{Value typing}
+  \label{fig:valuetyping}
+\end{figure}
+
+\begin{figure}[htpb]
+\begin{mathpar}
+  { \inferrule*[Left=Value]
+  { Γ |-_v t : τ  }
+  { Γ |- t : τ,\emptyset  }
   } \and
@@
     { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε1  ε2 }
-  } \and
-  { \inferrule*[Left=Value]
-  { Γ |-_v t : τ  }
-  { Γ |- t : τ,\emptyset  }
-  } \and
+  }
+\end{mathpar}
+  \caption{Program typing}
+  \label{fig:progtyping}
+\end{figure}
+
+\begin{figure}[htbp]
+  \begin{mathpar}
   { \inferrule*[Left=Logic]
@@
     { Γ |-_l t : \ceil{τ} }
+  } \and
+  { \inferrule*[Left=LetLogic]
+    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |- t : τ  }
+    { Γ |-_l \letml~x~[\alist{χ}] = t'~\inml~t : τ }
   }
   \end{mathpar}
-  \caption{The Typing relation}
-  \label{fig:typing}
+  \caption{Logic typing}
+  \label{fig:logictyping}
 \end{figure}
@@
 
+\begin{figure}[htbp]
+  \begin{eqnarray*}
+    \ceil{x~[\alist{κ}]} &=& x~[\alist{κ}]\\
+    \ceil{λx:τ.t} &=& λx:τ.\ceil{t}\\
+    \ceil{\recml~f~(x:τ).\{p\} (t : ε ) \{q\}} &=& 
+    (λx:\ceil{τ}.λcur:\efft{ε}.p,λx:\ceil{τ}.λold,cur:\efft{ε}.q)\\
+    \ceil{t_{τ' -> τ}~t'} &=& \ceil{t}~\ceil{t'}
+  \end{eqnarray*}
+  \caption{Lifting values}
+  \label{fig:valuelift}
+\end{figure}
+
+We have already seen that programs and specification share in fact the same
+syntax. Though they are differently typed, their typing relations still share
+a lot. For this reason, and to underline the fact that in Hoare logic, values
+are shared between programs and the logic, we present three different typing
+relations. One for programs, one for specifications, and the common part is
+factored out in a third typing relation for values.
+
 Again in order to simplify future definitions, the definition typing relation
-differs somewhat from the standard way (see Fig.~\ref{fig:typing}). First of
+differs somewhat from the standard way (see Fig.~\ref{fig:progtyping}). First of
 all, we present the language in {\em A-normal form} (ANF), where all values
@@
 $t$ is a semantic value of type $τ$ and cannot have any effect. It can be seen
-that the set of terms described by $|-_v$ is the precisely the one obtained by
+that the set of terms described by $|-_v$ is precisely the one obtained by
 combining syntactic values and pure application in a well-typed way.
@@
 
-unusual : 
 missing : 
-* value injection
 * subeffecting
-* logic typing
 
@@
 
+\subsection{predefined constants}
+
+arithmetic operations, operations on pairs, logic operations on prop, pre/post
+
+reference operations
+!!
+
+operations on maps : restrict, combine
 \subsection{Sharing between logic and programs}
 
-To obtain sharing between
+use logic application in programs : no problem;
+use lambdas with effects in logic : specification
+
 \subsection{References}
@@
     { \inferrule*[Left=LetRef, Right={ρ \textup{fresh}, $ρ\notin τ$}]
-      {Γ |-_v t1 : τ' \\ Γ, x : τ'~\refml_ρ |- t2 : τ }
-      {Γ |- \letml~x = ref~ t1 ~\inml~ t2 : τ}
+      {Γ |-_v t1 : τ' \\ Γ, x : τ'~\refml_ρ |- t2 : τ, ε }
+      {Γ |- \letml~x = ref~ t1 ~\inml~ t2 : τ, ε \setminus ρ}
     }
@@
 
-\section{A Simple Weakest Preconditions Calculus.}
+\subsection{Some properties}
+
+$Γ|-_v v : τ$ implies $Γ |-_l \ceil{v} : \ceil{τ}$
+
+\section{A Simple Weakest Preconditions Calculus}
 
@@
 
-\section{Extensions.}
+\begin{figure}[htbp]
+  \begin{eqnarray*}
+    \wpre_s(v, q) &=& q~s~\ceil{v}\\
+    \wpre_s( t_{τ'->^ε τ} t' ,q) &=& \correct{t} /\ \correct(t') /\
+    fst~\ceil{t}~\ceil{t'}~s /\ \\ 
+    & & ∀s':\efft{ε}, x:τ.~  snd~\ceil{t}~\ceil{t'}~s~s'~x => q~s'~x\\
+    \wpre_s(\letml~x~[\alist{χ}] = v~\inml~t, q) &=&
+      c(v) /\ \letml~x~[\alist{χ}]=\ceil{v}~\inml~ \wpre(t,q)\\
+    \wpre_s(\letml~x = ( t1 : ε1 ) ~\inml~ ( t2 : ε2 ) , q) &=&
+    \wpre_s( t1 ,λs':\efft{ ε1  ε2 }.\wpre_{s'}(t2, q ))\\
+    \wpre_s(\ifml~ t1 ~\thenml~ t2 ~\elseml~ t3 , q) &=&
+    (\ceil{ t1 } = \trueml => \wpre_s( t2 , q)) /\ \\
+    & &(\ceil{ t1 } = \falseml => \wpre_s( t3 , q)) 
+  \end{eqnarray*}
+  \caption{The weakest precondition calculus}
+  \label{fig:wp}
+\end{figure}
+
+\section{Extensions}
 
@@
 
-\section{Implementation.}
+\section{Implementation}
 
-\section{Examples.}
+%what has been implemented
 
-\section{Related Work.}
+A prototype implementation by the name ``\who'' exists and is freely available. It implements the
+described language with type, region and effect polymorphism, higher-order
+logics and store types in the logic. It takes as an input a fully annotated
+program and outputs proof obligations which have to be proved to guarantee the
+correctness of the program with respect to its specification. Currently, only
+output to the interactive proof assistant Coq has been implemented. Our logic
+is a subset of Coq's, so the translation is particularly simple.
 
-\section{Conclusion.}
+%details
+% lines of code : less than 3000 lines of ocaml
+% limited form of type and effect inference
+%
+
+\section{Examples}
+
+\begin{center}
+    \begin{tabular}{ | r | r | r | r |}
+    \hline
+    Program & Code & Spec & Proofs \\ \hline
+    Array &  &  &  \\
+    Hashtbl & & &  \\ \hline
+    Memo/Ymemo & & &  \\ \hline
+    Koda-Ruskey~\cite{KanigFilliatre09wml} & & &  \\ \hline
+    \end{tabular}
+\end{center}
+
+\subsection{memo}
+\subsection{ymemo}
+
+\section{Related Work}
+
+\section{Conclusion}
+
+\paragraph{Future Work.}
+
+We are currently investigating the possibility to obtain first-order proof
+obligations, to be able to use SMT solvers~\cite{RanTin-SMTLIB}, which would
+avoid manual proofs in many cases. An obvious solution would be to reuse
+existing encodings of higher-order logic in first-order logic, such as the one
+by Pottier and Gauthier~\cite{pottier-gauthier-hosc}. This technique has been
+successfully employed by the Pangolin system~\cite{regis-gianas-pottier-08}.
+Another option comes from our observation that in many cases, the proof
+obligations for higher-order functions can actually be expressed in
+first-order logic, when one replaces top-level {\em quantifications} of
+functional arguments and invariants by top-level {\em declarations}. Doing
+this translation manually, we have obtained first-order proof
+obligations for the {\tt Array.iter} function which are all discharged by
+SMT solvers. We don't know if such a transformation can always be carried out.
+
+what about correspondence between $τ ->^\emptyset τ$ and $τ -> τ$?
+
+\bibliographystyle{plain}
+\bibliography{biblio}{}
 

### 2009-10-01 papers/esop2010/main.tex
go

@@
 
-\begin{abstract}
-  %TODO
+% ICFP
+%  We present a weakest precondition calculus for an imperative,
+%  higher-order programming language, restricted to non-aliasing global
+%  references. The main feature of the calculus is {\em effect
+%    polymorphism}, {\em i.e.}, higher-order functions can be specified by
+%  generalizing over the effects and specifications of their functional
+%  arguments. This allows truly modular reasoning about effectful
+%  higher-order functions. We achieve this by defining an extension of
+%  a standard higher-order logic, more suitable to reason about
+%  modifications of the store. The obtained proof obligations, however,
+%  are expressible in the standard logic and can be proved using
+%  standard theorem provers. We show that the necessary annotations
+%  and the obtained proof obligations are quite natural.
+%
+
+% Workshop ML
+%  We present \who, a tool for verifying effectful higher-order
+%  functions. It features {\em Effect polymorphism}, higher-order logic
+%  and the possibility to reason about state in the logic, which enable
+%  highly modular specifications of generic code. Several small
+%  examples and a larger case study demonstrate its usefulness. The
+%  \who~tool is intended to be used as an intermediate language for
+%  verification tools targeting ML-like programming languages.
+%
+
+\begin{abstract} We present a specification language suited for reasoning
+  about effectful higher-order programs. The language's main features are
+  higher-order logic and effect polymorphism, both of which enable a truly
+  modular treatment of higher-order functions. To our knowledge, this is the
+  first time that effect polymorphism is put to use in program verification.
+  We also define a weakest precondition calculus for this language which
+  generates proof obligations in higher-order logic. A simple mechanism based
+  on regions permits to restrict the analysis to alias-free programs, thus
+  obtaining much simpler proof obligations. Several examples demonstrate the
+  practicality of the system. A prototype implementation exists and has been
+  used to prove the correctness of these and more examples.  
 \end{abstract}
@@
 
+We have already seen that programs and specification share in fact the same
+syntax. Though they are differently typed, their typing relations still share
+a lot. For this reason, and to underline the fact that in Hoare logic, values
+are shared between programs and the logic, we present three different typing
+relations. One for programs, one for specifications, and the common part is
+factored out in a third typing relation for values. All typing relations
+contain an environment $Γ$, which is simply a map from variable names to {\em
+type schemes} of the form $∀\alist{χ}.τ$ .
+
 \begin{figure}[htbp]
   \begin{mathpar}
-  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~\alist{κ} :
+  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~[\alist{κ}] :
   τ[\alist{χ}|->\alist{κ}]} } \and
@@
   { \inferrule*[Left=Rec] 
-  { Γ' = Γ, x : τ' \\ Γ'' = Γ',~f : τ' ->^ε τ \\ Γ'' |- t : τ,ε \\ Γ', cur : \efft{ε} |-_l p : \propml \\ 
-  Γ', old, cur : \efft{ε} |-_l q : τ -> \propml } 
+  { Γ' = Γ, x : τ' \\ Γ'' = Γ',~f : τ' ->^ε τ \\ Γ'' |- t : τ,ε \\ Γ' |-_l
+  p : \efft{ε} -> \propml \\ Γ' |-_l q : \efft{ε} ->
+  \efft{ε} -> τ -> \propml } 
       {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
@@
 \end{figure}
+
+\paragraph{Value typing.}
+Let us first look at the common part concerning values. We define a relation
+$Γ|-_v t : τ$, which means ``$t$ is well-typed wrt. the environment $Γ$, and
+is of type $τ$''. Values cannot have any effect.  Variables and pure
+abstractions are typed in the usual way; type instantiations at variable
+occurences are substituted in the type. The rule {\sc REC} simply states that
+the body of the function can use the argument $x$ as well as the function $f$.
+The specifications $p$ and $q$, however, cannot use $f$ and must be well-typed
+wrt. the typing relation $|-_l$ described later on. Finally, pure applications
+between values are considered as values as well, as they don't have any
+effect.
 
@@
   { \inferrule*[Left=Let]
-    { Γ |- t1 : τ', ε1 \\ Γ, x : τ' |- t2 : τ, ε2 }
-    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε1  ε2 }
+    { Γ |- t1 : τ', ε \\ Γ, x : τ' |- t2 : τ, ε }
+    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε }
+  } \and
+  { \inferrule*[Left=Sub]
+    { Γ |- t : τ', ε \\ ε \subseteq ε' }
+    { Γ |- t : τ, ε' }
   }
@@
 \end{figure}
+
+\paragraph{Program typing.}
+The typing of programs, written $Γ|- t : τ,ε$, assigns a type $τ$ and an
+effect $ε$ to an expression $t$, given an environment. This typing is
+presented in {\em A-normal form}, a normal form where all intermediate values
+are \letml-introduced. All values are also programs with no effect (rule {\sc
+Value}). An application must consist of two values and the effect of the
+overall expression is precisely the effect of the called function (rule {\sc
+App}). There are two rules for \letml, due to the value restriction. One only
+can generalize type, effect and region variables when the bound expression is
+a value. In this case (rule {\sc LetV}), only the second term may produe
+effects. In the other case (rule {\sc Let}), no generalization takes place,
+but both terms can have effects. The {\sc Let} rule actually obliges both
+terms to have the same effect, but together with the {\sc Sub} rule, which can
+increase the effect of an expression, this can always be achieved.
+Finally, \ifml-expressions are typed in the usual way, both branches have to
+have the same effect.
+
+\paragraph{Logic typing.}
+
+The typing of annotations is basically an extension of the typing for values;
+logic terms never have any effect. To ensure this, we need to forbid calling
+effectful functions in the logic. One way is to simply omit the typing rule
+for effectful function calls. There is, however, another way which also
+enables us to inspect the pre/ and postcondition of a function. To this end,
+let us introduce the operation of {\em lifting} a type to the logic level (see
+Fig.~\ref{fig:typelift}).
+\begin{figure}[htpb]
+  \begin{eqnarray*}
+    \ceil{α} &=& α\\
+    \ceil{τ~\refml_ρ} &=& \ceil{τ}~\refml_ρ\\
+    \ceil{τ -> τ'} &=& \ceil{τ} -> \ceil{τ'}\\
+    \ceil{τ \times τ'} &=& \ceil{τ} \times \ceil{τ'}\\
+    \ceil{\efft{ε}} &=& \efft{ε}\\
+    \ceil{τ ->^ε τ'} &=& (\ceil{τ} -> \efft{ε} -> \propml)\times (\ceil{τ} ->
+    \efft{ε} -> \efft{ε} -> \ceil{τ'} -> \propml)
+  \end{eqnarray*}
+  \caption{Lifting types}
+  \label{fig:typelift}
+\end{figure}
+This translation traverses the type and replaces all effectful types by a
+tuple type which describes a tuple of a pre- and a postcondition. The
+precondition takes as an argument the argument of the function as well as the
+part of the store, at the moment of the function call, which corresponds to
+the effect of the function. The postcondition's two first arguments are the
+same, but in addition it expects the store {\em at the exit of the function},
+as well as the function's return value.
 
@@
   { \inferrule*[Left=LetLogic]
-    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |- t : τ  }
+    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |-_l t : τ  }
     { Γ |-_l \letml~x~[\alist{χ}] = t'~\inml~t : τ }
+  }\and
+  { \inferrule*[Left=IteLogic]
+    { Γ |- _l t1 : \boolml \\ Γ |-_l t2 : τ \\ Γ |-_l t3 : τ }
+    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ }
   }
@@
 
-\begin{figure}[htpb]
-  \begin{eqnarray*}
-    \ceil{α} &=& α\\
-    \ceil{τ~\refml_ρ} &=& \ceil{τ}\refml_ρ\\
-    \ceil{τ -> τ'} &=& \ceil{τ} -> \ceil{τ'}\\
-    \ceil{τ \times τ'} &=& \ceil{τ} \times-> \ceil{τ'}\\
-    \ceil{\efft{ε}} &=& \efft{ε}\\
-    \ceil{τ ->^ε τ'} &=& (\ceil{τ} -> \efft{ε} -> \propml)\times (\ceil{τ} ->
-    \efft{ε} -> \efft{ε} -> \ceil{τ'} -> \propml)()
-  \end{eqnarray*}
-  \caption{Lifting types}
-  \label{fig:typelift}
-\end{figure}
+Now we can come back to logic typing. Every value is also a logic term,
+but now it is assigned its {\em lifted} type. The two other rules are just 
+pure variants of their impure counterparts in the program typing.
 
-\begin{figure}[htbp]
-  \begin{eqnarray*}
-    \ceil{x~[\alist{κ}]} &=& x~[\alist{κ}]\\
-    \ceil{λx:τ.t} &=& λx:τ.\ceil{t}\\
-    \ceil{\recml~f~(x:τ).\{p\} (t : ε ) \{q\}} &=& 
-    (λx:\ceil{τ}.λcur:\efft{ε}.p,λx:\ceil{τ}.λold,cur:\efft{ε}.q)\\
-    \ceil{t_{τ' -> τ}~t'} &=& \ceil{t}~\ceil{t'}
-  \end{eqnarray*}
-  \caption{Lifting values}
-  \label{fig:valuelift}
-\end{figure}
+\paragraph{Predefined constants.}
 
-We have already seen that programs and specification share in fact the same
-syntax. Though they are differently typed, their typing relations still share
-a lot. For this reason, and to underline the fact that in Hoare logic, values
-are shared between programs and the logic, we present three different typing
-relations. One for programs, one for specifications, and the common part is
-factored out in a third typing relation for values.
+To actually be able to write programs in this language, we need many
+predefined constants and types. The typing relations already mention two,
+\propml\ and \boolml. \propml\ is the logical truth and contains the values
+\Trueml\ and \Falseml, while \boolml\ describes boolean (decidable) properties
+and contains \trueml\ and \falseml. The distinction between \propml\ and
+\boolml\ is the same one as in \coq, for example. Properties expressed in
+\propml\ are potentially undecidable and one cannot take decisions by testing
+their truth value. On the other hand, \boolml\ talks about decidable
+properties and so it makes sense to expect the predicate of the
+\ifml-statement to be of type \boolml.
 
-Again in order to simplify future definitions, the definition typing relation
-differs somewhat from the standard way (see Fig.~\ref{fig:progtyping}). First of
-all, we present the language in {\em A-normal form} (ANF), where all values
-obtained by intermediate computations are named. However, our definition of
-values is not syntactical: our ``semantic'' values are objects which are
-either syntactic values (abstractions and variables) or are obtained by {\em
-pure application}, the application which involves the pure arrow $τ->τ$.
-Furthermore, the system is a type and effect system. There are in fact two
-typing relations, the first one is of the form $Γ|-_v  t : τ$. In this case,
-$t$ is a semantic value of type $τ$ and cannot have any effect. It can be seen
-that the set of terms described by $|-_v$ is precisely the one obtained by
-combining syntactic values and pure application in a well-typed way.
+The reader may have wondered why the specification language does not contain
+any logical connectives. These can of course be declared independently, for
+example the conjuction is just a function: \begin{equation*} /\ :  \propml ->
+  \propml -> \propml \end{equation*} Actually, when equality is given, all
+  logical operations (even quantification) can be defined in the logic.
+%TODO citation needed
+In the remainder of the paper, we suppose predefined the usual logical
+connectives conjunction, disjunction, implication and equivalence, as well as
+universal and existential quantification. We also assume all the arithmetic
+operations as well as the constructor $(t,t)$  and the accessors $fst$ and
+$snd$ for pairs.
 
-In the other case, we have $Γ|- t : τ, ε$, where $t$ is an expression of
-type $τ$ and effect $ε$. Impure applications, \letml-expressions and
-\ifml-expressions can only be typed in this way. Effects can be produced only
-by function application, where the impure arrow $τ'->^ετ$ is involved. In this
-case, the effect of the application is precisely the effect $ε$ of the arrow,
-because the two subterms are semantic values and don't have any effect.
-
-The typing relies on two predefined types, \propml\ and \boolml. \propml\ is
-the logical truth and contains the values \Trueml\ and \Falseml, while
-\boolml\ describes boolean (decidable) properties and contains \trueml\ and
-\falseml. The distinction between \propml\ and \boolml\ is the same one as in
-\coq, for example. Properties expressed in \propml\ are potentially
-undecidable and one cannot take decisions by testing their truth value. On the
-other hand, \boolml\ talks about decidable properties and so it makes sense to
-expect the predicate of the \ifml-statement to be of type \boolml.
-
-missing : 
-* subeffecting
-
-%TODO is the LOGIC rule correct ??
-
-\subsection{predefined constants}
-
-arithmetic operations, operations on pairs, logic operations on prop, pre/post
-
-reference operations
-!!
-
-operations on maps : restrict, combine
-\subsection{Sharing between logic and programs}
-
-use logic application in programs : no problem;
-use lambdas with effects in logic : specification
+% TODO TODO
+%Finally, we introduce two symbols
+%\begin{eqnarray*}
+%  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
+%  |_{ε} &:& ∀ e. \efft{ e ε } -> \efft{ ε }.
+%  !! & : & ∀αρ e. α~\refml_ρ -> \efft{ ρ e } -> α
+%\end{eqnarray*}
+%The first one denotes a combining operation for stores; the resulting store
+%contains all the mappings of the second store, as well as all the mappings of
+%the first one, as long as they are not present in the second one.
+%union of the two arguments, in the sense that its domain is the union of its
+%arguments.
 
@@
 \subsection{Alias restriction}
+
+The presented programming language, along with the specifications, can, in
+fact be used to prove programs. The weakest precondition calculus presented in
+the next section can be used to obtain proof obligations which, when proved,
+guarantee the correctness of the program with respect to its specification. On
+the other hand, in its presented form, and because of the presence of
+aliasing, one would constantly need equalities and inequalities between
+reference variables to be able to reason about a program.
 
@@
 
+A way out is to restrict programs to alias free programs. Fortunately, it is
+easy to modify the presented typing rules to render all programs with aliasing
+ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
+to identify memory locations entirely; each region shall describe precisely
+one memory location. Starting from this assumption, a program can actually
+exhibit {\em two} forms of aliasing. In the first one, it may use several
+names (region variables) for the same memory location. This can happen when a
+function is generalized over, say, two region variables, assuming they are
+different, and is then instantiated with the same region variable twice. The
+classical example is the following program
+%TODO
+
+which is only correct if $ ρ1 $ and $ ρ2 $ are indeed different. To enforce
+this, we add a side condition to the instantiation: the substitution $σ$ must
+be {\em compatible} with the type $τ$. To understand what compatibility means,
+consider an {\em atomic} substitution $[ρ|->ρ']$. It is compatible with any
+type which does not contain $ρ'$. Similarly, the substitution $[e|-> ρ1  ρ2
+e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
+Expressed more generally, an atomic substitution is compatible with a type if
+the type does not contain any region or effect variables which are in the
+range of the substitution. Finally, a composite substitution is compatible
+%TODO atomic and composite
+with a type if each atomic substitution is compatible with the result of the
+application of the previous substitution. In this way, it becomes impossible
+to have different names for a single reference cell.
+
+The second form of aliasing is when we use the same name for {\em different}
+memory locations. This can happen when we allow the user to call $ref$ several
+times with the same region instantiation. A simple possibility is to attach
+the suppress the $ref$ function entirely and to replace it by a \letml\refml\ 
+construct which combines name creation and memory allocation. In this way,
+there can of course be only one reference per name. On the other hand, because
+of the side condition $ρ\notin τ$, the reference cannot escape the scope of
+its \letml\refml\ construct, and so, a function cannot return a reference
+which has been created in its body. This restriction could be leveraged a bit if
+one seperates name creation and allocation, while still maintaining that one
+can only allocate a single reference in each region. This is how it is done in
+the correctness proof of this restriction. The presented \letml\refml\
+construct is of course a special case.
+%TODO now, all operations on maps can be statically resolved.
+
 \subsection{Some properties}
 
+%TODO
 $Γ|-_v v : τ$ implies $Γ |-_l \ceil{v} : \ceil{τ}$
+
+Subject reduction of our type system is a consequence of subject reduction of
+the region calculus of Tofte/Talpin~\cite{tofte97ic}.
 
@@
 
-In the following, we use the metasymbol $v$ to denote well-typed semantic
-values, i.e. terms with a typing derivation $Γ|-_v v : τ$.
+We now proceed to explain how proof obligations are obtained from the
+annotated program.
+
+In the following, we use the metasymbol $v$ to denote well-typed values, i.e.
+terms with a typing derivation $Γ|-_v v : τ$.
 
@@
   \begin{eqnarray*}
+    \ceil{x~[\alist{κ}]} &=& x~[\alist{κ}]\\
+    \ceil{λx:τ.t} &=& λx:τ.\ceil{t}\\
+    \ceil{\recml~f~(x:τ).\{p\} t \{q\}} &=& 
+    (λx:\ceil{τ}.p,λx:\ceil{τ}.q)\\
+    \ceil{t_{τ' -> τ}~t'} &=& \ceil{t}~\ceil{t'}
+  \end{eqnarray*}
+  \caption{Lifting values}
+  \label{fig:valuelift}
+\end{figure}
+
+\begin{figure}[htbp]
+  \begin{eqnarray*}
+    c(x~[\alist{κ}]) &=& \Trueml \\
+    c(λx:τ.t) &=& ∀x:\ceil{τ}.c(t) \\
+    c(t~t') &=& c(t) /\ c(t') \\
+    c(\recml~f~(x:τ).\{p\}(t : τ',ε)\{q\}) &=&
+    ∀x:\ceil{τ}.∀f:\ceil{τ->^ετ'}.  f = \ceil{v} => \\& &
+    ∀s:\efft{ε}.p~s => \wpre_s(t,q \\
     \wpre_s(v, q) &=& q~s~\ceil{v}\\
@@
     fst~\ceil{t}~\ceil{t'}~s /\ \\ 
-    & & ∀s':\efft{ε}, x:τ.~  snd~\ceil{t}~\ceil{t'}~s~s'~x => q~s'~x\\
+    & & ∀s':\efft{ε}∀ x:\ceil{τ}.~  snd~\ceil{t}~\ceil{t'}~s~s'~x => q~s'~x\\
     \wpre_s(\letml~x~[\alist{χ}] = v~\inml~t, q) &=&
       c(v) /\ \letml~x~[\alist{χ}]=\ceil{v}~\inml~ \wpre(t,q)\\
-    \wpre_s(\letml~x = ( t1 : ε1 ) ~\inml~ ( t2 : ε2 ) , q) &=&
-    \wpre_s( t1 ,λs':\efft{ ε1  ε2 }.\wpre_{s'}(t2, q ))\\
+    \wpre_s(\letml~x = ( t1 : τ, ε1 ) ~\inml~ ( t2 : ε2 ) , q) &=&
+    \wpre_s( t1 ,λs':\efft{ ε1  ε2 }.λx:\ceil{τ}.\wpre_{s'}(t2, q ))\\
     \wpre_s(\ifml~ t1 ~\thenml~ t2 ~\elseml~ t3 , q) &=&
-    (\ceil{ t1 } = \trueml => \wpre_s( t2 , q)) /\ \\
-    & &(\ceil{ t1 } = \falseml => \wpre_s( t3 , q)) 
+      \ifml~ \ceil{ t1 }~\thenml~ \wpre_s( t2 , q))~\elseml~\wpre_s( t3 , q)) \\
+      \wpre_s( (t : ε <: ε'), q) &=& \wpre_{s|_ε}(t, λs':\efft{ε}.q~(s\oplus s'))
   \end{eqnarray*}
@@
 
-A prototype implementation by the name ``\who'' exists and is freely available. It implements the
+A prototype implementation by the name ``\who''
+exists~\cite{KanigFilliatre09wml} and is freely available. It implements the
 described language with type, region and effect polymorphism, higher-order

### 2009-10-01 papers/esop2010/main.tex
ok

@@
 
-\newcommand{\wpre}{ {\bf wp}}
+\newcommand{\wpre}{{\bf wp}}
 \newcommand{\sep}{ ~|~ }
@@
 \newcommand{\recml}{{\bf rec}}
-\newcommand{\efft}[1]{ \langle #1 \rangle }
-\newcommand{\lift}[1]{ \langle #1 \rangle }
-\newcommand{\alist}[1]{ \overline{#1} }
+\newcommand{\efft}[1]{\langle #1 \rangle}
+\newcommand{\lift}[1]{\langle #1 \rangle}
+\newcommand{\alist}[1]{\overline{#1} }
 \newcommand{\Trueml}{{\bf True}}
@@
 \newcommand{\who}{Who}
-\newcommand{\correct}[1]{ c(#1) }
+\newcommand{\correct}[1]{c(#1)}
 
@@
 
-We now proceed to explain how proof obligations are obtained from the
-annotated program.
+The aim of the weakest precondition calculus is to obtain proof obligations
+from the annotated program. To do this, the calculus starts from a formula
+which holds, or is supposed to hold, at the end of the program, and walks
+backwards in the program source to obtain the weakest formula which has to
+hold at the beginning of the program (hence the name {\em weakest
+precondition}.
 
 In the following, we use the metasymbol $v$ to denote well-typed values, i.e.
-terms with a typing derivation $Γ|-_v v : τ$.
+terms with a typing derivation $Γ|-_v v : τ$. In the definition of the weakest
+precondition calculus, we will use the notation $(t : τ, ε)$ to denote that
+$t$ has type $τ$ and effect $ε$. We assume that applications of the
+subeffecting rule {\sc Sub} rule are explicitely inserted in the term; they
+are written $(t:ε<:ε')$.
 
@@
 
+First of all, analogous to the lifting of types to the logic level, we define
+a lifting of values to the logic level. Similarly, this translation traverses
+the value until it encounters an impure function. From the logic point of
+view, an impure function is represented by its specification, so the
+translation keeps only the pair of pre- and postcondition, both abstracted
+over the function argument. The function body is thrown away.
+
 \begin{figure}[htbp]
@@
     ∀x:\ceil{τ}.∀f:\ceil{τ->^ετ'}.  f = \ceil{v} => \\& &
-    ∀s:\efft{ε}.p~s => \wpre_s(t,q \\
-    \wpre_s(v, q) &=& q~s~\ceil{v}\\
-    \wpre_s( t_{τ'->^ε τ} t' ,q) &=& \correct{t} /\ \correct(t') /\
+    ∀s:\efft{ε}.p~s => \wpre_s(t,q~s) \\
+    \wpre_s(v, q) &=& c(v) /\ q~s~\ceil{v}\\
+    \wpre_s( t_{τ'->^ε τ} t' ,q) &=& \correct{t} /\ \correct{t'} /\
     fst~\ceil{t}~\ceil{t'}~s /\ \\ 
@@
     \wpre_s(\letml~x~[\alist{χ}] = v~\inml~t, q) &=&
-      c(v) /\ \letml~x~[\alist{χ}]=\ceil{v}~\inml~ \wpre(t,q)\\
-    \wpre_s(\letml~x = ( t1 : τ, ε1 ) ~\inml~ ( t2 : ε2 ) , q) &=&
-    \wpre_s( t1 ,λs':\efft{ ε1  ε2 }.λx:\ceil{τ}.\wpre_{s'}(t2, q ))\\
+      ∀\alist{χ}.c(v) /\ \letml~x~[\alist{χ}]=\ceil{v}~\inml~ \wpre(t,q)\\
+    \wpre_s(\letml~x = ( t1 : τ,ε) ~\inml~ t2 , q) &=&
+    \wpre_s( t1 ,λs':\efft{ ε }.λx:\ceil{τ}.\wpre_{s'}( t2 , q ))\\
     \wpre_s(\ifml~ t1 ~\thenml~ t2 ~\elseml~ t3 , q) &=&
@@
 \end{figure}
+
+The weakest precondition calculus itself comes in two parts, one for values
+and one for effectful terms. Both are reasonably straightforward. The {\em
+correctness} of values, written $c(v)$, ensures that any effectful function
+appearing in a value adheres to its specification. To this end, the value is
+traversed, quantifying over bound variables and using conjunction for pure
+applications. When an effectful function is encountered, one has to prove that
+the precondition of the function implies the postcondition, given the function
+body.
+
+The part concerning effectful functions, denoted $\wpre_s(t,q)$ takes as
+argument a term $t$ of type $τ$ and effect $ε$, a postcondition $q$ of type
+$\efft{ε} -> τ -> \propml$ and a current state $s$ of type $\efft{ε}$. If the
+term is in fact a value, the value has to be correct and the postcondition
+must be true for the lifted value and the current state. In the case of an
+application, both values must be correct, the precondition of the function
+must hold in the current state, and for any state and return value of the
+function, the function's postcondition must imply the condition $q$ to be
+proved.
+
+As for the typing relation, we have two rules for the \letml-binding. If the
+bound term is a value, we can translate the \letml-binding in the program by a
+\letml-binding in the language. In the other case, the \wpre\ of $ t2 $ becomes
+the argument of the \wpre of $ t1 $ (we denote $ε$ the effect of $ t1 $ and $
+t2 $). The rule for \ifml-expressions is straightforward. The rule for
+subeffecting adapts the domain of the current store and the expected domain of
+the postcondition so that the types work out.
 

### 2009-10-02 papers/esop2010/main.tex
go

@@
 \newcommand{\who}{Who}
+\newcommand{\ocaml}{Ocaml}
 \newcommand{\correct}[1]{c(#1)}
@@
 \begin{document}
-\title{\mbox{Specifying Imperative Higher Order Programs}}
+\title{\mbox{Specifying Imperative Higher-Order Programs}}
 
@@
 %TODO affiliation
-%TODO find notions for:
-%  * ``semantic values'' v
-%  * ``store types'' <ε>
+%TODO explain somewhere the correspondence store type <-> map, effect variable
+%<-> domain, region <-> key
+%TODO the constants dealing with maps
+
+%external TODO : proofs for Hashtbl
+%external TODO : proofs of system
 %
@@
 
-% ICFP
-%  We present a weakest precondition calculus for an imperative,
-%  higher-order programming language, restricted to non-aliasing global
-%  references. The main feature of the calculus is {\em effect
-%    polymorphism}, {\em i.e.}, higher-order functions can be specified by
-%  generalizing over the effects and specifications of their functional
-%  arguments. This allows truly modular reasoning about effectful
-%  higher-order functions. We achieve this by defining an extension of
-%  a standard higher-order logic, more suitable to reason about
-%  modifications of the store. The obtained proof obligations, however,
-%  are expressible in the standard logic and can be proved using
-%  standard theorem provers. We show that the necessary annotations
-%  and the obtained proof obligations are quite natural.
-%
-
-% Workshop ML
-%  We present \who, a tool for verifying effectful higher-order
-%  functions. It features {\em Effect polymorphism}, higher-order logic
-%  and the possibility to reason about state in the logic, which enable
-%  highly modular specifications of generic code. Several small
-%  examples and a larger case study demonstrate its usefulness. The
-%  \who~tool is intended to be used as an intermediate language for
-%  verification tools targeting ML-like programming languages.
-%
-
 \begin{abstract} We present a specification language suited for reasoning
@@
   higher-order logic and effect polymorphism, both of which enable a truly
-  modular treatment of higher-order functions. To our knowledge, this is the
-  first time that effect polymorphism is put to use in program verification.
-  We also define a weakest precondition calculus for this language which
-  generates proof obligations in higher-order logic. A simple mechanism based
-  on regions permits to restrict the analysis to alias-free programs, thus
-  obtaining much simpler proof obligations. Several examples demonstrate the
-  practicality of the system. A prototype implementation exists and has been
-  used to prove the correctness of these and more examples.  
+  modular treatment of higher-order functions.  We also define a weakest
+  precondition calculus for this language which generates proof obligations in
+  higher-order logic.  To our knowledge, this is the first time that effect
+  polymorphism is put to use in program verification.  A simple mechanism
+  based on regions permits to restrict the analysis to alias-free programs,
+  thus obtaining much simpler proof obligations. Several examples demonstrate
+  the practicality of the system. A prototype implementation exists and has
+  been used to prove the correctness of these and more examples.
 \end{abstract}
@@
 \section{Introduction}
+
+The purpose of the present paper is to introduce a simple weakest precondition
+calculus for imperative higher-order programs. Higher-order functions are a
+key feature, not only of purely functional languages like Haskell and the pure
+fragment of ML, but also in combination with imperative features, like
+references. However, reasoning about higher-order functions is difficult, and
+especially so in the presence of effectful computations. Any function's
+specification depends on its arguments. In a higher-order setting, arguments
+may be functions that have their own specification. In consequence, to be
+modular, the specification of a higher-order function can only be formulated
+{\em depending on the specifications of its functional arguments}. In a
+setting where side effects are possible, one also needs to formulate the
+specification {\em depending on the side effects of the functional arguments}.
+We claim that to be truly modular, the specification of a higher-order
+function must be generic with respect to the specification {\em and} the
+effect of its arguments.
+
+In this paper, we first present a specification language combining features of
+a programming language and logic. It uses an effect system and a region
+calculus to track references. The effect system contains {\em effect
+polymorphism}, thus dealing well with higher-order functions. The
+specifications can be expressed in higher-order logic, to be able to state
+properties of functional arguments. We show how two very simple modifications
+to the system let us obtain alias-free programs, with the aim of obtaining
+much simpler proof obligations.
+
+We then proceed to define a weakest precondition calculus based on this
+language, transforming an annotated program into a logic formula, which has to
+be valid for the program to be correct. This calculus is very straight-forward
+and syntax-directed.
+
+Finally, we show the usefulness of the system by giving several examples of
+annotated programs. All the programs have been proved correct using our
+prototype implementation called \who.
+
+\paragraph{Outline.}
+Section~\ref{sec:genpic} gives an introduction to the described language as
+well as an example. Sections~\ref{sec:syntax} and \ref{sec:wp} give the
+details about the programming language and the weakest precondition calculus.
+In section~\ref{sec:implem} we speak about the implementation and show some
+examples. Finally, section~\ref{sec:related} discusses related work.
+
+\section{The general picture and a motivating example}
+\label{sec:genpic}
 %TODO Hoare logic = sharing between logic and prog language + alias free
-
 also one or two examples in the described language
@@
 \section{A programming language for specification}
+\label{sec:syntax}
 
@@
 \begin{itemize}
-  \item ANF
-  \item values defined by typing 
-  \item  two different types of arrows
-  \item  distinction bool and prop
   \item  notion of substitution - correspondence χ κ
@@
 presented in {\em A-normal form}, a normal form where all intermediate values
-are \letml-introduced. All values are also programs with no effect (rule {\sc
+are \letml-introduced. This is not a restriction, as any program can be
+transformed an A-normal form.
+
+All values are also programs with no effect (rule {\sc
 Value}). An application must consist of two values and the effect of the
@@
 \propml -> \propml \end{equation*} Actually, when equality is given, all
-  logical operations (even quantification) can be defined in the logic.
-%TODO citation needed
-In the remainder of the paper, we suppose predefined the usual logical
-connectives conjunction, disjunction, implication and equivalence, as well as
-universal and existential quantification. We also assume all the arithmetic
-operations as well as the constructor $(t,t)$  and the accessors $fst$ and
-$snd$ for pairs.
+logical operations (even quantification) can be defined in the
+logic~\cite{andrews86}.  In the remainder of the paper, we suppose
+predefined the usual logical connectives conjunction, disjunction,
+implication and equivalence, as well as universal and existential
+quantification. We also assume all the arithmetic operations as well as the
+constructor $(t,t)$  and the accessors $fst$ and $snd$ for pairs.
 
+%TODO pre and post
 % TODO TODO
@@
 %TODO
+\begin{who}
+  let f [ρ ρ'] (x : int ref(ρ)) (y : int ref(ρ')) =
+  { }
+  x := !x + 1;
+  !y
+  { r : r = !!y old }
+\end{who}
 
-which is only correct if $ ρ1 $ and $ ρ2 $ are indeed different. To enforce
+which is only correct if $ ρ $ and $ ρ' $ are indeed different. To enforce
 this, we add a side condition to the instantiation: the substitution $σ$ must
@@
 \section{A Simple Weakest Preconditions Calculus}
+\label{sec:wp}
 
@@
 
-\section{Extensions}
-
-subtyping: effects
-
+%TODO lemmas about correctness
 \section{Implementation}
-
-%what has been implemented
+\label{sec:implem}
 
@@
 
-%details
-% lines of code : less than 3000 lines of ocaml
-% limited form of type and effect inference
-%
+The implementation consists of less than 3000 lines of \ocaml\ code, including
+a limited form of type and effect inference.
 
@@
 
+In this section, we will show two simple examples which use higher-order
+features.
+\subsection{A simple memoization function}
+
+\begin{figure}[tpb]
+\begin{who}
+type ('a,'b) map
+logic mem : forall ('a, 'b) . 'a -> ('a,'b) map -> bool
+logic get : forall ('a, 'b) . 'a -> ('a,'b) map -> 'b
+logic set : forall ('a, 'b) . 'a -> 'b -> ('a,'b) map -> ('a,'b) map
+
+let stores f0 m =
+  forall x : Z, mem x m = true -> get x m = f0 x
+
+parameter table : (int,int) map ref(ρ)
+
+let memo [e] (f0 : int -> int) (f : int ->{} int) (x : int) = 
+  {pre f x /\ stores f0 !!table}
+  if mem x !table then get x !table
+  else
+    let z = f x in
+    table := set x z !table;
+    z
+  {r : post f x r}
+\end{who}
+  \caption{The code along with the specification of the {\tt memo} function}
+  \label{fig:memo}
+\end{figure}
+
+Our first example is a simple memoization function {\tt memo} for (pure)
+functions.  It takes as an argument a function (from {\tt int} to {\tt int}
+here) and an integer and returns the same result as the functional argument
+would; however, before executing the argument, it will look up the integer in
+a table to see if the function hasn't been called yet on this integer. If so,
+the table's value is returned, otherwise we call the function, store the
+result in the table, and return it. Fig~\ref{fig:memo} shows the code along
+with the specification. The specification is very simple; it states that {\tt
+memo} will return a value that verifies the postcondition of {\tt f}, as long
+is initially we have the precondition of f and as long as the table stores
+only valid key/value pairs. As we are not interested in the exact way of the
+implemenation of the map, we simply model it here by a reference to a
+functional map with get/set functions.
+
+\subsection{ymemo}
+
+\begin{figure}[tpb]
+\begin{who}
+  let realizes 
+  let ymemo (f0 : int -> int) 
+            (ff : (int ->{table} int) ->{} int ->{table} int) = 
+  { forall (k : [[int ->{table} int ]]),
+      { realizes f0 k} ff k { r : realizes f0 r} }
+  let rec f (x : int) : int ->{table} int =
+    { stores f0 !!table }
+    if mem x !table then get x !table
+    else
+      let z = ff f x in
+      table := set x z !table; z
+    {r : r = f0 x /\ stores f0 !!table}
+  in f
+  { rf : realizes f0 rf }
+\end{who}
+\caption{The {\tt ymemo function} and its specifacation}
+  \label{fig:<+label+>}
+\end{figure}<++>
+
+\subsection{All examples - an overview}
+\begin{figure}[tbp]
 \begin{center}
@@
     \hline
-    Program & Code & Spec & Proofs \\ \hline
-    Array &  &  &  \\
-    Hashtbl & & &  \\ \hline
-    Memo/Ymemo & & &  \\ \hline
-    Koda-Ruskey~\cite{KanigFilliatre09wml} & & &  \\ \hline
+    Program & Ocaml Code & Code + Spec in \who & Proofs \\ \hline
+    Memo/ymemo & 15 loc & 26 loc & 90 lop  \\ \hline
+    {\tt Array} & 69 loc & 170 loc  & 320 lop  \\ \hline
+    {\tt Hashtbl} (incomplete) & & &  \\ \hline
+    Koda-Ruskey~\cite{KanigFilliatre09wml} & 10 loc & 30 loc & $\sim$700 lop  \\ \hline
     \end{tabular}
 \end{center}
+  \caption{A summary of all programs proved so far using the \who\ tool}
+  \label{fig:whoproofs}
+\end{figure}
 
-\subsection{memo}
-\subsection{ymemo}
+Fig.~\ref{fig:whoproofs} shows an overview over the programs we have been able
+to prove using the \who\ tool. The first line summarizes the numbers of the
+memo/ymemo example of the previous subsections.
+
+The next two lines concern the two modules {\tt Array} and {\tt Hashtbl} of
+the \ocaml\ standard library. For the {\tt Array} module, all functions
+present in the \ocaml\ module have been proved, with the exception of the {\tt
+Array.map} family, whose specification is very similar to {\tt iter}, and the
+{\tt Array.sort} function, which is quite complicated to prove, but not very
+interesting concerning higher-order functions or side effects. In this proof
+development, arrays are modeled as references to functional arrays with the
+usual axioms on access and update.
+
+The proof development concerning the {\tt Hashtbl} module is ongoing work. The
+difficult part of this development is the model of the \ocaml\ hash table; In
+\ocaml, when one overwrites a previously existing entry in the table, the old
+value still exists, and is only {\em shadowed} by the new key/value pair. If
+one calls {\tt Hashtbl.remove} using this key, the old value reappears. Also,
+if several bindings for a key exist, the key may be encountered several times,
+during a call of {\tt Hashtbl.iter}, for example. A possible way of modeling
+this is an association list instead of a simple mapping, where keys in front
+of the list can shadow the same key further down in the list. 
+% TODO Currently, only
+%two functions of this module have been proved correct.
+
+Finally, the last line shows the numbers for the proof development of the
+Koda-Ruskey algorithm~\cite{KodaRuskey93}. A small, efficient
+implementation~\cite{FilliatrePottier02} is shown to be correct. More details
+about this development can be found in~\cite{KanigFilliatre09wml}.
 
 \section{Related Work}
+\label{sec:related}
 
+%TODO rewrite
+%TODO new citation for Ynot
+Honda, Yoshida and Berger~\cite{Honda05anobservationally} present a
+logic for imperative higher-order programs, using a slightly more
+expressive language than ours. They present their calculus using rules
+in Hoare logic style, while we presented a \wpre-calculus. Their
+system does not provide the same modularity than ours, as one can not
+reason modularly about side effects of functional arguments. It also
+has the drawback of being difficult to implement, due to the presence
+of a large number of non-structural rules. Our system, starting from a
+correctly annotated program, is syntax-directed and thus entirely
+automated. Their logic of proof obligations is non-standard, while we
+obtain obligations in higher-order logic. However, their calculus can
+deal easily with functions in references, even if the stored function
+reads or modifies the reference it is stored in. While, with the
+appropriate extensions, one can reason about functions in references
+in our system, this particular kind of programs cannot be reasoned
+about. Honda, Yoshida and Berger prove a completeness result for their
+system. We have proved no such result. Their extended
+system~\cite{BergerHondaYoshida05aliasing} has the same advantages and
+drawbacks compared to ours, but additionally deals with aliasing
+situations.
+
+The Coq proof assistant~\cite{CoqManualV81}, in particular in
+combination with the Program extension~\cite{sozeau07icfp}, can be
+used to implement and prove arbitrary properties of purely functional
+programs, using dependent types. The appeal of that approach is that
+the only limit of expressiveness is the one of Coq. One can also do
+meta-reasoning about programs inside Coq and, at the end of the
+development, extract a certified implementation in \ocaml\ or
+Haskell. The confidence in programs proved in this way is very high,
+because in the end one obtains a Coq proof term that corresponds to
+the correctness of the program. This term has been checked by Coq's
+kernel. However, without extensions such as Ynot (see next paragraph),
+it is impossible to implement (and reason about) effectful
+computations. Another drawback of this approach is lack of automation.
+
+The Ynot System~\cite{Nanevski08Awkward} is an extension to the Coq
+proof assistant, capable of reasoning about imperative higher-order
+programs, including effectful functions as arguments, using a monad in
+which effectful computations can take place. It uses higher-order
+separation logic as the device to speak about modifications of the
+state. Ynot uses an (arguably less intuitive) monadic representation
+of effectful computations. Ynot is more expressive than our system and
+features modular reasoning (abstracting over the specification and
+effect of a function as argument) while being able to reason about
+aliasing situations thanks to separation logic.  On the other hand,
+Ynot shares the Coq's main weakness: lack of automation. While we
+haven't yet presented a platform to automatically discharge generated
+proof obligations, this seems easily feasible, along the same vein as
+Régis-Gianas and Pottier~\cite{regis-gianas-pottier-08}, who argue
+that first-order SMT solvers~\cite{RanTin-SMTLIB} are suitable even
+for higher-order proof obligations. They also present an encoding
+scheme to support their claim. Their practical results are
+promising. The Ynot system can not use this kind of automation because
+it uses separation logic. Automated provers for separation logic
+exist, but we are not aware of any attempts to use these tools in a
+higher-order setting.
+
+%TODO also say that an approach like Ynot ou Coq doesn't give any help
+%for the mechanical part of the proof.
+
+The Pangolin system, the implementation of the theoretical system
+of~\cite{regis-gianas-pottier-08}, can also be used to reason about
+purely functional programs. It is also one of the starting points of
+our work, and our purely functional fragment is basically the Pangolin
+system, although we removed algebraic data types for clarity of
+presentation. For this reason, we share the potential advantages (good
+automation, relatively simple system) and drawbacks (no
+meta-reasoning or observational equality, no machine-checked proof
+term). Pangolin cannot deal with side effects.
+
+Another line of work consists in giving a translation from an
+imperative to a functional language, with the aim of applying one of
+the techniques for purely functional programs to the
+translation. O'Hearn and Reynolds~\cite{ohearn-reynolds-00} present a
+translation from Algol with first class references and dynamic
+allocation to a linear $\lambda$-calculus. They impose an
+alias-avoiding restriction which is similar to ours. Charguéraud and
+Pottier~\cite{chargueraud-pottier-08} give a translation from Core ML
+with first-class references, dynamic allocation and aliasing to a
+functional language, using a very expressive type system with linear
+capabilities. Both systems handle much more expressive languages than
+ours. However, they do not describe a mechanism to obtain proof
+obligations, for example a \wpre-calculus. Although we use a simpler
+programming language, we present a complete cycle from imperative
+programs to proof obligations. We also claim that extending our
+language and our \wpre-calculus is simple, as long as aliasing remains
+excluded.  We leave these extensions for future work.
+
+Systems for verification of first-order programs like
+Why~\cite{Filliatre00a} or Spec\#~\cite{BarnettLS04} are the other
+starting points of this work. These systems generally deal very well
+with usual features of first-order programs, for example arrays, use
+SMT solvers to discharge proof obligations and strive for the best
+possible automation. We try to improve on this work by adding
+higher-order features, trying to maintain the degree of automation and
+ease of use.
 \section{Conclusion}
+\label{sec:conclusion}
 
@@
 obligations for the {\tt Array.iter} function which are all discharged by
-SMT solvers. We don't know if such a transformation can always be carried out.
+SMT solvers. It is future work to find out if such a transformation can always
+be carried out.
 
-what about correspondence between $τ ->^\emptyset τ$ and $τ -> τ$?
+Currently, our effect system is quite limited. One possible extension is the
+addition of subtyping: an effectful function $f$ of type $τ ->^ετ'$ can be seen
+as a function of type $τ ->^{ε'}τ'$, if $ε\subseteq ε'$. In particular, such a
+subtyping operation should come at no cost for program verification; it should
+still come for free that $f$ does not modify the variables in $ε'\setminus ε$. It
+would also be interesting to see if an ``impure'' function of type $τ
+->^\emptyset τ'$ can be seen as pure function $τ -> τ'$ (and be used in the
+logic).
+
+Another possibility would be to modify the effect system itself to allow for
+aliasing. In the setting with the aliasing restriction, our regions actually
+become {\em singleton regions}, regions containing exactly one memory
+location. Other, more powerful effect systems allow for {\em group regions},
+with the possibility to temporarily {\em focus } on a particular memory cell
+in that region. We see no fundamental difficulty in adopting such an effect
+system, but haven't worked out the details.
+
+Finally, from a programmer's point of view, it would be nice to have more
+features in the programming language, such as recursive types and algebraic
+datatypes. As long as so defined types verify a {\em positivity condition},
+%TODO citation needed
+these extensions would be completely orthogonal to the weakest precondition
+calculus and the type and effect system used.
 

### 2009-10-05 papers/esop2010/main.tex
advancing on the paper

@@
 \newcommand{\refml}{{\bf ref}}
+\newcommand{\preml}{{\bf pre}}
+\newcommand{\postml}{{\bf post}}
 \newcommand{\propml}{{\bf prop}}
@@
 \newcommand{\unitml}{{\bf unit}}
+\newcommand{\intml}{{\bf int}}
 \newcommand{\coq}{Coq}
@@
 
-\author{Johannes Kanig \and Jean-Christophe Fill\^atre}
+\author{Johannes Kanig\inst{1,}\inst{2} \and Jean-Christophe
+Fill\^atre\inst{2,}\inst{1}}
 \institute{
-Laboratoire de Recherche en Informatique, Univ Paris-Sud, CNRS, Orsay F-91405\\
-INRIA Futurs, ProVal, Parc Orsay Universit\'e, F-91893
+INRIA Saclay - Île-de-France, ProVal, Orsay, F-91894 \and
+ LRI, Univ Paris-Sud, CNRS, Orsay, F-91405
 }
-%TODO affiliation
 %TODO explain somewhere the correspondence store type <-> map, effect variable
@@
 %TODO the constants dealing with maps
+
+%TODO clearly state possibilities/limitations of the system
+% * one can store everything in refs
+% * but aliasing restriction (refs in refs, lists of refs etc)
 
@@
 details about the programming language and the weakest precondition calculus.
-In section~\ref{sec:implem} we speak about the implementation and show some
+In section~\ref{sec:examples} we speak about the implementation and show some
 examples. Finally, section~\ref{sec:related} discusses related work.
@@
 \label{sec:genpic}
-%TODO Hoare logic = sharing between logic and prog language + alias free
-also one or two examples in the described language
 
-characteristics : 
-\begin{itemize}
-  \item Hoare logic
-  \item logic and program share values  
-  \item Type and effect system
-  \item Region typing to track references
-  \item the wp needs absence of aliasing to produce good POs
-\end{itemize}
+Consider the following first-order program\footnote{Throughout the document,
+we use \ocaml~Syntax for programs.} that accumulates the sum of the contents
+of some array $a$ in some global reference $c$, using an integer $i$ as
+a counter:
+\begin{ocaml}
+  let sum () = 
+    for i = 0 to Array.length a - 1 do
+      c := !c + a.(i)
+\end{ocaml}
+This simple  can be annotated in a standard Hoare logic (for
+example the system of \cite{Filliatre00a}) as follows:
+\begin{who}
+  let sum () =
+  {!!c = 0}
+    for i = 0 to length !a - 1 do
+    { !!c = {{{ $\sum_0^{i-1}$ get !!a i }}} }
+
+      c := !c + get !a i
+
+      {!!c = {{{ $\sum_0^{ {\tt length~!a} - 1}$ get !!a i }}} }
+\end{who}
+
+Here, assuming we model the array in the programming language by a reference
+to a functional array, with functions {\tt length} and {\tt get} with the
+obvious meaning. Also, in the logic we use a derefence operator {\tt !!}
+instead of the usual {\tt !}. In this case, our program is a {\tt for} loop,
+so it needs an {\em invariant annotation}, indicating which relation is
+preserved by the loop body. The equalities that need to be proved are very
+simple. Also, it is obvious that the program modifies exactly the references
+$r$ and $c$, and this fact can be formally confirmed by an effect analysis.
+
+A common reaction of a functional programmer to a program like
+$\mathit{sum}$ is to factor out the structure of the recursion, which
+may be useful in other situations, in a higher-order function. The
+more specific program is then recovered by function application. In this case,
+the more general function is of course simply {\tt Array.iter}:
+\begin{ocaml}
+  let iter f a = 
+    for i = 0 to Array.length a - 1 do
+      f a.(i)
+
+  let sum () =
+    iter (fun n -> c := !c + n) a
+\end{ocaml}
+It is obvious that the two programs do exactly the same
+thing. However, how do we prove the correctness of the second one? One
+possibility is to fix the future usage of ${\tt iter}$ by
+specifying what exactly is the specification of its argument {\tt f}:
+\begin{who}
+  let iter 
+    (f : (x : int) ->{c} 
+        { !!c = {{{ $\sum_0^{i-1}$ get !!a i }}} } 
+        unit 
+        { !!c = {{{ $\sum_0^{length !a - 1}$ get !!a i }}} } ) 
+    (a : int array) =
+    ...
+\end{who}
+Here we impose precisely the specification of {\tt f}, and thus we can
+derive that {\tt sum} computes the right value,
+using the same method as for the first-order case before. Note
+that, to be able to apply an effect analysis, we have to specify that
+$f$ modifies $c$, by giving it the type $\intml->^c\unitml$.
+
+Of course, this approach is not very satisfactory. While the program achieves
+the separation of the iterating function ({\tt iter}) and the combining
+function (the anonymous function), the specification does not do this at all;
+as specified,  {\tt iter} can basically only be applied to the same function,
+and can only compute a certain sum. If one wants to use {\tt iter} in another
+context, one has to copy the code, give a new specification adapted to the
+application and prove this new specification again. In addition, in order to
+be able to write the specification at all, we had to specialize the type of
+{\tt iter} to integer arrays, while its \ocaml\ counterpart is of course
+polymorphic. Said otherwise, this approach is not modular at all.
+
+There are two obstacles related to higher-order functions to be
+overcome in order to achieve modularity on the level of specification
+in the presence of side effects. We must be able to refer to the
+effect of a function parameter, and we must be able to refer to the
+specification (pre- and postconditions) of a function parameter, in a
+generic way.
+
+The first obstacle has a well known solution using a type and effect
+system~\cite{talpin94taed} with effect variables: effect polymorphism
+does exactly what we want here. Using effect polymorphism, we can say
+that {\tt f} (the argument of {\tt iter}) has type
+$\intml->^e\unitml$, where $e$ is an {\em effect variable}, which
+amounts to say that {\tt f} has a {\em certain} effect $e$. The total
+effect of {\tt iter} is then $e$, meaning {\tt iter}
+modifies whatever {\tt f} modifies.
+
+A solution to the second problem concerning specifications consists in
+representing a program function in the logic as the pair of its pre- and
+postcondition. This is an old idea, and it has recently been applied to purely
+functional programs by Régis-Gianas and
+Pottier~\cite{regis-gianas-pottier-08}.
+
+The main idea of this paper is to combine the effect-based Hoare logic of
+\cite{Filliatre00a} with the more powerful effect system of
+\cite{talpin94taed} and the more powerful annotations of
+\cite{regis-gianas-pottier-08} to obtain a modular system for proving
+higher-order programs with side effects. As in~\cite{regis-gianas-pottier-08},
+we use operators \preml\ and \postml\ to access the pre- and postcondition of
+a function. However, in our system, a precondition takes as an argument not
+only the function argument, but also the initial state, before the execution
+of the function. The postcondition takes the argument and return value, but
+also the states before {\em and} after the function call. Th so-called
+two-state postconditions are a way to get rid of most uses of auxiliary
+variables. However, the pre- and postcondition of {\tt f} may not refer to
+arbitrary parts of the state, but only the part {\tt f} reads or modifies.
+With these conventions, one can now write a generic specification for the {\tt
+iter} function, see ~Fig.~\ref{fig:iter}.
+
+\begin{figure}[tbp]
+\begin{who}[lines]
+let iter [e] (a : 'a farray ref) (inv : <e> -> int ->  prop)
+  (f : 'a ->{e} unit) =
+  { inv cur 0 /\ forall (i:int). 0 <= i /\ i < length !!a ->
+    forall (m : <e>).  inv m i -> pre f (get i !!a) m /\
+    forall (n : <e>).  post f (get i !!a) m n () -> inv n (i+1)
+  }
+  for index = 0 to length !a - 1 do
+    { inv cur index }
+    f (get index !a)
+  done
+  {inv cur (length !!a) }
+\end{who}
+\caption{The specification of the {\tt iter} function}
+  \label{fig:iter}
+\end{figure}
+
+There are several new elements here. First of all, we use effect polymorphism
+to introduce an effect variable {\tt e} standing for the effect of {\tt f}.
+Also, {\tt iter} has an additional argument {\tt inv}, the {\em invariant},
+taking as argument the portion of the state modified by f and the integer
+indicating the current iteration count. This is a logical argument, which
+means that it cannot be used in the program. Also, {\tt inv} is a {\em pure}
+function, which is indicated by the arrows without effect annotation. The
+postcondition is now quite clear: we simply want to establish the invariant
+for the final state (the keyword {\tt cur} for the current state of the
+postcondition) and the length of the array, indicating that we have iterated
+over the entire array. It is also easy to see that the invariant {\tt inv} is
+precisely the invariant of the {\tt for} loop. However, the precondition is a
+bit complicated: it states that we need the invariant to hold initially (for
+the current state of the precondition and index $0$) and we need the invariant
+to be preserved by the function call. This is expressed by quantifying over
+two states, saying that the invariant has to imply the precondition of {\tt f}
+and the postcondition has to imply the invariant for the next array index.
+With a bit of syntactic sugar, this could be expressed much more concisely
+using a Hoare triple:
+\begin{who}
+  { inv cur i} f (get i !!a) {inv cur (i+1)}
+\end{who}
+
+The {\tt iter} function and its specification raise a question: is the
+specification correct if {\tt f} modifies the array {\tt a}, which is required
+for the good behavior of {\tt iter}? The answer is clearly no; if {\tt f}
+messes up the contents of {\tt a}, {\tt iter} may not actually maintain the
+invariant.  This is an instance of the {\em aliasing problem} (see, for
+example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
+does not support aliasing. In this paper we make the same choice and disallow
+aliasing in the programming language we introduce. Under this assumption, the
+specification we gave is correct.
+
+Entirely excluding aliasing renders our system inappropriate for
+reasoning about shared mutable structures. However, it is interesting
+to see that our alias exclusion restriction, in many cases,
+corresponds to a natural condition imposed on higher-order functions
+that operate on some data structure. It is a common
+requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration
+functions, for example, that the functional argument must not modify
+the structure to be iterated on. In our system, this fact follows from
+the restriction on effect instantiations, as will be seen in
+section~\ref{sec:syntax}. In the example, the array {\tt a}
+of the iterating function {\tt iter} must not be modified by its
+argument.
 
@@
 
-technical presentation:
-\begin{itemize}
-  \item  notion of substitution - correspondence χ κ
-  \item semantics is standard (goes to appendix )
-\end{itemize}
-
+%TODO  notion of substitution - correspondence χ κ
 \subsection{Syntax}
-\begin{figure}[htpb]
+\begin{figure}[tpb]
 \begin{eqnarray*}
@@
 of the same syntax as programs, but are differently typed (typing will be
-discussed in the next section). The term $p$ corresponds to the {\em
+discussed in the next subsection). The term $p$ corresponds to the {\em
 precondition}, $q$ to the {\em postcondition} of the function. The analogy to
@@
 
-\begin{figure}[htbp]
+\begin{figure}[tbp]
   \begin{mathpar}
@@
 abstractions are typed in the usual way; type instantiations at variable
-occurences are substituted in the type. The rule {\sc REC} simply states that
+occurrences are substituted in the type. The rule {\sc REC} simply states that
 the body of the function can use the argument $x$ as well as the function $f$.
@@
 
-\begin{figure}[htpb]
+\begin{figure}[tpb]
 \begin{mathpar}
@@
 can generalize type, effect and region variables when the bound expression is
-a value. In this case (rule {\sc LetV}), only the second term may produe
+a value. In this case (rule {\sc LetV}), only the second term may produce
 effects. In the other case (rule {\sc Let}), no generalization takes place,
@@
 for effectful function calls. There is, however, another way which also
-enables us to inspect the pre/ and postcondition of a function. To this end,
+enables us to inspect the pre- and postcondition of a function. To this end,
 let us introduce the operation of {\em lifting} a type to the logic level (see
 Fig.~\ref{fig:typelift}).
-\begin{figure}[htpb]
+\begin{figure}[tpb]
   \begin{eqnarray*}
@@
 any logical connectives. These can of course be declared independently, for
-example the conjuction is just a function: \begin{equation*} /\ :  \propml ->
+example the conjunction is just a function: \begin{equation*} /\ :  \propml ->
 \propml -> \propml \end{equation*} Actually, when equality is given, all
@@
 quantification. We also assume all the arithmetic operations as well as the
-constructor $(t,t)$  and the accessors $fst$ and $snd$ for pairs.
+constructor $(t,t)$  and the accessors $fst$ and $snd$ for pairs. We will also
+use the names \preml\ and \postml\ for $fst$ and $snd$, respectively.
 
-%TODO pre and post
 % TODO TODO
@@
 
-\begin{figure}[htpb]
+\begin{figure}[tpb]
   \begin{mathpar}
@@
 classical example is the following program
-%TODO
 \begin{who}
@@
 \end{who}
-
 which is only correct if $ ρ $ and $ ρ' $ are indeed different. To enforce
@@
 be {\em compatible} with the type $τ$. To understand what compatibility means,
-consider an {\em atomic} substitution $[ρ|->ρ']$. It is compatible with any
-type which does not contain $ρ'$. Similarly, the substitution $[e|-> ρ1  ρ2
-e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
-Expressed more generally, an atomic substitution is compatible with a type if
-the type does not contain any region or effect variables which are in the
-range of the substitution. Finally, a composite substitution is compatible
-%TODO atomic and composite
+consider an {\em atomic} region substitution (whose domain contains only one
+variable) $[ρ|->ρ']$. It is compatible with any type which does not contain
+$ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2 e'$] is compatible with
+any type which does not contain $ ρ1 , ρ2 $ nor $e'$.  Expressed more
+generally, an atomic substitution is compatible with a type if the type does
+not contain any region or effect variables which are in the range of the
+substitution. Finally, a {\em composite} substitution (whose domain contains more
+than one variable) is compatible
 with a type if each atomic substitution is compatible with the result of the
@@
 which has been created in its body. This restriction could be leveraged a bit if
-one seperates name creation and allocation, while still maintaining that one
+one separates name creation and allocation, while still maintaining that one
 can only allocate a single reference in each region. This is how it is done in
@@
 $t$ has type $τ$ and effect $ε$. We assume that applications of the
-subeffecting rule {\sc Sub} rule are explicitely inserted in the term; they
+subeffecting rule {\sc Sub} rule are explicitly inserted in the term; they
 are written $(t:ε<:ε')$.
 
-\begin{figure}[htbp]
+\begin{figure}[tbp]
   \begin{eqnarray*}
@@
 
-\begin{figure}[htbp]
+\begin{figure}[tbp]
   \begin{eqnarray*}
@@
 %TODO lemmas about correctness
-\section{Implementation}
-\label{sec:implem}
-
-A prototype implementation by the name ``\who''
-exists~\cite{KanigFilliatre09wml} and is freely available. It implements the
-described language with type, region and effect polymorphism, higher-order
-logics and store types in the logic. It takes as an input a fully annotated
-program and outputs proof obligations which have to be proved to guarantee the
-correctness of the program with respect to its specification. Currently, only
-output to the interactive proof assistant Coq has been implemented. Our logic
-is a subset of Coq's, so the translation is particularly simple.
-
-The implementation consists of less than 3000 lines of \ocaml\ code, including
-a limited form of type and effect inference.
-
 \section{Examples}
+\label{sec:examples}
 
@@
 features.
+
 \subsection{A simple memoization function}
@@
 only valid key/value pairs. As we are not interested in the exact way of the
-implemenation of the map, we simply model it here by a reference to a
+implementation of the map, we simply model it here by a reference to a
 functional map with get/set functions.
@@
 
+TODO %TODO
 \begin{figure}[tpb]
@@
 \end{who}
-\caption{The {\tt ymemo function} and its specifacation}
-  \label{fig:<+label+>}
-\end{figure}<++>
+\caption{The {\tt ymemo function} and its specification}
+  \label{fig:ymemo}
+\end{figure}
 
 \subsection{All examples - an overview}
+
+A prototype implementation of the presented system by the name ``\who''
+exists~\cite{KanigFilliatre09wml} and is freely available. It implements the
+described language with type, region and effect polymorphism, higher-order
+logic and store types in the logic. It takes as an input a fully annotated
+program and outputs proof obligations which have to be proved to guarantee the
+correctness of the program with respect to its specification. Currently, only
+output to the interactive proof assistant Coq has been implemented. Our logic
+is a subset of Coq's, so the translation is particularly simple.
+The implementation consists of less than 3000 lines of \ocaml\ code, including
+a limited form of type and effect inference.
+
 \begin{figure}[tbp]
@@
 
-%TODO rewrite
-%TODO new citation for Ynot
-Honda, Yoshida and Berger~\cite{Honda05anobservationally} present a
-logic for imperative higher-order programs, using a slightly more
-expressive language than ours. They present their calculus using rules
-in Hoare logic style, while we presented a \wpre-calculus. Their
-system does not provide the same modularity than ours, as one can not
-reason modularly about side effects of functional arguments. It also
-has the drawback of being difficult to implement, due to the presence
-of a large number of non-structural rules. Our system, starting from a
-correctly annotated program, is syntax-directed and thus entirely
-automated. Their logic of proof obligations is non-standard, while we
-obtain obligations in higher-order logic. However, their calculus can
-deal easily with functions in references, even if the stored function
-reads or modifies the reference it is stored in. While, with the
-appropriate extensions, one can reason about functions in references
-in our system, this particular kind of programs cannot be reasoned
-about. Honda, Yoshida and Berger prove a completeness result for their
-system. We have proved no such result. Their extended
-system~\cite{BergerHondaYoshida05aliasing} has the same advantages and
-drawbacks compared to ours, but additionally deals with aliasing
-situations.
+Honda, Yoshida and
+Berger~\cite{Honda05anobservationally,BergerHondaYoshida05aliasing} present a
+logic for imperative higher-order programs. They present their calculus using
+rules in Hoare logic style, while we presented a \wpre-calculus. Their system
+does not provide the same modularity than ours, as one can not reason
+modularly about side effects of functional arguments. It also has the drawback
+of being difficult to implement, due to the presence of a large number of
+non-structural rules. Our system, starting from a correctly annotated program,
+is syntax-directed and thus entirely automated. Their logic of proof
+obligations is non-standard, while we obtain obligations in higher-order
+logic. They deal with aliasing using equalities and disequalities between
+pointers, which we chose not to do. They can reason about the classic {\em
+Landin's knot}, which consists in obtaining recursion by storing a function in
+a reference which calls this same reference. Our typing restriction prohibits
+this particular kind of programs.  Honda, Yoshida and Berger prove a
+completeness result for their system. We have proved no such result.
 
-The Coq proof assistant~\cite{CoqManualV81}, in particular in
-combination with the Program extension~\cite{sozeau07icfp}, can be
-used to implement and prove arbitrary properties of purely functional
-programs, using dependent types. The appeal of that approach is that
-the only limit of expressiveness is the one of Coq. One can also do
-meta-reasoning about programs inside Coq and, at the end of the
-development, extract a certified implementation in \ocaml\ or
-Haskell. The confidence in programs proved in this way is very high,
-because in the end one obtains a Coq proof term that corresponds to
-the correctness of the program. This term has been checked by Coq's
-kernel. However, without extensions such as Ynot (see next paragraph),
-it is impossible to implement (and reason about) effectful
-computations. Another drawback of this approach is lack of automation.
+%, using a slightly more
+%expressive language than ours. However, their calculus can
+%deal easily with functions in references, even if the stored function
+%reads or modifies the reference it is stored in. While, with the
+%appropriate extensions, one can reason about functions in references
+%in our system, this particular kind of programs cannot be reasoned
+%about.  Their extended
+%system~\cite{BergerHondaYoshida05aliasing} has the same advantages and
+%drawbacks compared to ours, but additionally deals with aliasing
+%situations.
 
-The Ynot System~\cite{Nanevski08Awkward} is an extension to the Coq
-proof assistant, capable of reasoning about imperative higher-order
-programs, including effectful functions as arguments, using a monad in
-which effectful computations can take place. It uses higher-order
-separation logic as the device to speak about modifications of the
-state. Ynot uses an (arguably less intuitive) monadic representation
-of effectful computations. Ynot is more expressive than our system and
-features modular reasoning (abstracting over the specification and
-effect of a function as argument) while being able to reason about
-aliasing situations thanks to separation logic.  On the other hand,
-Ynot shares the Coq's main weakness: lack of automation. While we
-haven't yet presented a platform to automatically discharge generated
-proof obligations, this seems easily feasible, along the same vein as
-Régis-Gianas and Pottier~\cite{regis-gianas-pottier-08}, who argue
-that first-order SMT solvers~\cite{RanTin-SMTLIB} are suitable even
-for higher-order proof obligations. They also present an encoding
-scheme to support their claim. Their practical results are
-promising. The Ynot system can not use this kind of automation because
-it uses separation logic. Automated provers for separation logic
-exist, but we are not aware of any attempts to use these tools in a
-higher-order setting.
+The Coq proof assistant~\cite{CoqManualV81}, in particular in combination with
+the Program extension~\cite{sozeau07icfp}, can be used to implement and prove
+arbitrary properties of purely functional programs, using dependent types. The
+appeal of that approach is that the only limit of expressiveness is the one of
+Coq. One can also do meta-reasoning about programs inside Coq and, at the end
+of the development, extract a certified implementation in \ocaml\ or Haskell.
+The confidence in programs proved in this way is very high, because in the end
+one obtains a Coq proof term that corresponds to the correctness of the
+program. This term has been checked by Coq's kernel. However, without
+extensions such as Ynot (see next paragraph), it is impossible to implement
+(and reason about) effectful computations. Another drawback of this approach
+is lack of automation.
+
+The Ynot System~\cite{Nanevski08Awkward,chlipalaicfp09} is an extension to the
+Coq proof assistant, capable of reasoning about imperative higher-order
+programs, including effectful functions as arguments, using a monad in which
+effectful computations can take place. It uses higher-order separation logic
+as the device to speak about modifications of the state. Ynot uses an
+(arguably less intuitive) monadic representation of effectful computations.
+Ynot is more expressive than our system because it uses Coq as the underlying
+system, and features modular reasoning (abstracting over the specification and
+effect of a function as argument) while being able to reason about aliasing
+situations thanks to separation logic. Ynot also has limited support for
+proof automation, thanks to tactics dealing with separation logic
+formulas~\cite{chlipalaicfp09}. We claim that we can obtain much better
+automation for alias-free programs, and even higher-order functions. We
+explain in the next section how we plan to achieve this.
+
+%On the other hand, Ynot shares the
+%Coq's main weakness: lack of automation. While we haven't yet presented a
+%platform to automatically discharge generated proof obligations, this seems
+%easily feasible, along the same vein as Régis-Gianas and
+%Pottier~\cite{regis-gianas-pottier-08}, who argue that first-order SMT
+%solvers~\cite{RanTin-SMTLIB} are suitable even for higher-order proof
+%obligations. They also present an encoding scheme to support their claim.
+%Their practical results are promising. The Ynot system can not use this kind
+%of automation because it uses separation logic.  Automated provers for
+%separation logic exist, but we are not aware of any attempts to use these
+%tools in a higher-order setting.
 
@@
 
-Another line of work consists in giving a translation from an
-imperative to a functional language, with the aim of applying one of
-the techniques for purely functional programs to the
-translation. O'Hearn and Reynolds~\cite{ohearn-reynolds-00} present a
-translation from Algol with first class references and dynamic
-allocation to a linear $\lambda$-calculus. They impose an
-alias-avoiding restriction which is similar to ours. Charguéraud and
-Pottier~\cite{chargueraud-pottier-08} give a translation from Core ML
-with first-class references, dynamic allocation and aliasing to a
-functional language, using a very expressive type system with linear
-capabilities. Both systems handle much more expressive languages than
-ours. However, they do not describe a mechanism to obtain proof
-obligations, for example a \wpre-calculus. Although we use a simpler
-programming language, we present a complete cycle from imperative
-programs to proof obligations. We also claim that extending our
-language and our \wpre-calculus is simple, as long as aliasing remains
-excluded.  We leave these extensions for future work.
+Another line of work consists in giving a translation from an imperative to a
+functional language, with the aim of applying one of the techniques for purely
+functional programs to the translation. O'Hearn and
+Reynolds~\cite{ohearn-reynolds-00} present a translation from Algol with first
+class references and dynamic allocation to a linear $\lambda$-calculus. They
+impose an alias-avoiding restriction which is similar to ours. Charguéraud and
+Pottier~\cite{chargueraud-pottier-08} give a translation from Core ML with
+first-class references, dynamic allocation and aliasing to a functional
+language, using a very expressive type system with linear capabilities and
+group regions, while we only accept singleton regions. The second system can
+deal with aliasing programs. However, none of these works describe a mechanism
+to obtain proof obligations, for example a \wpre-calculus.  We present a
+complete cycle from imperative programs to proof obligations. We also claim
+that extending our language and our \wpre-calculus is simple, as long as
+aliasing is tracked precisely.  We leave these extensions for future work.
 
@@
 ease of use.
+
 \section{Conclusion}
 \label{sec:conclusion}
+
+We have presented a Hoare logic for higher-order programs with side
+effects. Its main feature is the ability to reason about higher-order
+functions in a {\em modular} way. A higher-order function can be
+specified by abstracting over the specification and the effect of its
+arguments. This specification can be reused when the higher-order
+function is instantiated with a particular effect and applied to a
+particular argument. This modularity is achieved by combining two
+previously known mechanisms. Effect polymorphism enables us to
+abstract over the modifications of a function. Reflecting values and
+types in a higher-order logic enables us to abstract over the
+specification of a function. We have proposed a way to
+reason about the evolution of a store, with potentially unknown
+components represented by effect variables. We have limited ourselves to
+programs without aliasing.
+
+In its simplicity and its drawbacks, our system is similar to the one
+of ~\cite{regis-gianas-pottier-08}. The system's definition holds
+entirely in Figure~\ref{fig:wp}, and it can be easily
+implemented because it is syntax-directed. On the other hand, we only
+consider partial correctness. Also, we can not reason about
+observational equivalence of programs. In our system, two functions
+are equivalent if their specifications are.
 
@@
 
-Another possibility would be to modify the effect system itself to allow for
+We see two possibilities to add the possibility to reason about aliasing
+programs. The first one is to modify the effect system itself to allow for
 aliasing. In the setting with the aliasing restriction, our regions actually
@@
 in that region. We see no fundamental difficulty in adopting such an effect
-system, but haven't worked out the details.
+system, but haven't worked out the details. A second possibility is to deal
+with aliasing via equalities and disequalities over pointers, using regions to
+obtain some separation automatically. On itself, we believe that this
+technique does not scale, due to the explosion of needed disequalities.
+However, in combination with other separation techniques, for example the
+Burstall-Bornat memory model which separates automatically objects of
+different structural types, this approach has worked well applied to
+first-order programs.
 
@@
 features in the programming language, such as recursive types and algebraic
-datatypes. As long as so defined types verify a {\em positivity condition},
-%TODO citation needed
-these extensions would be completely orthogonal to the weakest precondition
-calculus and the type and effect system used.
+data types. As long as so defined types verify a {\em positivity
+condition}~\cite{paulintlca93}, these extensions would be completely
+orthogonal to the weakest precondition calculus and the type and effect system
+used.
 

### 2009-10-06 papers/esop2010/main.tex
remarques de JC

@@
 \institute{
-INRIA Saclay - Île-de-France, ProVal, Orsay, F-91894 \and
+INRIA Saclay - Île-de-France, ProVal, Orsay, F-91893 \and
  LRI, Univ Paris-Sud, CNRS, Orsay, F-91405
@@
   higher-order logic.  To our knowledge, this is the first time that effect
-  polymorphism is put to use in program verification.  A simple mechanism
-  based on regions permits to restrict the analysis to alias-free programs,
-  thus obtaining much simpler proof obligations. Several examples demonstrate
-  the practicality of the system. A prototype implementation exists and has
-  been used to prove the correctness of these and more examples.
-\end{abstract}
+  polymorphism is put to use in program verification. Additionally, we propose
+  a region-based analysis to exclude aliasing, in order to obtain much simpler
+  proof obligations. Several examples demonstrate the practicality of the
+  system. A prototype implementation exists and has been used to prove the
+  correctness of these and more examples.  \end{abstract}
 
@@
 
-The purpose of the present paper is to introduce a simple weakest precondition
-calculus for imperative higher-order programs. Higher-order functions are a
-key feature, not only of purely functional languages like Haskell and the pure
+The purpose of the present paper is to propose a language suited for reasoning
+about effectful higher-order programs.  Higher-order functions are a key
+feature, not only of purely functional languages like Haskell and the pure
 fragment of ML, but also in combination with imperative features, like
-references. However, reasoning about higher-order functions is difficult, and
-especially so in the presence of effectful computations. Any function's
-specification depends on its arguments. In a higher-order setting, arguments
-may be functions that have their own specification. In consequence, to be
-modular, the specification of a higher-order function can only be formulated
-{\em depending on the specifications of its functional arguments}. In a
-setting where side effects are possible, one also needs to formulate the
+references or arrays. However, reasoning about higher-order functions is
+difficult, and especially so in the presence of effectful computations. Any
+function's specification depends on its arguments. In a higher-order setting,
+arguments may be functions that have their own specification. In consequence,
+to be modular, the specification of a higher-order function can only be
+formulated {\em depending on the specifications of its functional arguments}.
+In a setting where side effects are possible, one also needs to formulate the
 specification {\em depending on the side effects of the functional arguments}.
@@
 specifications can be expressed in higher-order logic, to be able to state
-properties of functional arguments. We show how two very simple modifications
-to the system let us obtain alias-free programs, with the aim of obtaining
-much simpler proof obligations.
-
-We then proceed to define a weakest precondition calculus based on this
-language, transforming an annotated program into a logic formula, which has to
-be valid for the program to be correct. This calculus is very straight-forward
-and syntax-directed.
-
-Finally, we show the usefulness of the system by giving several examples of
-annotated programs. All the programs have been proved correct using our
-prototype implementation called \who.
+properties of functional arguments.  We then proceed to define a weakest
+precondition calculus for this language, transforming an annotated
+program into a logic formula, which has to be valid for the program to be
+correct. This calculus is very straight-forward and syntax-directed.
+
+Programs in our language may contain aliases. However, we obtain much simpler
+proof obligations when aliases are excluded. Therefore, we also propose a
+static region-based analysis rejecting programs with aliasing. The weakest
+precondition calculus on the alias-free fragment has been implemented in a
+prototype tool called \who~\cite{KanigFilliatre09wml}.  We show the usefulness
+of this tool on several examples, which have all been proved correct.
 
 \paragraph{Outline.}
-Section~\ref{sec:genpic} gives an introduction to the described language as
-well as an example. Sections~\ref{sec:syntax} and \ref{sec:wp} give the
-details about the programming language and the weakest precondition calculus.
-In section~\ref{sec:examples} we speak about the implementation and show some
+Section~\ref{sec:genpic} introduces our language using a motivating example.
+Sections~\ref{sec:syntax} and \ref{sec:wp} give the details about the
+programming language and the weakest precondition calculus.  In
+section~\ref{sec:examples} we present our implementation and show some
 examples. Finally, section~\ref{sec:related} discusses related work.
@@
 Consider the following first-order program\footnote{Throughout the document,
-we use \ocaml~Syntax for programs.} that accumulates the sum of the contents
-of some array $a$ in some global reference $c$, using an integer $i$ as
-a counter:
+we use \ocaml~syntax for programs.} that accumulates the sum of the contents
+of some array {\tt a} in some global reference {\tt c}, using an integer {\tt
+i} as a counter:
 \begin{ocaml}
@@
       c := !c + a.(i)
+    done
 \end{ocaml}
-This simple  can be annotated in a standard Hoare logic (for
+This simple program can be annotated in a standard Hoare logic (for
 example the system of \cite{Filliatre00a}) as follows:
@@
   let sum () =
-  {!!c = 0}
-    for i = 0 to length !a - 1 do
-    { !!c = {{{ $\sum_0^{i-1}$ get !!a i }}} }
-
-      c := !c + get !a i
-
-      {!!c = {{{ $\sum_0^{ {\tt length~!a} - 1}$ get !!a i }}} }
+    { !c = 0 }
+    for i = 0 to Array.length a - 1 do
+      { !c = {{{ $\sum_0^{i-1}$ }}}  a.(i) }
+      c := !c + a.(i)
+    done
+    { !c = {{{ $\sum_0^{ {\tt Array.length~a} - 1}$ }}}  a.(i) }
 \end{who}
-
-Here, assuming we model the array in the programming language by a reference
-to a functional array, with functions {\tt length} and {\tt get} with the
-obvious meaning. Also, in the logic we use a derefence operator {\tt !!}
-instead of the usual {\tt !}. In this case, our program is a {\tt for} loop,
-so it needs an {\em invariant annotation}, indicating which relation is
-preserved by the loop body. The equalities that need to be proved are very
-simple. Also, it is obvious that the program modifies exactly the references
-$r$ and $c$, and this fact can be formally confirmed by an effect analysis.
+Our program is a {\tt for} loop, so it requires an {\em invariant
+annotation}, indicating which relation is preserved by the loop body. The
+equalities that need to be proved are very simple. Also, it is obvious that
+the program only modifies the reference {\tt c}, and this fact can be
+formally confirmed by an effect analysis.
 
 A common reaction of a functional programmer to a program like
-$\mathit{sum}$ is to factor out the structure of the recursion, which
-may be useful in other situations, in a higher-order function. The
-more specific program is then recovered by function application. In this case,
-the more general function is of course simply {\tt Array.iter}:
+{\tt sum} is to factor out the structure of the recursion, which
+may be useful in other situations, in a higher-order function. In this case,
+the more general function is the following iterator over arrays:
 \begin{ocaml}
@@
       f a.(i)
-
+    done
+\end{ocaml}
+The more specific program is then recovered by function application. 
+\begin{ocaml}
   let sum () =
@@
 thing. However, how do we prove the correctness of the second one? One
-possibility is to fix the future usage of ${\tt iter}$ by
-specifying what exactly is the specification of its argument {\tt f}:
+possibility is to impose the future usage of ${\tt iter}$ by
+specifying a particular specification for its argument {\tt f}:
 \begin{who}
@@
     (f : (x : int) ->{c} 
-        { !!c = {{{ $\sum_0^{i-1}$ get !!a i }}} } 
+        { !c = {{{ $\sum_0^{i-1}$ }}} a.(i) } 
         unit 
-        { !!c = {{{ $\sum_0^{length !a - 1}$ get !!a i }}} } ) 
+        { !c = {{{ $\sum_0^{ {\tt Array.length~a} - 1}$ }}} a.(i) } ) 
     (a : int array) =
@@
 Here we impose precisely the specification of {\tt f}, and thus we can
-derive that {\tt sum} computes the right value,
+derive that {\tt sum} computes the expected value,
 using the same method as for the first-order case before. Note
 that, to be able to apply an effect analysis, we have to specify that
-$f$ modifies $c$, by giving it the type $\intml->^c\unitml$.
+$f$ modifies $c$, by giving it the type $\intml->^{\{c\}}\unitml$.
 
@@
 and can only compute a certain sum. If one wants to use {\tt iter} in another
-context, one has to copy the code, give a new specification adapted to the
+context, one has to copy the code, give it a new specification adapted to the
 application and prove this new specification again. In addition, in order to
@@
 
-There are two obstacles related to higher-order functions to be
-overcome in order to achieve modularity on the level of specification
-in the presence of side effects. We must be able to refer to the
-effect of a function parameter, and we must be able to refer to the
-specification (pre- and postconditions) of a function parameter, in a
-generic way.
+There are two obstacles related to higher-order functions to be overcome in
+order to achieve modularity on the level of specification in the presence of
+side effects. We must be able to refer to the effect of a function parameter,
+and we must be able to refer to its specification (pre- and postconditions),
+in a generic way.
 
 The first obstacle has a well known solution using a type and effect
-system~\cite{talpin94taed} with effect variables: effect polymorphism
-does exactly what we want here. Using effect polymorphism, we can say
-that {\tt f} (the argument of {\tt iter}) has type
-$\intml->^e\unitml$, where $e$ is an {\em effect variable}, which
-amounts to say that {\tt f} has a {\em certain} effect $e$. The total
-effect of {\tt iter} is then $e$, meaning {\tt iter}
+system~\cite{talpin94taed} with effect variables: effect polymorphism does
+exactly what we want here. Using effect polymorphism, we can say that {\tt f}
+(the argument of {\tt iter}) has type $α->^e\unitml$, whera $α$ is the type of
+the array elements and where $e$ is an {\em effect variable}, which amounts to
+say that {\tt f} has a {\em certain} effect $e$. The complete type of {\tt
+iter} is thus:
+\begin{displaymath}
+  ∀α∀e. (α ->^e \unitml) ->^\emptyset α~\mathtt{array} ->^e \unitml
+\end{displaymath}
+The total effect of {\tt iter} is then $e$, meaning {\tt iter}
 modifies whatever {\tt f} modifies.
@@
 of the function. The postcondition takes the argument and return value, but
-also the states before {\em and} after the function call. Th so-called
-two-state postconditions are a way to get rid of most uses of auxiliary
-variables. However, the pre- and postcondition of {\tt f} may not refer to
-arbitrary parts of the state, but only the part {\tt f} reads or modifies.
-With these conventions, one can now write a generic specification for the {\tt
-iter} function, see ~Fig.~\ref{fig:iter}.
+also the states before {\em and} after the function call. Two-state
+postconditions are a standard way to get rid of most auxiliary variables.
+However, the pre- and postcondition of {\tt f} may not refer to arbitrary
+parts of the state, but only the part {\tt f} reads or modifies.  With these
+conventions, one can now write a generic implementation and specification for
+the {\tt iter} function. It is given~Fig.~\ref{fig:iter}, in the input syntax
+of our prototype \who. There are a few syntactic differences with \ocaml, in
+particular for arrays.
+
+%Here, assuming we model the array in the programming language by a reference
+%to a functional array, with functions {\tt length} and {\tt get} with the
+%obvious meaning. Also, in the logic we use a derefence operator {\tt !!}
+%instead of the usual {\tt !}.
 
@@
 
-There are several new elements here. First of all, we use effect polymorphism
-to introduce an effect variable {\tt e} standing for the effect of {\tt f}.
-Also, {\tt iter} has an additional argument {\tt inv}, the {\em invariant},
-taking as argument the portion of the state modified by f and the integer
-indicating the current iteration count. This is a logical argument, which
-means that it cannot be used in the program. Also, {\tt inv} is a {\em pure}
-function, which is indicated by the arrows without effect annotation. The
-postcondition is now quite clear: we simply want to establish the invariant
-for the final state (the keyword {\tt cur} for the current state of the
-postcondition) and the length of the array, indicating that we have iterated
-over the entire array. It is also easy to see that the invariant {\tt inv} is
-precisely the invariant of the {\tt for} loop. However, the precondition is a
-bit complicated: it states that we need the invariant to hold initially (for
-the current state of the precondition and index $0$) and we need the invariant
-to be preserved by the function call. This is expressed by quantifying over
-two states, saying that the invariant has to imply the precondition of {\tt f}
-and the postcondition has to imply the invariant for the next array index.
-With a bit of syntactic sugar, this could be expressed much more concisely
-using a Hoare triple:
+There are several new elements here. First of all, we use effect polymorphism.
+An effect variable {\tt e}, standing for the effect of {\tt f}, is introduced
+by the notation ${\mathtt [e]}$.  {\tt iter} also has an additional argument
+{\tt inv}, the {\em invariant}, taking as argument the portion of the state
+modified by {\tt f} and the integer indicating the current iteration count.
+This is a logical argument, which means that it cannot be used in the program.
+Also, {\tt inv} is a {\em pure} function, which is indicated by the arrows
+without effect annotation. The postcondition is now quite clear: we simply
+want to establish the invariant for the final state (the keyword {\tt cur}
+stands for the current state of the postcondition) and the length of the
+array, indicating that we have iterated over the entire array. It is also easy
+to see that the invariant {\tt inv} is precisely the invariant of the {\tt
+for} loop.  However, the precondition is a bit complicated: it states that we
+need the invariant to hold initially (for the current state of the
+precondition and index $0$) and we need the invariant to be preserved by the
+function call.  This is expressed by quantifying over two states, saying that
+the invariant has to imply the precondition of {\tt f} and the postcondition
+has to imply the invariant for the next array index.  With a bit of syntactic
+sugar, this could be expressed much more concisely using a Hoare triple:
 \begin{who} 
-  { inv cur i} f (get i !!a) {inv cur (i+1)}
+  { inv cur i } f (get i !!a) { inv cur (i+1) } 
 \end{who}
 
-The {\tt iter} function and its specification raise a question: is the
-specification correct if {\tt f} modifies the array {\tt a}, which is required
-for the good behavior of {\tt iter}? The answer is clearly no; if {\tt f}
-messes up the contents of {\tt a}, {\tt iter} may not actually maintain the
-invariant.  This is an instance of the {\em aliasing problem} (see, for
-example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
-does not support aliasing. In this paper we make the same choice and disallow
-aliasing in the programming language we introduce. Under this assumption, the
-specification we gave is correct.
-
-Entirely excluding aliasing renders our system inappropriate for
-reasoning about shared mutable structures. However, it is interesting
-to see that our alias exclusion restriction, in many cases,
-corresponds to a natural condition imposed on higher-order functions
-that operate on some data structure. It is a common
-requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration
-functions, for example, that the functional argument must not modify
-the structure to be iterated on. In our system, this fact follows from
-the restriction on effect instantiations, as will be seen in
-section~\ref{sec:syntax}. In the example, the array {\tt a}
-of the iterating function {\tt iter} must not be modified by its
-argument.
+% TODO rewrite
+%The {\tt iter} function and its specification raise a question: is the
+%specification correct if {\tt f} modifies the array {\tt a}, which is required
+%for the good behavior of {\tt iter}? The answer is clearly no; if {\tt f}
+%messes up the contents of {\tt a}, {\tt iter} may not actually maintain the
+%invariant.  This is an instance of the {\em aliasing problem} (see, for
+%example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
+%does not support aliasing. In this paper we make the same choice and disallow
+%aliasing in the programming language we introduce. Under this assumption, the
+%specification we gave is correct. 
+%
+%Entirely excluding aliasing renders our system inappropriate for
+%reasoning about shared mutable structures. However, it is interesting
+%to see that our alias exclusion restriction, in many cases,
+%corresponds to a natural condition imposed on higher-order functions
+%that operate on some data structure. It is a common
+%requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration
+%functions, for example, that the functional argument must not modify
+%the structure to be iterated on. In our system, this fact follows from
+%the restriction on effect instantiations, as will be seen in
+%section~\ref{sec:syntax}. In the example, the array {\tt a}
+%of the iterating function {\tt iter} must not be modified by its
+%argument.
 
@@
 variable stands for an unknown effect. Effects are used to specify
-of memory locations an expression can write. They are also used in function
+the memory locations an expression can write. They are also used in function
 types to describe the memory locations which may be written if the function is
@@
 wrt. the typing relation $|-_l$ described later on. Finally, pure applications
-between values are considered as values as well, as they don't have any
+between values are considered to be values as well, as they don't have any
 effect.
@@
 are \letml-introduced. This is not a restriction, as any program can be
-transformed an A-normal form.
+transformed into A-normal form.
 
@@
 effects. In the other case (rule {\sc Let}), no generalization takes place,
-but both terms can have effects. The {\sc Let} rule actually obliges both
+but both terms can have effects. The {\sc Let} rule actually requires both
 terms to have the same effect, but together with the {\sc Sub} rule, which can
@@
     { Γ |-_l t : \ceil{τ} }
-  } \and
+  } \qquad \and
   { \inferrule*[Left=LetLogic]
@@
 any logical connectives. These can of course be declared independently, for
-example the conjunction is just a function: \begin{equation*} /\ :  \propml ->
-\propml -> \propml \end{equation*} Actually, when equality is given, all
-logical operations (even quantification) can be defined in the
-logic~\cite{andrews86}.  In the remainder of the paper, we suppose
-predefined the usual logical connectives conjunction, disjunction,
+example the conjunction is just a function: 
+\begin{equation*} 
+  /\ :  \propml -> \propml -> \propml 
+\end{equation*} 
+Actually, when equality is given, all logical operations (even quantification)
+can be defined in the logic~\cite{andrews86}.  In the remainder of the paper,
+we suppose predefined the usual logical connectives conjunction, disjunction,
 implication and equivalence, as well as universal and existential
 quantification. We also assume all the arithmetic operations as well as the
-constructor $(t,t)$  and the accessors $fst$ and $snd$ for pairs. We will also
-use the names \preml\ and \postml\ for $fst$ and $snd$, respectively.
+constructor $(t,t)$  and the accessors $\mathit{fst}$ and $\mathit{snd}$ for
+pairs. We will also use the names \preml\ and \postml\ for $\mathit{fst}$ and
+$\mathit{snd}$, respectively.
 
@@
 We haven't mentioned references yet, but all we need to support them is to add
-the three usual functions $ref, :=, !$ for reference creation, assignment and
+the three usual functions $\mathit{ref}, :=, !$ for reference creation, assignment and
 lookup, respectively, to the environment. In our system, these functions have
@@
 \begin{eqnarray*}
-  ref &:& ∀αρ.α ->^ρ α~\refml_ρ\\
+  \mathit{ref} &:& ∀αρ.α ->^ρ α~\refml_ρ\\
   := &:& ∀αρ.α~\refml_ρ ->^\emptyset α ->^ρ \unitml\\
@@
 A way out is to restrict programs to alias free programs. Fortunately, it is
-easy to modify the presented typing rules to render all programs with aliasing
+easy to modify the typing rules to render all programs with aliasing
 ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
-to identify memory locations entirely; each region shall describe precisely
-one memory location. Starting from this assumption, a program can actually
+in such a way that there is exactly one memory location in each region.
+Starting from this assumption, a program can actually
 exhibit {\em two} forms of aliasing. In the first one, it may use several
@@
 this, we add a side condition to the instantiation: the substitution $σ$ must
-be {\em compatible} with the type $τ$. To understand what compatibility means,
-consider an {\em atomic} region substitution (whose domain contains only one
-variable) $[ρ|->ρ']$. It is compatible with any type which does not contain
-$ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2 e'$] is compatible with
-any type which does not contain $ ρ1 , ρ2 $ nor $e'$.  Expressed more
-generally, an atomic substitution is compatible with a type if the type does
-not contain any region or effect variables which are in the range of the
-substitution. Finally, a {\em composite} substitution (whose domain contains more
-than one variable) is compatible
-with a type if each atomic substitution is compatible with the result of the
-application of the previous substitution. In this way, it becomes impossible
-to have different names for a single reference cell.
+be {\em compatible} with the type $τ$, written $σ\sim τ$. To understand what
+compatibility means, consider an {\em atomic} region substitution (whose
+domain contains only one variable) $[ρ|->ρ']$. It is compatible with any type
+which does not contain $ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2
+e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
+Expressed more generally, an atomic substitution is compatible with a type if
+the type does not contain any region or effect variables which are in the
+range of the substitution. Finally, a {\em composite} substitution (whose
+domain contains more than one variable) is compatible with a type if each
+atomic substitution is compatible with the result of the application of the
+previous substitution. In this way, it becomes impossible to have different
+names for a single reference cell.
 
 The second form of aliasing is when we use the same name for {\em different}
-memory locations. This can happen when we allow the user to call $ref$ several
-times with the same region instantiation. A simple possibility is to attach
-the suppress the $ref$ function entirely and to replace it by a \letml\refml\ 
-construct which combines name creation and memory allocation. In this way,
-there can of course be only one reference per name. On the other hand, because
-of the side condition $ρ\notin τ$, the reference cannot escape the scope of
-its \letml\refml\ construct, and so, a function cannot return a reference
-which has been created in its body. This restriction could be leveraged a bit if
-one separates name creation and allocation, while still maintaining that one
-can only allocate a single reference in each region. This is how it is done in
-the correctness proof of this restriction. The presented \letml\refml\
-construct is of course a special case.
+memory locations. This can happen when we allow the user to call
+$\mathit{ref}$ several times with the same region instantiation. A simple
+possibility is to  suppress the $\mathit{ref}$ function entirely and to
+replace it by a \letml\refml\ construct which combines name creation and
+memory allocation. In this way, there can of course be only one reference per
+name. On the other hand, because of the side condition $ρ\notin τ$, the
+reference cannot escape the scope of its \letml\refml\ construct, and so, a
+function cannot return a reference which has been created in its body. This
+restriction could be leveraged a bit if one separates name creation and
+allocation, while still maintaining that one can only allocate a single
+reference in each region. This is how it is done in the correctness proof of
+this restriction. The presented \letml\refml\ construct is of course a special
+case.
 %TODO now, all operations on maps can be statically resolved.
@@
 the region calculus of Tofte/Talpin~\cite{tofte97ic}.
-
-\section{A Simple Weakest Preconditions Calculus}
+%TODO ref vers l'annexe -  remarque sur la difference
+
+\section{A Weakest Preconditions Calculus}
 \label{sec:wp}
@@
 The aim of the weakest precondition calculus is to obtain proof obligations
-from the annotated program. To do this, the calculus starts from a formula
+from annotated programs. To do this, the calculus starts from a formula
 which holds, or is supposed to hold, at the end of the program, and walks
@@
 hold at the beginning of the program (hence the name {\em weakest
-precondition}.
+precondition}).
 
@@
 \letml-binding in the language. In the other case, the \wpre\ of $ t2 $ becomes
-the argument of the \wpre of $ t1 $ (we denote $ε$ the effect of $ t1 $ and $
+the argument of the \wpre\ of $ t1 $ (we denote $ε$ the effect of $ t1 $ and $
 t2 $). The rule for \ifml-expressions is straightforward. The rule for

### 2009-10-06 papers/esop2010/main.tex
remarques JC - 2

@@
 \begin{document}
-\title{\mbox{Specifying Imperative Higher-Order Programs}}
+\title{\mbox{Verifying Imperative Higher-Order Programs}}
 
@@
 
-%Here, assuming we model the array in the programming language by a reference
-%to a functional array, with functions {\tt length} and {\tt get} with the
-%obvious meaning. Also, in the logic we use a derefence operator {\tt !!}
-%instead of the usual {\tt !}.
-
 \begin{figure}[tbp]
@@
 
-% TODO rewrite
-%The {\tt iter} function and its specification raise a question: is the
-%specification correct if {\tt f} modifies the array {\tt a}, which is required
-%for the good behavior of {\tt iter}? The answer is clearly no; if {\tt f}
-%messes up the contents of {\tt a}, {\tt iter} may not actually maintain the
-%invariant.  This is an instance of the {\em aliasing problem} (see, for
-%example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
-%does not support aliasing. In this paper we make the same choice and disallow
-%aliasing in the programming language we introduce. Under this assumption, the
-%specification we gave is correct. 
-%
-%Entirely excluding aliasing renders our system inappropriate for
-%reasoning about shared mutable structures. However, it is interesting
-%to see that our alias exclusion restriction, in many cases,
-%corresponds to a natural condition imposed on higher-order functions
-%that operate on some data structure. It is a common
-%requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration
-%functions, for example, that the functional argument must not modify
-%the structure to be iterated on. In our system, this fact follows from
-%the restriction on effect instantiations, as will be seen in
-%section~\ref{sec:syntax}. In the example, the array {\tt a}
-%of the iterating function {\tt iter} must not be modified by its
-%argument.
+The specification of {\tt iter} is correct, but it will be difficult to prove
+in general, because it is not clear wether {\tt f} modifies the array or not,
+or more precisely, if the effect variable {\tt e} does contain {\tt a} or not.
+This is an instance of the {\em aliasing problem} (see, for
+example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
+is not well-suited for programs with aliasing. Without special techniques (for
+example {\em separation logic}), a large number of equalities and disequalities
+are needed to reason about potentially aliased references.
+
+To avoid this situation, we additionally introduce a simple means to exclude
+programs with aliasing. The instantiation of effect variables is restricted to
+the ones which do not create aliasing situations. Our system, with aliasing
+exclusion enabled, is inappropriate for reasoning about shared mutable
+structures. However, it is interesting to see that this alias exclusion
+restriction, in many cases, corresponds to a natural condition imposed on
+higher-order functions that operate on some data structure. It is a common
+requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration functions,
+for example, that the functional argument must not modify the structure to be
+iterated on. In our system, this fact follows from the restriction on effect
+instantiations, as will be seen in section~\ref{sec:syntax}. In the example,
+the array {\tt a} of the iterating function {\tt iter} must not be modified by
+its argument.
 
@@
     \wpre_s(\ifml~ t1 ~\thenml~ t2 ~\elseml~ t3 , q) &=&
-      \ifml~ \ceil{ t1 }~\thenml~ \wpre_s( t2 , q))~\elseml~\wpre_s( t3 , q)) \\
+      \ifml~ \ceil{ t1 }~\thenml~ \wpre_s( t2 , q)~\elseml~\wpre_s( t3 , q) \\
       \wpre_s( (t : ε <: ε'), q) &=& \wpre_{s|_ε}(t, λs':\efft{ε}.q~(s\oplus s'))
@@
 bound term is a value, we can translate the \letml-binding in the program by a
-\letml-binding in the language. In the other case, the \wpre\ of $ t2 $ becomes
+\letml-binding in the logic. In the other case, the \wpre\ of $ t2 $ becomes
 the argument of the \wpre\ of $ t1 $ (we denote $ε$ the effect of $ t1 $ and $
@@
 
-let memo [e] (f0 : int -> int) (f : int ->{} int) (x : int) = 
-  {pre f x /\ stores f0 !!table}
+let memo [e] (f0 : int -> int) (x : int) = 
+  {stores f0 !!table}
   if mem x !table then get x !table
   else
-    let z = f x in
+    let z = f0 x in
     table := set x z !table;
     z
-  {r : post f x r}
+  {r : r = f0 x /\ stores f0 !!table}
 \end{who}
@@
 Our first example is a simple memoization function {\tt memo} for (pure)
-functions.  It takes as an argument a function (from {\tt int} to {\tt int}
-here) and an integer and returns the same result as the functional argument
-would; however, before executing the argument, it will look up the integer in
-a table to see if the function hasn't been called yet on this integer. If so,
-the table's value is returned, otherwise we call the function, store the
-result in the table, and return it. Fig~\ref{fig:memo} shows the code along
-with the specification. The specification is very simple; it states that {\tt
-memo} will return a value that verifies the postcondition of {\tt f}, as long
-is initially we have the precondition of f and as long as the table stores
-only valid key/value pairs. As we are not interested in the exact way of the
-implementation of the map, we simply model it here by a reference to a
-functional map with get/set functions.
+functions.  It takes as an argument a function {\tt f} (from {\tt int} to {\tt
+int} here) and an integer and returns the same result as {\tt f} would;
+however, before executing the argument, it looks up the integer in a table to
+see if the function hasn't been called yet on this integer. If so, the table's
+value is returned, otherwise we call the function, store the result in the
+table, and return it. Fig~\ref{fig:memo} shows the code along with the
+specification. Since we are not interested in the implementation of the table,
+we simply model it by a reference to a functional map with mem/get/set functions.
+
+The specification of {\tt memo} is very simple. The only precondition is that
+the table should only contain bindings $y |-> {\mathtt f}~y$, which is
+expressend by the {\tt stores} predicate. The postcondition states that the
+result is equal to {\tt f0~x} and that the table still verifies the {\tt
+stores} predicate.
 
@@
 \begin{who}
-  let realizes 
+  let realizes = .. .
   let ymemo (f0 : int -> int) 
@@
         { realizes f0 k} ff k { r : realizes f0 r} }
-  let rec f (x : int) : int ->{table} int =
+    let rec f (x : int) =
       { stores f0 !!table }
@@
 
-A prototype implementation of the presented system by the name ``\who''
-exists~\cite{KanigFilliatre09wml} and is freely available. It implements the
+A prototype implementation of the presented system, called ``\who'',
+ is freely available~\cite{KanigFilliatre09wml}%TODO \footnote
+. It implements the
 described language with type, region and effect polymorphism, higher-order
@@
 is syntax-directed and thus entirely automated. Their logic of proof
-obligations is non-standard, while we obtain obligations in higher-order
-logic. They deal with aliasing using equalities and disequalities between
-pointers, which we chose not to do. They can reason about the classic {\em
+obligations is non-standard, while we obtain obligations in standard higher-order
+logic.
+
+They deal with aliasing using equalities and disequalities between
+pointers, which we chose not to do.
+
+They can reason about the classic {\em
 Landin's knot}, which consists in obtaining recursion by storing a function in
@@
 explain in the next section how we plan to achieve this.
-
-%On the other hand, Ynot shares the
-%Coq's main weakness: lack of automation. While we haven't yet presented a
-%platform to automatically discharge generated proof obligations, this seems
-%easily feasible, along the same vein as Régis-Gianas and
-%Pottier~\cite{regis-gianas-pottier-08}, who argue that first-order SMT
-%solvers~\cite{RanTin-SMTLIB} are suitable even for higher-order proof
-%obligations. They also present an encoding scheme to support their claim.
-%Their practical results are promising. The Ynot system can not use this kind
-%of automation because it uses separation logic.  Automated provers for
-%separation logic exist, but we are not aware of any attempts to use these
-%tools in a higher-order setting.
-
-%TODO also say that an approach like Ynot ou Coq doesn't give any help
-%for the mechanical part of the proof.
 
@@
 Systems for verification of first-order programs like
-Why~\cite{Filliatre00a} or Spec\#~\cite{BarnettLS04} are the other
-starting points of this work. These systems generally deal very well
-with usual features of first-order programs, for example arrays, use
-SMT solvers to discharge proof obligations and strive for the best
-possible automation. We try to improve on this work by adding
-higher-order features, trying to maintain the degree of automation and
-ease of use.
+Why~\cite{Filliatre00a,filliatre07cav} or Spec\#~\cite{BarnettLS04} are the
+other starting points of this work. These systems generally deal very well
+with usual features of first-order programs, for example arrays, use SMT
+solvers to discharge proof obligations and strive for the best possible
+automation. We improve on this work by adding higher-order features,
+while trying to maintain the degree of automation and ease of use.
 
@@
 
-We see two possibilities to add the possibility to reason about aliasing
-programs. The first one is to modify the effect system itself to allow for
-aliasing. In the setting with the aliasing restriction, our regions actually
-become {\em singleton regions}, regions containing exactly one memory
-location. Other, more powerful effect systems allow for {\em group regions},
-with the possibility to temporarily {\em focus } on a particular memory cell
-in that region. We see no fundamental difficulty in adopting such an effect
-system, but haven't worked out the details. A second possibility is to deal
-with aliasing via equalities and disequalities over pointers, using regions to
-obtain some separation automatically. On itself, we believe that this
-technique does not scale, due to the explosion of needed disequalities.
-However, in combination with other separation techniques, for example the
-Burstall-Bornat memory model which separates automatically objects of
-different structural types, this approach has worked well applied to
-first-order programs.
+We see two possibilities to deal with aliasing beyond plain usage of pointer
+disequalities. The first one is to modify the effect system itself to allow
+more precise tracking of aliasing. In the setting with the aliasing
+restriction, our regions actually become {\em singleton regions}, regions
+containing exactly one memory location. Other, more powerful effect systems
+allow for {\em group regions}, with the possibility to temporarily {\em focus
+} on a particular memory cell in that region. We see no fundamental difficulty
+in adopting such an effect system, but haven't worked out the details. A
+second possibility is to deal with aliasing via equalities and disequalities
+over pointers, using regions to obtain some separation automatically. On
+itself, we believe that this technique does not scale, due to the explosion of
+needed disequalities.  However, in combination with other separation
+techniques, for example the Burstall-Bornat memory model~\cite{bornat00mpc}
+which separates automatically objects of different structural types, this
+approach has worked well applied to first-order programs.
 
@@
 features in the programming language, such as recursive types and algebraic
-data types. As long as so defined types verify a {\em positivity
+data types. As long as defined types verify a {\em positivity
 condition}~\cite{paulintlca93}, these extensions would be completely

### 2009-10-06 papers/esop2010/main.tex
go

@@
 
-%external TODO : proofs for Hashtbl
 %external TODO : proofs of system
@@
 
-% TODO TODO
-%Finally, we introduce two symbols
-%\begin{eqnarray*}
-%  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
-%  |_{ε} &:& ∀ e. \efft{ e ε } -> \efft{ ε }.
-%  !! & : & ∀αρ e. α~\refml_ρ -> \efft{ ρ e } -> α
-%\end{eqnarray*}
-%The first one denotes a combining operation for stores; the resulting store
-%contains all the mappings of the second store, as well as all the mappings of
-%the first one, as long as they are not present in the second one.
-%union of the two arguments, in the sense that its domain is the union of its
-%arguments.
+Finally, we introduce three symbols
+\begin{eqnarray*}
+  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
+  |_{ε} &:& ∀e. \efft{ e ε } -> \efft{ ε }\\
+  {\tt !!} & : & ∀αρe. α~\refml_ρ -> \efft{ ρ e } -> α.
+\end{eqnarray*}
+The first one denotes a combining operation for stores; the resulting store
+contains all the mappings of the second store, as well as all the mappings of
+the first one, as long as they are not present in the second one. The second
+operations is indexed by an effect and it restricts a store to a certain
+effect. Finally, the double bang {\tt !!} is the logic equivalent to the
+lookup operator {\tt !}. It is pure and it is always with respect to a certain
+state.
 
@@
 the other hand, in its presented form, and because of the presence of
-aliasing, one would constantly need equalities and inequalities between
+aliasing, one would constantly need equalities and disequalities between
 reference variables to be able to reason about a program.
@@
 case.
-%TODO now, all operations on maps can be statically resolved.
+
+In alias-free programs, the operations $\oplus$ and $|_ε$ on stores we
+previously introduced can be executed {\em statically}, i.e. just by looking
+at the domains of the stores. Also, we statically know which value (or which
+named component) the accessor {\tt !!} will get from a store.
 
@@
 let stores f0 m =
-  forall x : Z, mem x m = true -> get x m = f0 x
+  forall x : int, mem x m = true -> get x m = f0 x
 
@@
 
-let memo [e] (f0 : int -> int) (x : int) = 
+let memo (f0 : int -> int) (x : int) = 
   {stores f0 !!table}
@@
 
-\subsection{ymemo}
-
-TODO %TODO
+\subsection{A more powerful memoization function}
+
 \begin{figure}[tpb]
 \begin{who}
-  let realizes = .. .
+  let realizes (f0 : int -> int) (f : [[ int ->{table} int ]]) =
+    forall (x : int),
+    { stores f0 !!table } f x { r : r = f0 x /\ stores f0 !!table }
+
   let ymemo (f0 : int -> int) 
@@
 
+Our second, more complicated example is the recursive version of {\tt memo},
+dubbed {\tt ymemo}. We want to be able to define a recursive function whose
+recursive calls are automatically memoized. The basic idea of {\tt ymemo} is
+analogous to the fixpoint combinator in λ-calculus. As an example, here is how
+one would implement a memoized Fibonacci function
+using ymemo (ignoring the logic argument $f0$):
+\begin{ocaml}
+  let fib = 
+    ymemo (fun g x -> if x <= 1 then x else g (x-1) + g (x-2))
+\end{ocaml}
+The anonymous function uses its first argument {\tt f} to do recursive calls.
+
+{\tt ymemo} achieves this by defining a recursive function {\tt f} inside its
+body (see Fig~\ref{fig:ymemo}). This function, when called with an integer
+argument {\tt x} first checks if the table contains {\tt x} and if not, calls
+{\tt ff} {\em with {\tt f} itself} and x as argument. As before, it stores the
+obtained result in the table and returns it. Now, the return value of {\tt
+ymemo} is precisely this recursive function {\tt f}. The only difference with
+the equivalent \ocaml\ code is the usage of a logical argument {\tt f0} which
+stands for the (mathematical) function to be defined.
+
+Let's move on to the specification of {\tt ymemo}. We first define the {\tt
+realizes} predicate between a pure function {\tt f0} and an impure one
+modifying the table. This predicate states that {\tt f} {\em realizes}, i.e.
+returns the same result as, {\tt f0} and additionally, if the table stored
+valid key/value pairs, this is also the case after executing {\tt f}.
+Actually, this is precisely the specification of {\tt memo}, so we could
+easily prove {\tt realizes f0~(memo f0)} for any pure {\tt f0}. The definition
+of the predicate uses the syntactic sugar for Hoare triples we introduced in
+section~\ref{sec:genpic}.
+
+The recursive function {\tt f} in the body of {\tt ymemo} has the same spec as
+{\tt memo}, and indeed here it is trivial to prove the postcondition of {\tt
+ymemo}, which states that its return value realizes {\tt f0}. Now, in order to
+prove that the specification of {\tt f} is indeed correct, the only
+information we need is that {\tt ff k} realizes {\tt f0} if {\tt k} is a
+function which realizes {\tt f0}.
+
 \subsection{All examples - an overview}
@@
     {\tt Array} & 69 loc & 170 loc  & 320 lop  \\ \hline
-    {\tt Hashtbl} (incomplete) & & &  \\ \hline
     Koda-Ruskey~\cite{KanigFilliatre09wml} & 10 loc & 30 loc & $\sim$700 lop  \\ \hline
@@
 
-The proof development concerning the {\tt Hashtbl} module is ongoing work. The
-difficult part of this development is the model of the \ocaml\ hash table; In
-\ocaml, when one overwrites a previously existing entry in the table, the old
-value still exists, and is only {\em shadowed} by the new key/value pair. If
-one calls {\tt Hashtbl.remove} using this key, the old value reappears. Also,
-if several bindings for a key exist, the key may be encountered several times,
-during a call of {\tt Hashtbl.iter}, for example. A possible way of modeling
-this is an association list instead of a simple mapping, where keys in front
-of the list can shadow the same key further down in the list. 
-% TODO Currently, only
+%The proof development concerning the {\tt Hashtbl} module is ongoing work. The
+%difficult part of this development is the model of the \ocaml\ hash table; In
+%\ocaml, when one overwrites a previously existing entry in the table, the old
+%value still exists, and is only {\em shadowed} by the new key/value pair. If
+%one calls {\tt Hashtbl.remove} using this key, the old value reappears. Also,
+%if several bindings for a key exist, the key may be encountered several times,
+%during a call of {\tt Hashtbl.iter}, for example. A possible way of modeling
+%this is an association list instead of a simple mapping, where keys in front
+%of the list can shadow the same key further down in the list. 
+% Currently, only
 %two functions of this module have been proved correct.

### 2009-10-06 papers/esop2010/appendix/appendix.tex
go

%; whizzy document
\documentclass[a4paper]{article}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{stmaryrd} % for semantic brackets
\usepackage{mathpartir} % for inference rules
\input{utf8symbols.tex}
\input{thmdefs.tex}
\input{generaldefs.tex}

\author{Johannes Kanig}
\title{Technical Appendix}
\date{\today}
%TODO describe label substitutions clearly
%TODO describe all kinds of substitutions clearly
%TODO introduce new syntax for records : <>, only labels accessible
%... is it reasonable ?
% traduire vers des n-uplets plutôt que des ->
% utiliser la syntaxe concrète de l'outil
%TODO invent new notation!

\input{macros.tex}
\newcommand{\bmark}[1]{\textbf{(#1)}}
\newcommand{\pre}{{\bf pre}}
\newcommand{\post}{{\bf post}}
\newcommand{\wpre}{{\bf wp}}
\newcommand{\proptype}{{\bf prop}}
\newcommand{\rec}{{\bf rec}}
\newcommand{\void}{{\tt ()}}
\newcommand{\unit}{{\tt unit}}
\newcommand{\inttype}{{\tt int}} 
\newcommand{\letst}{{\bf let }}
\newcommand{\inst}{{\bf in }}
\newcommand{\true}{{\tt true}}
\newcommand{\false}{{\tt false}}
\newcommand{\bool}{{\tt bool}} 
\newcommand{\ifst}{{\bf if }}
\newcommand{\thenst}{{\bf then }}
\newcommand{\elsest}{{\bf else }}
\newcommand{\evlist}{\alist{\ev}}
\newcommand{\False}{\bot}
\newcommand{\True}{\top}
\newcommand{\reftype}{{\bf ref}}

\begin{document}
\maketitle

\section{The Aliasing Restriction}

In this section, we deliver the correctness proof for restriction to singleton
regions. We show here that each region can in fact only contain a single
reference. Instead of the {\bf letref}-construct as in the paper, we develop a
more general approach using a {\bf letreg}-construct. The principal idea is
that to call the function $\mathit{ref}$ to create a reference in a certain
region $ρ$ , one needs a capacity, a special token, denoted $\{ ρ\}$, which is
consumed by the function call. These capacities are treated linearly, i.e.
they cannot be duplicated. The {\bf letreg} construct delimits not only the
scope of a region variable, but also gives a single creation capacity. So the
whole following development basically ensures that no region capacity can be
duplicated, and then we obtain for free that no region contains more than one
memory location. A region can be empty, though.

The {\bf letref}-construct can be trivially recovered of course:
\begin{equation*}
  \letst~x = \mathit{ref}~v~\inst~e \quad \equiv\quad 
   \letreg{ρ}{\mllet{x}{\mathit{ref}~\{ρ\}~v}{e}}
\end{equation*}

\paragraph{A guide to the proof.} After the usual definitions of syntax and
semantics, we define a splitting operation over typing environments which
takes care of the linear nature of creation capacities
(Sec.~\ref{sec:environments}). In Section~\ref{sec:typing} we prove many
auxiliary lemmas for the main theorem, which is stated in
Section~\ref{sec:subjectreduction}. Definition \ref{def:invariants} contains
the important property to be maintained: If a term is well-typed wrt. an
environment, then all regions are singleton regions, and if a creation
capacity on a region is still available, this region is empty. The desired
property follows.

\subsection{Syntax}

\begin{equation*}
  \begin{array}[]{llll}
  \mbox{Program Variables}& x & & \\
  \mbox{Region Variables} & \rho & &\\
  \mbox{Type Variables} & \alpha & &\\
  \mbox{Effect Variables} & \epsilon & &\\
  \mbox{Program Constants} & c & &\\
  \mbox{Values}& v & ::= & c \sep \tyapp{x}{\bar{\tau} \bar{\rho}\bar{\varepsilon}} \sep \lambda x:\tau.e \sep
  \regpr{\rho}{e} \sep y_\rho \\
  \mbox{Expressions} & e & ::= & v\sep  \napp{e}{e} \sep \rapp{e}{\rho} \sep \letreg{\rho}{e} \sep \\
  & & & (e : \varepsilon ) \sep \region{\rho}{e} \sep \mllet{\bar{\alpha} \bar{\rho} \bar{\epsilon}\ x}{v}{e}\\
  \mbox{Types} & \tau & ::= & \alpha \sep \iota \sep \tau \effarr{\varepsilon} \tau \sep  \refreg{\tau}{\rho}
  \sep \regty{\rho,\varepsilon}{\tau} \\ 
  \mbox{Type Schemes} & ο & ::= & \tau\sep  \forall\alpha.\sigma \sep \forall\rho.\sigma \sep \forall\epsilon.\sigma \\
  \mbox{Effects} & \varepsilon & ::= & \epsilon \sep \rho \sep \emptyset \sep \varepsilon \cup \varepsilon
  \end{array}
\end{equation*}

\subsection{Semantics}
\label{sec:semantics}

Before executing a term, effect annotations are erased.

In this overview, when we write $d\in s$ we mean $d\in dom(s)$ and
similar for $l\in s(d)$.
\begin{equation*}
   \begin{array}[t]{rrclr}
    \beta & s, (\lambda x:\tau.e)\ v & \basered & s, \esub{e}{x\mapsto v} & \\
    let & s, \mllet{\bar{\alpha} \bar{\rho} \bar{\epsilon}\ x}{v}{e} & \basered & s,
    \esub{e}{x\mapsto \Lambda\bar{\alpha}\bar{\rho}\bar{\epsilon}.v} \\
    \capreg{\rho} & s, \rapp{(\regpr{\rho}{e})}{\rho} & \basered & s,e & \\
    letreg & s, \letreg{\rho}{e} & \basered & \statesub{s}{\rho_1\mapsto \left\{
    \right\}}, \region{\rho_1}{\regsub{e}{\rho\mapsto \rho_1}} & \rho_1\notin s \\
    region & s, \region{\rho}{v} & \basered & s, v \\
    ref & s, \reft \left[ \xi_1 \dots \xi_n \right] \capreg{\rho_1}\  v &
    \basered & s\left[ \rho_1 \mapsto s(\rho_1)\left[ y\mapsto v \right]\right],
    y_{\rho_1} & \rho_1\in s, y\notin s(\rho_1)\\
    := & s, := \left[ \xi_1 \dots \xi_n \right] y_{\rho_1} v & \basered & s\left[ \rho_1 \mapsto s(\rho_1)\left[
    y \mapsto v \right] \right], \void & \rho_1\in s, y\in s(\rho_1)\\
    ! & s, ! \left[ \xi_1 \dots \xi_n \right] y_{\rho_1} & \basered & s, s(\rho_1)(y) & \rho_1\in s, y\in
    s(\rho_1) \\
    \rec & s, (\rec\ f\ v_1)\ v_2 & \basered & s, v_1\left[ f \mapsto (\rec\
    v_1)  \right] v_2
  \end{array}
\end{equation*}

The relation $\basered$ describes a single reduction step at the top of a
term. To lift this relation to arbitrary expressions, we define call-by-value
contexts:
\begin{equation*}
  E ::= \left[  \right] \sep E\ e \sep v\ E\sep E \left[ \xi \right] \sep
  \rapp{E}{p} \sep \region{\rho}{E} \sep \mllet{x}{E}{e}
\end{equation*}

Now we define the one step reduction relation $\onestep$ as follows:
\begin{mathpar}
  \inferrule* 
    { s, e \basered s', e' }
    { s, E\left[ e \right] \onestep s', E\left[ e' \right] }
\end{mathpar}

The general reduction relation $\reduction$ is defined as the reflexive and
transitive closure of $\onestep$.

\subsection{Environments}
\label{sec:environments}
\begin{eqnarray*}
  \Gamma & ::= & \emptyset \sep \Gamma, x : \tau \sep \Gamma, \chi \sep \Gamma, \capreg{\rho}\\
\end{eqnarray*}

\subsubsection*{Wellformedness}
We write $\chi\notin fv(\Gamma)$ to express that $\chi$ never occurs free in $\Gamma$. We write
$\chi\notin \Gamma$ to express that $\chi$ is not an {\em element} of the environment
$\Gamma$.

\begin{mathpar}
  (1)\ \cwf{\emptyset} \and
  {\inferrule* [left=(2)] {\cwf{\Gamma}\quad \chi\notin \Gamma} {\cwf{\Gamma,\chi}}}\and
  {\inferrule* [left=(3)] {\cwf{\Gamma}\quad \ctwf{\Gamma}{\rho} } 
                          {\cwf{\Gamma,\capreg{\rho}}}}
  \and
  {\inferrule* [left=(4)] {\cwf{\Gamma}\quad \ctwf{\Gamma}{\tau}} {\cwf{\Gamma,x:\tau}}}
  \and
  {\inferrule* [left=(5)] {\chi\in \Gamma}{\ctwf{\Gamma}{\chi}}}\and
  (6)\  \ctwf{\Gamma}{\iota}\and
  {\inferrule* [left=(7)] {\ctwf{\Gamma}{\tau_1}\quad\ctwf{\Gamma}{\tau_2}\quad\ctwf{\Gamma}{\varepsilon}}
  {\ctwf{\Gamma}{\tau_1\effarr{\varepsilon} \tau_2}}}\and
  {\inferrule* [left=(8)] {\ctwf{\Gamma,\bar{\chi}}{\tau}}{\ctwf{\Gamma}{\forall\bar{\chi}.\tau}}}\and
  {\inferrule* [left=(9)] {\ctwf{\Gamma}{\tau} \\ \ctwf{\Gamma}{\rho}}
      {\ctwf{\Gamma}{\refreg{\tau}{\rho}}}} \and
  {\inferrule* [left=(10)] {\ctwf{\Gamma}{\tau} \\ \ctwf{\Gamma}{\rho}}
      {\ctwf{\Gamma}{\regty{\rho}{\tau}}}} \and
  {\inferrule* [left=(11)] {\ctwf{\Gamma}{\varepsilon_1} \\ \ctwf{\Gamma}{\varepsilon_2}}
      {\ctwf{\Gamma}{\varepsilon_1\cup \varepsilon_2}}}
\end{mathpar}

The rules (1) to (4) concern wellformed environment construction, The rules (5) to (10)
concern wellformedness of types. Rule (11) concerns wellformedness of effects
and recycles rule (5) for effect and region variables.

\begin{prop}
  $\chi'\notin \Gamma$ and $\ctwf{\Gamma}{\xi}$ implies $\chi'\notin fv(\xi)$.
  \label{prop:cwffvty}
\end{prop}
\begin{proof}
  Proof by induction on the derivation of $\ctwf{\Gamma}{\xi}$.
  \begin{description}

    \item[Case (5)] If $\chi = \chi'$, then we have a contradiction between the
      premise of the rule and the hypothesis of our claim. In the other case
      we have $\chi'\notin fv(\chi)$.

    \item[Case (6)] There are no free variables in base types.

    \item[Cases (7), (9), (10), (11)] By induction hypotheses.

    \item[Case (8)] We can assume any $\chi$ of the rule to be different to the
      $\chi'$ in the claim. Now suppose $\chi'\notin \Gamma,\bar{\chi}$ and $\chi'\notin fv(\tau)$. We have
      $\chi'\notin \Gamma$, and $\chi'\notin fv(\tau)$ iff $\chi'\notin fv(\forall\bar{\chi}.\tau)$.
  \end{description}
\end{proof}

\begin{prop}
  \cwf{\Gamma,\chi',\Delta} and \ctwf{\Gamma}{\xi} imply $\chi'\notin fv(\Gamma)$ and
  \cwf{\Gamma,\tsub{\Delta}{\chi'\mapsto \xi}}.
  \label{prop:cwftysub}
\end{prop}
\begin{proof}
  We proceed by induction on the wellformedness derivation of \cwf{\Gamma,\chi',\Delta}. 
  \begin{description}

    \item[Case (1)] The rule states \cwf{\emptyset}, but this cannot be the
      last rule applied to establish \cwf{\Gamma,\chi',\Delta}.

    \item[Case (2)] If $\chi \neq \chi'$ then we can simply conclude by induction
      hypothesis. If $\chi = \chi'$, then either $\Delta=\emptyset$ and the side
      condition $\chi\notin\Gamma$ together with proposition \ref{prop:cwffvty} gives
      exactly the property we want to prove for the first claim, and the
      second claim is trivial since $\tsub{\emptyset}{\chi\mapsto \xi} =
      \emptyset$. Or $\Delta$ is not empty and that side condition cannot be true.

    \item[Case (3)] Either $\chi' = \rho$, in which case $\xi = \rho'$ and we conclude by
      the obvious fact $\ctwf{\Gamma,\tsub{\Delta}{\rho\mapsto \rho'}}{\capreg{\rho'}}$. In the case
      where $\chi' \neq \rho$, we conclude simply by induction hypothesis.

    \item[Case(4)] For the first claim, there is nothing to prove. For the
      second claim, we need to prove that $\ctwf{\Gamma,\chi',\Delta}{\tau}$ implies
      $\ctwf{\Gamma,\tsub{\Delta}{\chi'\mapsto \xi}}{\tsub{\tau}{\chi'\mapsto \xi}}$.

      \begin{description}

        \item[Case (5)] If $\chi = \chi'$, this amounts to proving
          \ctwf{\Gamma,\tsub{\Delta}{\chi\mapsto \chi}}{\xi}, which is obvious because of
          \ctwf{\Gamma}{\xi} by hypothesis.

        \item[Case (6)] This rule is independent of the environment.

        \item[Case (7),(9),(10),(11)] Simply by applying the induction
          hypotheses.

        \item[Case (8)] We can assume $\chi'\notin \bar{\chi}$, and thus conclude by
          induction hypothesis.

      \end{description}
  \end{description}
\end{proof}

\begin{prop}
  \cwf{\Gamma} and $\Delta$ is a prefix of $\Gamma$ implies \cwf{\Delta}.
  \label{prop:cwfprefix}
\end{prop}
\begin{proof}
  We prove the claim by induction over the wellformedness derivation for $\Gamma$. 
\begin{description}
  \item[Case (1)] This implies $\Gamma$ is empty and thus $\Delta$ is empty, so \cwf{\Delta}
    by (1).
  \item[Case (2)] 
    The conclusion of the rule is \cwf{\Gamma,\chi}(a); it's premises are
    $\cwf{\Gamma}$ (b) and $\chi\notin fv(\Gamma)$ (c). If $\Delta = \Gamma,\chi$, then conclude by (a).
    If this is not the case $\Delta$ is also a prefix of $\Gamma$, and by induction
    hypothesis and (b), we can conclude.
  \item[Cases (3) and (4)] completely analogous.
\end{description}
\end{proof}

\begin{prop}
  \label{prop:cwfomit}
  \cwf{\Gamma,x : \tau, \Delta} implies \cwf{\Gamma,\Delta}.
\end{prop}
\begin{proof}
  By induction over the wellformedness derivation for $\Gamma$.
  \begin{description}
    \item[Case (1)] impossible.
    \item[Cases (2), (3), (4) and (5)] By induction hypothesis.  
    \item[Case (6)] By proving that \ctwf{\Gamma,x : \tau, \Delta}{y : \tau'} implies
      \ctwf{\Gamma,\Delta}{y : \tau'}. This is obvious, as one can see that wellformedness
      of types depends only on the presence of type and region variables, and
      not on variable bindings in the environment.
  \end{description}

\end{proof}

\subsubsection*{Splitting}

We define a non-deterministic splitting operation on the environment. The
operation is entirely symmetric, so only the ``left''
case of $\oplus_4$ are given.

\begin{equation*}
  \begin{array}[]{lrcl}
  \oplus_1 & \emptyset \oplus \emptyset & = & \emptyset \\
  \oplus_2 & (\Gamma_1, \chi) \oplus (\Gamma_2, \chi) & = & (\Gamma_1 \oplus \Gamma_2),\chi \\
  \oplus_3 & (\Gamma_1, x : \tau) \oplus (\Gamma_2, x : \tau) & = & (\Gamma_1 \oplus \Gamma_2),x:\tau \\
  \oplus_4 & (\Gamma_1,\capreg{\rho}) \oplus \Gamma_2 & = & (\Gamma_1 \oplus
              \Gamma_2),\capreg{\rho}\qquad \capreg{\rho}\notin \Gamma_2 \\
  \end{array}
\end{equation*}

\begin{prop}
  \cwf{\Gamma_1} and \cwf{\Gamma_2} imply \cwf{\Gamma_1\oplus \Gamma_2}, if $\Gamma_1\oplus \Gamma_2$ is
  defined.
  \label{prop:cwfoplus}
\end{prop}
\begin{proof}
  By induction on the construction of $\Gamma_1\oplus \Gamma_2$.
  \begin{description}
    \item[Case $\oplus_1$] Obvious.

    \item[Case $\oplus_2$] The corresponding rule states $(\Gamma_1, \chi) \oplus
      (\Gamma_2, \chi) = (\Gamma_1 \oplus \Gamma_2),\chi$. By the premises and proposition
      \ref{prop:cwfprefix}, we have \cwf{\Gamma_1} and \cwf{\Gamma_2}, and by induction
      hypothesis we have \cwf{\Gamma_1\oplus \Gamma_2}. By the premises of the
      wellformedness derivations for \cwf{\Gamma_1,\chi} and \cwf{\Gamma_2,\chi} we obtain
      $\chi\notin fv(\Gamma_1)$ and $\chi\notin fv(\Gamma_2)$. As can be simply checked, this
      implies $\chi\notin fv(\Gamma_1 \oplus \Gamma_2)$, and by the rule (2) we conclude
      \cwf{(\Gamma_1\oplus \Gamma_2),\chi}.

   \item[Case $\oplus_3$] analogously, we can conclude $\cwf{\Gamma_1}$ and
     \cwf{\Gamma_2} and thus, by induction hypothesis, \cwf{\Gamma_1 \oplus \Gamma_2}. We are
     left to prove that if \ctwf{\Gamma_1}{\tau} then \ctwf{\Gamma_1 \oplus \Gamma_2}{\tau}, which
     is very easily proved by induction. We can conclude \cwf{(\Gamma_1 \oplus
     \Gamma_2), x:\tau} by (4).

   \item[Case $\oplus_4$ ] As before, we can derive $\cwf{\Gamma_1}$ and \cwf{\Gamma_2}
     and thus, by induction hypothesis, \cwf{\Gamma_1 \oplus \Gamma_2}. As in the case
     $\oplus_3$, we are left to prove \ctwf{\Gamma_1 \oplus \Gamma_2}{\rho}. This can
     easily be derived by induction. 
 \end{description}
\end{proof}

\begin{prop}
  \label{prop:oplusabsorb}
  If $\capreg{\_}\notin \Gamma_2$, and $\Gamma_1 \oplus \Gamma_2$ is defined, then $\Gamma_1
  \oplus \Gamma_2 = \Gamma_1$.
\begin{proof}
  By induction of the construction of $\Gamma_1 \oplus \Gamma_2$.
  \begin{description}
    \item[Case $\oplus_1$] There is nothing to prove.

    \item[Case $\oplus_2$] We have $\capreg{\_}\notin \Gamma_2, \chi$, which gives us
      $\capreg{\_}\notin \Gamma_2$ and thus $\Gamma_1\oplus \Gamma_2 = \Gamma_1$ by induction
      hypothesis, and finally $(\Gamma_1,\chi) \oplus (\Gamma_2,\chi) = (\Gamma_1 \oplus \Gamma_2), \chi =
      \Gamma_1, \chi$.

    \item[Cases $\oplus_3$ and $\oplus_4$] Similar.

    \item[Case $\oplus_5$ ] Similar as well. Because of symmetry, we also
      should consider the case where $\Gamma_1$ and $\Gamma_2$ are inversed, but this
      leads to the contradiction $\capreg{\_}\notin \Gamma_2,\capreg{p}$. 
  \end{description}
\end{proof}
\end{prop}

\begin{prop}
  \label{prop:defplusplus}
  If $\Gamma_1 = \Gamma_a \oplus \Gamma_b$ and $\Gamma = \Gamma_1 \oplus \Gamma_2$ are defined, then 
  $\Gamma_a \oplus \Gamma_2$ is defined as well.
\end{prop}
%\begin{proof}
%  %TODO
%\end{proof}

\subsection{Typing}
\label{sec:typing}

We assume given a function $TypeOf : c \mapsto \sigma$ with the property that
$TypeOf(c) = \sigma$ implies $fv(\sigma) = \emptyset$. We have at least $TypeOf(\void) =
\unit$ and the following typings for functions manipulating references:

\begin{eqnarray*}
  TypeOf(\reft) & = & \forall\alpha\rho\epsilon.\regty{\rho}{\alpha\effarr{\epsilon\cup \rho} \refreg{\alpha}{\rho}} \\
  TypeOf( := ) & = & \forall\alpha\rho\epsilon\epsilon'. \refreg{\alpha}{\rho} \effarr{\epsilon} \alpha
  \effarr{\epsilon'\cup \rho} \unit \\
  TypeOf( !) & = & \forall\alpha\rho\epsilon. \refreg{\alpha}{\rho} \effarr{\epsilon\cup \rho} \alpha \\
\end{eqnarray*}

The typing rules are summarised in figure \ref{fig:bastyprules}.

\begin{figure}[htpb]
  \begin{center}
\begin{mathpar}
  {\inferrule* [left=\varrule] 
  {\Gamma(x) = \forall\bar{\chi}.\tau \quad \ctwf{\Gamma}{\bar{\xi}}\quad \capreg{\_}\notin \Gamma}
  {\typerel{\Gamma}{\tyapp{x}{\bar{\xi}}}{\tsub{\tau}{\bar{\chi}\mapsto\bar{\xi}}}{\emptyset}} }\and
  {\inferrule* [left=\conrule] 
    {TypeOf(c) = \tau \\ \cwf{\Gamma}\\ \capreg{\_}\notin \Gamma}
    {\typerel{\Gamma}{c}{\tau}{\emptyset}}}\and
  {\inferrule* [left=\absrule] 
    {\typerel{\Gamma,x:\tau'}{e}{\tau}{\varepsilon}\\ \capreg{\_}\notin\Gamma} 
    {\typerel{\Gamma}{\lambda x:\tau'.e}{\tau'\effarr{\varepsilon} \tau}{\emptyset}}}\and
  {\inferrule* [left=\apprule] 
    {\typerel{\Gamma_1}{e_1}{\tau'\effarr{\varepsilon} \tau}{\varepsilon_1}\\ \typerel{\Gamma_2}{e_2}{\tau'}{\varepsilon_2}} 
    {\typerel{\Gamma_1\oplus \Gamma_2}{e_1\ e_2}{\tau}{\varepsilon\cup \varepsilon_1 \cup \varepsilon_2}}}\and
  {\inferrule* [left=\letregrule] 
    {\typerel{\Gamma,\rho,\capreg{\rho}}{e}{\tau}{\varepsilon}} 
    {\typerel{\Gamma}{\letreg{\rho}{e}}{\tau}{\varepsilon \setminus \rho}}}\and
  {\inferrule* [left=\absregrule] 
    {\typerel{\Gamma,\capreg{\rho}}{e}{\tau}{\varepsilon}\\ \capreg{\_}\notin\Gamma} 
    {\typerel{\Gamma}{\regpr{\rho}{e}}{\regty{\rho,\varepsilon}{\tau}}{\emptyset}}}\and
  {\inferrule* [left=\appregrule] 
    {\typerel{\Gamma,\rho}{e}{\regty{\rho,\varepsilon'}{\tau}}{\varepsilon}}
    {\typerel{\Gamma,\rho,\capreg{\rho}}{\rapp{e}{\rho}}{\tau}{\varepsilon\cup \varepsilon'}}} \and
  {\inferrule* [left=\letrule]
    {\typerel{\Gamma_1,\bar{\alpha}\bar{\rho}\bar{\epsilon}}{v}{\tau'}{\emptyset} \\
     \typerel{\Gamma_2,x : \forall\bar{\alpha}\bar{\rho}\bar{\epsilon}.\tau'}{e}{\tau}{\varepsilon}}
    {\typerel{\Gamma_1 \oplus \Gamma_2}{\mllet{\bar{\alpha}\bar{\rho}\bar{\epsilon}\ x}{v}{e}}{\tau}{\varepsilon}}
  }
\end{mathpar}
  \end{center}
  \caption{The Core Typing Rules}
  \label{fig:bastyprules}
\end{figure}

\begin{figure}[htpb]
  \begin{center}
    \begin{mathpar}
      {\inferrule* [left=\regionrule] 
        {\typerel{\Gamma}{e}{\tau}{\varepsilon}} 
        {\typerel{\Gamma}{\region{\rho}{e}}{\tau}{\varepsilon \setminus \rho}}}\and
      {\inferrule* [left=\locrule] 
        {\Gamma(y) = \refreg{\tau}{\rho}\and \cwf{\Gamma} \and \capreg{\_}\notin\Gamma}
        { \typerel{\Gamma}{y_\rho}{\refreg{\tau}{\rho}}{\emptyset}}} \and
      {\inferrule* [left=\subeffrule] 
        {\typerel{\Gamma}{e}{\tau}{\varepsilon}\\ \varepsilon \subeff \varepsilon' \and \ctwf{\Gamma}{\varepsilon'}} 
        {\typerel{\Gamma}{(e : \varepsilon')}{\tau}{\varepsilon'}}}
    \end{mathpar}
  \end{center}
  \caption{Bookkeeping typing rules}
  \label{fig:advtyprules}
\end{figure}

\begin{prop}
  \label{prop:capvalues}
  $\typerel{\Gamma}{v}{\tau}{\varepsilon}$ for any value $v$ implies {\normalfont
  $\capreg{\_}\notin\Gamma$} and $\typerel{\Gamma}{v}{\tau}{\emptyset}$.
\end{prop}
\begin{proof}
  By looking at the premises for the rules for values: \varrule, \conrule,
  \absrule{} and \absregrule.
\end{proof}

\begin{prop}
  \typerel{\Gamma}{e}{\tau}{\varepsilon} implies \cwf{\Gamma} and \ctwf{\Gamma}{\tau,\varepsilon}.
\end{prop}
\begin{proof}
  We proceed by induction on the typing derivation. 
  \begin{description}

    \item[Cases \varrule, \conrule{} and \locrule] $\cwf{\Gamma}$ is a premise of
      these rules. This implies the other claims as well.

    \item[Cases \absrule{} and \absregrule] By applying the induction
      hypothesis to the pre\-mise of any of the rules, we have \cwf{\Gamma, \nu},
      where $\nu$ is the object (program variable or region capability)
      introduced by the corresponding rule. By proposition
      \ref{prop:cwfprefix}, we can conclude.

    \item[Case \apprule] By applying proposition \ref{prop:cwfoplus} and the
      induction hypothesis.

    \item[Case \subeffrule] By induction hypothesis.

    \item[Cases \letregrule{} and \regionrule] By proposition
      \ref{prop:cwfprefix} and the induction hypothesis.

    \item[Case \appregrule] by applying the rule (3). The needed hypotheses
      are obtained by induction hypothesis (\cwf{\Gamma,p} implies \cwf{\Gamma} and
      \ctwf{\Gamma}{p}).

    \item[Case \letrule] By combining the rules (2),(8) and proposition
      \ref{prop:cwfoplus}.

  \end{description}
\end{proof}

\begin{lem}[Variable substitution lemma]
  \newcommand{\thesub}[1]{\tsub{#1}{\chi' \mapsto \xi}} 
  \label{lem:typesubst}
  \typerel{\Gamma,\chi',\Delta}{e}{\tau}{\varepsilon} and \ctwf{\Gamma}{\xi} imply 
  \typerel{\Gamma,\thesub{\Delta}}{\thesub{e}}{\thesub{\tau}}{\thesub{\varepsilon}}.
\end{lem}
\begin{proof}
  Let $\theta$ be the substitution $\left[ \chi'\mapsto \xi \right]$.
  \newcommand{\thesub}[1]{\ensuremath{#1 \theta}}
  We proceed by induction over the typing derivation of
  \typerel{\Gamma,\chi',\Delta}{e}{\tau}{\varepsilon}.  We use the facts $\chi'\notin fv(\Gamma)$ (by
  proposition \ref{prop:cwftysub}) and $\chi'\notin\xi$ (by proposition
  \ref{prop:cwffvty}).
  \begin{description}

    \item[Case \varrule] The conclusion of the rule is
      \typerel{\Gamma,\chi',\Delta}{x}{\tau}{\emptyset}. Its premises are $(\Gamma,\chi',\Delta)(x) = \tau$
      (1) and \cwf{\Gamma,\chi',\Delta}. By proposition \ref{prop:cwftysub}, we obtain
      $\cwf{\Gamma,\thesub{\Delta}}$. If $x\in \Gamma$, then $(\Gamma,\chi',\Delta)(x) = \Gamma(x) =
      (\Gamma,\thesub{\Delta})(x) = \tau = \thesub{\tau}$, as $\chi'\notin \Gamma$ and by proposition
      \ref{prop:cwffvty}. If $x\in \Delta$, then $\Delta(x) = \tau$ and $(\thesub{\Delta})(x) =
      \thesub{\tau}$ and we can conclude.

    \item[Case \conrule] obvious, as $\tau$ does not contain free variables.

    \item[Case \absrule] The conclusion of the rule is \typerel{\Gamma,\chi',\Delta}{\lambda
      x:\tau'.e}{\tau'\effarr{\varepsilon} \tau}{\emptyset}. Its premises are
      \typerel{\Gamma,\chi,\Delta,x:\tau'}{e}{\tau}{\varepsilon} (1) and $\capreg{\_} \notin \Gamma,\chi',\Delta$.  We
      can immediately conclude $\capreg{\_} \notin \Gamma,\thesub{\Delta}$, and by
      induction hypothesis, we have
      \typerel{\Gamma,\thesub{\Delta},x:\thesub{\tau'}}{\thesub{e}}{\thesub{\tau}}{\varepsilon}.  We
      get, by application of \absrule{}, \typerel{\Gamma,\thesub{\Delta}}{\thesub{(\lambda
      x:\tau'.e)}}{\thesub{(\tau'\effarr{\varepsilon} \tau)}}{\emptyset}, which is the statement
      we wanted.

    \item[Case \recrule] similar.

    \item[Cases \absregrule] Similar.

    \item[Case \subeffrule] by induction hypothesis.

    \item[Cases \appregrule, \letregrule, \regionrule{} and \locrule] By
      reconstructing the subs\-ti\-tu\-ted de\-ri\-vation tree, as in the case
      for \absrule{}.

    \item[Case \letrule] By induction hypothesis. One can assume $\chi'$ and
      $fv(\xi)$ different from the $\bar{\alpha},\bar{\rho},\bar{\epsilon}$ of the rule.

  \end{description} 
\end{proof}

\begin{prop}[Irrelevant hypotheses]
  \label{prop:irrelhyp}
  If \typerel{\Gamma,x : \tau',\Delta}{e}{\tau}{\varepsilon} and $x\notin fv(e)$, then
  \typerel{\Gamma,\Delta}{e}{\tau}{\varepsilon}.
\end{prop}
\begin{proof}
  By induction on the typing derivation. This proof poses no difficulties.
\end{proof}

\begin{prop}[Weak Weakening]
  \label{prop:weakweak}
  If $\Gamma_1 \oplus \Gamma_2$ is defined, then \typerel{\Gamma_1}{e}{\tau}{\varepsilon} implies
  \typerel{\Gamma_1 \oplus \Gamma_2}{e}{\tau}{\varepsilon}.
\end{prop}

\begin{prop}[Strong Weakening]
  \label{prop:strongweak}
  If $\typerel{\Gamma}{e}{\tau}{\varepsilon}$ and $\ctwf{\Gamma}{\Delta}$ and $\capreg{\_}\notin \Delta$, we
  have $\capreg{\rho}\notin \Delta$, then \typerel{\Gamma,\Delta}{e}{\tau}{\varepsilon}.
\end{prop}

\begin{prop}[General Exchange]
  \label{prop:genexchange}
  The following statements are true:
  \begin{itemize}
    \item $\typerel{\Gamma, x : \tau, y : \tau', \Delta}{e}{\tau''}{\varepsilon}\Leftrightarrow \typerel{\Gamma, y :
      \tau', x : \tau, \Delta}{e}{\tau''}{\varepsilon}$.
   \item  $\typerel{\Gamma, x : \tau, \chi , \Delta}{e}{\tau''}{\varepsilon}\Leftrightarrow\typerel{\Gamma,\chi, x
     : \tau, \Delta}{e}{\tau''}{\varepsilon}$, if $\chi\notin fv(\tau)$.
   \item  $\typerel{\Gamma, x : \tau, \capreg{\rho} ,
     \Delta}{e}{\tau''}{\varepsilon}\Leftrightarrow\typerel{\Gamma,\capreg{\rho}, x : \tau, \Delta}{e}{\tau''}{\varepsilon}$
 \end{itemize}
\end{prop}
\begin{proof}
  straightforward induction over the typing derivation, observing that the
  respective environments are always well-formed.
\end{proof}

\begin{lem}[Substitution Lemma]
  \label{lem:substlem}
  If $\Gamma = \Gamma_1 \oplus \Gamma_2$ is defined and
  $\capreg{\_}\notin \Gamma_2$, then \typerel{\Gamma_1,x:\tau'}{e}{\tau}{\varepsilon} and
  \typerel{\Gamma_2}{v}{\tau'}{\varepsilon'} imply \typerel{\Gamma_1}{\esub{e}{x\mapsto
  v}}{\tau}{\varepsilon}.
\end{lem}
\begin{proof}
  Let us call $\sigma$ the substitution $\left[ x \mapsto v \right]$.  We proceed
  by induction on the derivation tree for \typerel{\Gamma_1,x:\tau'}{e}{\tau}{\varepsilon}.
  \begin{description}

    \item[Case \varrule] This means that $e$ is of the form $y$. If $x = y$,
      then we have $\tau' = \Gamma_1(x) = \Gamma_1(y) = \tau$. Using $x\sigma = v$, we obtain
      \typerel{\Gamma_1}{e\sigma}{\tau}{\emptyset} by the second premise and proposition
      \ref{prop:capvalues}. If $x \neq y$, we
      know that $e\sigma = y\sigma = y$, and we can conclude by proposition
      \ref{prop:irrelhyp}.

    \item[Case \conrule] As in the case of \varrule, with $x \neq y$.

    \item[Case \absrule] The conclusion of the rule is \typerel{\Gamma_1,x:\tau'}{\lambda
      y:\tau''.e}{\tau''\effarr{\varepsilon} \tau}{\emptyset}. We can assume that $y \neq x$. The
      premises of the rule are \typerel{\Gamma_1,x:\tau',y:\tau''}{e}{\tau}{\varepsilon} and
      $\capreg{\_}\notin \Gamma_1,x:\tau'$ (a).  By proposition
      \ref{prop:genexchange}, we obtain \typerel{\Gamma_1,y:\tau'',x:\tau'}{e}{\tau}{\varepsilon}. As
      $\Gamma_1 \oplus \Gamma_2$ is defined, $\Gamma_1, y : \tau'') \oplus (\Gamma_2, y : \tau'')$ is
      defined as well and is equal to $(\Gamma_1 \oplus \Gamma_2), y : \tau''$. We also
      have $\typerel{\Gamma_2, y : \tau}{v}{\tau'}{\varepsilon'}$ by proposition
      \ref{prop:strongweak}, and by induction hypothesis we have
      \typerel{\Gamma,y:\tau''}{e\sigma}{\tau}{\varepsilon'}. Now we know that $\capreg{\_}\notin \Gamma$
      (by (a) and proposition \ref{prop:capvalues}) and we can conclude with
      the rule \absrule, and by $\lambda y:\tau''.e\sigma = (\lambda y:\tau''.e)\sigma$.

    \item[Case \apprule] The conclusion of the rule is $\typerel{\Gamma_1, x :
      \tau'}{e_1\ e_2}{\tau}{\varepsilon \cup \varepsilon_1 \cup \varepsilon_2}$. But we also have $\Gamma_1, x : \tau' =
      \Gamma_a \oplus \Gamma_b$.  It can be seen that we have $\Gamma_a = \Gamma_a', x:\tau'$ and
      $\Gamma_b = \Gamma_b', x : \tau'$. The premises of the rule are $\typerel{\Gamma_a', x :
      \tau'}{e_1}{\tau'' \effarr{\varepsilon} \tau}{\varepsilon_1}$ and $\typerel{\Gamma_b',
      x:\tau'}{e_2}{\tau''}{\varepsilon_2}$. By induction hypothesis and proposition
      \ref{prop:defplusplus}, we obtain $\typerel{\Gamma_a'\oplus \Gamma_2}{e_1
      \sigma}{\tau''\effarr{\varepsilon} \tau}{\varepsilon_1}$ and $\typerel{\Gamma_b'}{e_2 \sigma}{\tau''}{\varepsilon_2}$, and
      by the rule \apprule{} we have $\typerel{(\Gamma_a'\oplus \Gamma_2) \oplus (\Gamma_b'
      \oplus \Gamma_2')}{(e_1\ e_2) \sigma}{\tau}{\varepsilon\cup\varepsilon_1\cup\varepsilon_2}$. Observing that
      $(\Gamma_a'\oplus \Gamma_2) \oplus (\Gamma_b' \oplus \Gamma_2) = \Gamma_1 \oplus \Gamma_2 = \Gamma$, we can
      conclude.

    \item[Cases \letregrule, \regionrule{} and \locrule{}] By induction
      hypothesis and proposition \ref{prop:genexchange}.

    \item[Case \absregrule] The conclusion of the rule is \typerel{\Gamma, x :
      \tau'}{\regpr{p}{e}}{\regty{p,\varepsilon}{\tau}}{\emptyset}. Its premises are
      \typerel{\Gamma, x : \tau', \capreg{p}}{e}{\tau}{\varepsilon} and $\capreg{\_}\notin \Gamma$. By
      general exchange, we can derive \typerel{\Gamma, \capreg{p}, x : \tau'}{e}{\tau}{\varepsilon}
      and thus, by induction hypothesis, \typerel{\Gamma, \capreg{p}}{e\sigma}{\tau}{\varepsilon}. By
      \absregrule{}, we obtain
      \typerel{\Gamma}{\regpr{p}{e}}{\regty{p,\varepsilon}{\tau}}{\emptyset}, as desired.

    \item[Case \appregrule] Impossible

    \item[Case \letrule] Simply by induction hypothesis. One can assume
      that $\tau'$ does not contain any of the bound variables $\chi$.
  \end{description}
\end{proof}

\subsection{Subject Reduction}
\label{sec:subjectreduction}

\begin{definition}
  For an environment $\Gamma$, define 
  \begin{eqnarray*}
    dom_\rho(\Gamma) &=& \left\{ \rho \mid \rho\in \Gamma \right\}, \\
    \Gamma(\rho) &=& \left\{ y \mapsto \tau \mid y : \refreg{\tau}{\rho} \in \Gamma\right\}.
  \end{eqnarray*}
  Now define $\storetype{\Gamma}{s}$ iff
  \begin{itemize}
    \item $dom_\rho(\Gamma) = dom(s)$
    \item $\forall \rho\in dom(s), dom(\Gamma(\rho)) = dom(s(\rho))$
    \item $\forall \rho\in dom(s), \forall y\in dom(s(\rho)),
      \typerel{\Gamma}{s(\rho)(y)}{\Gamma(\rho)(y)}{\emptyset}$.
  \end{itemize}
\end{definition}

\begin{definition}
  \label{def:invariants}
  For an environment $\Gamma$, a store $s$, an expression $e$, a type $\tau$ and an
  effect $\varepsilon$, we define the following four properties:
  \begin{description}
    \item[(a)] $\typerel{\Gamma}{e}{\tau}{\varepsilon}$
    \item[(b)] $\storetype{\Gamma}{s}$
    \item[(c)] $\capreg{\rho} \in \Gamma$ implies $s(\rho) = \emptyset$
    \item[(d)] $singleton(s)$, that is $\forall\rho\in dom(s)$, either $s(\rho) =
      \emptyset$ or $s(\rho) = \{ y \mapsto v\}$ for some $y,v$.
  \end{description}
\end{definition}

\begin{thm}[Subject Reduction]
  \label{thm:subred}
  Suppose the properties of \ref{def:invariants} are true for $\Gamma,s,e,\tau,\varepsilon$ and
  suppose $s,erase(e) \basered s',erase(e')$ for some $s',e'$. Then there is
  an environment $\Gamma'$ such that these properties are true as well for
  $\Gamma',s',e',\tau,\varepsilon$. $\Gamma'$ relates to $\Gamma$ in one of the following ways:
  \begin{itemize}
    \item $\Gamma' = \Gamma, \rho,\capreg{\rho}$ with $\rho\notin \Gamma$
    \item $\Gamma' = (\Gamma - \capreg{\rho}), y : \refreg{\tau}{\rho}$
    \item $\Gamma' = \Gamma$ 
  \end{itemize}
\end{thm}
\begin{proof}
  By induction on the reduction $s, e \basered s', e'$. In these subcases we
  usually disregard the possibility that the last rule applied to establish
  the typing of $e$ has been the $\subeffrule$ rule. This does not alter the
  proofs.   \begin{description}
    \item[Case $\beta$] In this case, $e = (\lambda x : \tau'. e_1)\ v$ and we have the
      reduction $s,(\lambda x : \tau'. e_1)\ v \basered s, \esub{e}{x\mapsto v}$.
      Every derivation tree for $\typerel{\Gamma}{(\lambda x : \tau'. e_1)\ v}{\tau}{\varepsilon}$ has
      the following shape:
      \begin{equation*}
      \inferrule* [left=\apprule]
        {\inferrule* [left=\subeffrule] 
          {\inferrule* [left = \absrule]
            {\typerel{\Gamma_1, x : \tau'}{e_1}{\tau}{\varepsilon} \\ \capreg{\_}\notin \Gamma_1}
            {\typerel{\Gamma_1}{\lambda x : \tau'.e_1}{\tau'\effarr{\varepsilon} \tau}{\emptyset}
            } }
        {\typerel{\Gamma_1}{\lambda x : \tau'.e_1}{\tau'\effarr{\varepsilon} \tau}{\varepsilon_1}} \\
        {\typerel{\Gamma_2}{v}{\tau'}{\varepsilon_2} }}
        {\typerel{\Gamma_1}{(\lambda x : \tau'. e_1)\ v}{\tau}{\varepsilon \cup \varepsilon_1 \cup \varepsilon_2}}
      \end{equation*}

      As a remark, we are allowed to write $\Gamma_1$ instead of $\Gamma_1\oplus \Gamma_2$ in
      the last line thanks to propositions \ref{prop:capvalues} and
      \ref{prop:oplusabsorb}. We still can use the fact that $\Gamma_1 \oplus \Gamma_2$
      is defined. This will also be used in many of the other cases.

      Now the substitution lemma \ref{lem:substlem} uses the established facts
      to exactly prove (a) : \typerel{\Gamma}{\esub{e}{v \mapsto v}}{\tau}{\varepsilon}. Using
      the \subeffrule{} rule, we can increase the effect as desired to $\varepsilon \cup
      \varepsilon_1 \cup \varepsilon_2$. We now can take $s' = s$ and the properties (b), (c) and
      (d) are true by hypothesis. Also note that if the last rule applied was
      an additional \subeffrule{} rule, we could have increased the effect
      accordingly in the resulting typing derivation, by the bias of the last
      typing rule.

    \item[Case $\chi$] In this case, $e = \tyapp{(\lambda\chi.e)}{\xi}$, and we have the
      following reduction: $s, \tyapp{(\lambda\chi.e)}{\xi} \basered s,
      \tsub{e}{\chi\mapsto \xi}$.  Every derivation tree for
      $\typerel{\Gamma}{\tyapp{(\lambda\chi.e)}{\xi}}{\tau}{\varepsilon}$ has the following shape:
      \begin{equation*}
        \inferrule* [left=\apptyrule]
          {\inferrule* [left=\abstyrule]
            {\typerel{\Gamma,\chi}{v}{\tau}{\emptyset} \\ \capreg{\_}\notin \Gamma}
            {\typerel{\Gamma}{\lambda\chi.v}{\forall\chi.\tau}{\emptyset} \\ \ctwf{\Gamma}{\xi} } }
            {\typerel{\Gamma}{\tyapp{(\lambda\chi.e)}{\xi}}{\tsub{\tau}{\chi\mapsto \xi}}{\emptyset}}
      \end{equation*}
      Now the type substitution lemma \ref{lem:typesubst} gives us
      $\typerel{\tsub{e}{\chi \mapsto \xi}}{\tsub{\tau}{\chi\mapsto \xi}}{\emptyset}$, as
      desired. Again we can take $s' = s$ $\Sigma' = \Sigma$ and the properties (b), (c)
      and (d) are true by hypothesis.

\item[Case $p$] In this case, $e = \rapp{\regpr{p}{e}}{p}$, and we have the
  following reduction: $s, \rapp{\regpr{p}{e}}{p} \basered s,e$. The
  derivation tree for $\typerel{\Gamma}{\rapp{\regpr{p}{e}}{p}}{\tau}{\varepsilon}$ has the
  following shape:
  \begin{equation*}
    \inferrule* [left=\appregrule]
      { \inferrule* [left=\absregrule]
        {\typerel{\Gamma,p,\capreg{p}}{e}{\tau}{\varepsilon} \\ \capreg{\_}\notin \Gamma,\Sigma}
        {\typerel{\Gamma,p}{\regpr{p}{e}}{\regty{p,\varepsilon}{\tau}}{\emptyset} \\ \capreg{p}\notin \Gamma}}
        { \typerel{\Gamma,p,\capreg{p}}{\rapp{(\regpr{p}{e})}{p}}{\tau}{\varepsilon}}
  \end{equation*}
    The conclusion is now obvious.
    \item[Case $letreg$]
      In this case, $e = \letreg{\rho}{e'}$ and we have the following reduction:
      $s,\letreg{\rho}{e'} \basered \statesub{s}{\rho_1\mapsto \left\{ \right\}},
      \region{\rho_1}{\regsub{e'}{\rho\mapsto \rho_1}}$, where $\rho_1\notin s$. The
      derivation tree looks as follows:
    \begin{equation*}
      \inferrule* [left=\letregrule]
        {\typerel{\Gamma,\rho,\capreg{\rho}}{e'}{\tau}{\varepsilon} \and \capreg{\rho}\notin \Gamma}
        {\typerel{\Gamma}{\letreg{\rho}{e'}}{\tau}{\varepsilon\setminus \rho }}
    \end{equation*}

    Using the \regionrule{} rule, and defining $\Gamma' = \Gamma, \rho_1, \capreg{\rho_1}$, we
    obtain (a).  Property (d) is still true for $s'$, as we only added an
    empty region.  Property (c) is correct for $\Gamma'$ and $s'$, as we added one
    capacity to $\Gamma$, whose region is empty. Finally, $(b)$ is correct, because
    we added the same to the store as to the environment.

  \item[Case $region$] In this case, $e = \region{\rho_1}{v}$ and we have the
    following reduction: $s, \region{\rho_1}{v} \basered s,v$. The derivation
    tree for $\typerel{\Gamma}{\region{\rho_1,\rho}}{\tau}{\varepsilon}$ looks as follows:
    \begin{equation*}
      { \inferrule* [right=\regionrule]
        { \typerel{\Gamma}{v}{\tau}{\varepsilon} }
        { \typerel{\Gamma}{\region{\rho_1}{v}}{\tau}{\varepsilon\setminus \rho_1} }
        }
    \end{equation*}
    Taking $\Gamma' = \Gamma$, (a) is trivial as the type $\tau$ is preserved and $v$ is a
    value. As $s' = s$, the other conditions are true by hypothesis.
  \item[Case $ref$] In this case, $e = \reft \left[ \xi_1 \dots \xi_n \right]
    \capreg{\rho_1}\ v$, and we have the following reduction starting from $s,e$:
    $s, (\reft \left[ \xi_1 \dots \xi_n \right] \ \capreg{\rho_1})\ v \basered
    s\left[ \rho_1 \mapsto s(\rho_1)\left[ y\mapsto v \right]\right], y_{\rho_1}$ where
    $\rho_1\in s, y\notin s(\rho_1)$. We have the following derivation tree:
    \begin{equation*}
    { \inferrule* [right=\apprule]
      { \inferrule* [right=\appregrule]
        {\typerel{\Gamma_1}
                   {\reft \left[ \xi_1 \dots \xi_n \right] }
                   {\regty{\rho,\varepsilon'}{(\tau\effarr{\varepsilon}\refreg{\tau}{\rho})}}
                   {\varepsilon_1} }
        {\typerel{\Gamma_1, \capreg{\rho_1}}
                 {(\reft \left[ \xi_1 \dots \xi_n \right] )@\rho_1}
                 {\tau\effarr{\varepsilon}\refreg{\tau}{\rho}}{\varepsilon'\cup \varepsilon_1}} \\
                 \typerel{\Gamma_2}{v}{\tau}{\varepsilon_2} 
      }
      {\typerel{\Gamma_1, \capreg{\rho_1}}{\reft \left[ \xi_1 \dots \xi_n \right] \rho_1\ v}{\refreg{\tau}{\rho}}{\varepsilon\cup \varepsilon_1 \cup \varepsilon_2 \cup \varepsilon'}}
    }
    \end{equation*}
      with $\rho_1\in \Gamma_1$. Now, define $\Gamma' = (\Gamma - \capreg{\rho_1}), y :
      \refreg{\tau}{\rho_1}$. By construction, we have (a) (by application of the
      \locrule{} rule). We also easily have (c): The store typing only changed
      for $\rho_1$, but we removed the capacity $\capreg{\rho_1}$ from $\Gamma$. We also
      have (d): As $\capreg{\rho_1}\in \Gamma$, we have by hypothesis that $s(\rho_1) =
      \emptyset$. Now, $s'$ differs from $s$ only in a reference allocated in
      $\rho_1$, so $singleton(s')$ is still true. For (b), we know that $\rho_1\in
      dom(s)$, so $dom(s) = dom(s')$, and thus $dom_\rho(\Gamma') = dom(s')$.  We now
      have $y\in dom(\Gamma'(\rho_1))$, but we also have $y\in dom(s'(\rho_1))$. As
      $\typerel{\Gamma_2}{v}{\tau}{\varepsilon_2}$, by proposition \ref{prop:weakweak}, we
      have $\typerel{\Gamma'}{v}{\tau}{\varepsilon_2}$. By construction, we have $\Gamma'(\rho_1)(y) =
      \tau$.  This proves $\typerel{\Gamma'}{s'(\rho_1)(y)}{\Gamma'(\rho_1)(y)}{\emptyset}$.

    \item[Case $:=$] In this case, $e = (:= \left[ \xi_1 \dots \xi_n \right]
      y_{\rho_1}\  v)$ and we have the following reduction $s, := \left[ \xi_1
      \dots \xi_n \right] \ y_{\rho_1}\ v \basered s\left[ \rho_1 \mapsto
      s(\rho_1)\left[ y \mapsto v \right] \right], \void$, with $\rho_1\in s, y\in
      s(\rho_1)$. The derivation tree for $\typerel{\Gamma}{e}{\tau}{\varepsilon}$ looks as
      follows:
      \begin{equation*}
        { \inferrule* [right=\apprule]
          { 
            \inferrule* [right=\apprule]
              {\typerel{\Gamma_1}
                       { := \left[ \xi_1 \dots \xi_n \right] }
                       { \refreg{\tau}{\rho} \effarr{\varepsilon} \tau \effarr{\varepsilon'} \unit }
                       {\varepsilon_1}
                \\
              \typerel{\Gamma_1}{y_{\rho_1}}{\refreg{\tau}{\rho}}{\varepsilon_2}\  (1) }
              {\typerel{\Gamma_1}
                    {:= \left[ \xi_1 \dots \xi_n \right] y_{\rho_1}}
                    { \tau \tyarr \unit}
                    {\varepsilon\cup \varepsilon_1 \cup \varepsilon_2}
               } \\ 
             \typerel{\Gamma_2}{v}{\tau}{\varepsilon_3} \   (2)
          }
          { \typerel{\Gamma}
                    {:= \left[ \xi_1 \dots \xi_n \right] y_{\rho_1}\ v)}{\unit}
                    {\varepsilon \cup \varepsilon_1 \cup \varepsilon_2 \cup \varepsilon_3 \cup \varepsilon'}
          }
        }
      \end{equation*}
      Set $\Gamma' = \Gamma$. As $e'$ is \void, (a) is trivial to prove by the
      \conrule{} and \subeffrule{} rules. For (b), it suffices to show that
      $\typerel{\Gamma}{s'(\rho_1)(y)}{\Gamma(\rho_1)(y)}{\emptyset}$. Using (1) and (2), this
      is obvious. (c) and (d) are true by hypothesis.

    \item[Case $!$]
      In this case, $e = ! \left[\xi_1 \dots \xi_n\right] y_{\rho_1}$, and we have
      the reduction $s, ! \left[\xi_1 \dots\xi_n\right] y_{\rho_1} \basered s,
      s(\rho_1)(y)$, where $\rho_1\in s, y\in s(\rho_1)$. The derivation tree for
      $\typerel{\Gamma}{e}{\tau}{\varepsilon}$ looks as follows:
      \begin{equation*}
        { \inferrule* [right=\apprule]
          { \typerel{\Gamma}{!\left[\xi_1 \dots\xi_n\right]}
            {\refreg{\tau}{\rho} \effarr{\varepsilon} \tau}{\varepsilon_1} \\
            \typerel{\Gamma} { y_{\rho_1}} {\refreg{\tau}{\rho} }{\varepsilon_2}\ (1)}
          { \typerel{\Gamma} {! \left[\xi_1 \dots\xi_n\right] y_{\rho_1}}{\tau}{\varepsilon\cup
          \varepsilon_1\cup \varepsilon_2} }
        }
      \end{equation*}
      Again, we set $\Gamma' = \Gamma$. As $s' = s$, we have the properties (b), (c),
      and (d) for free by hypothesis. Using (1), we easily prove (a).
  \end{description}
\end{proof}

\begin{thm}[One-Step Subject Reduction]
  If the properties of definition \ref{def:invariants} are true for $\Gamma,e,\tau,s$
  and $\varepsilon$, then for any $s',e'$ such that $s,e\onestep s',e'$ there is a $\Gamma'$  
  such that these properties are true for $\Gamma',e',\tau,s'$ and $\varepsilon$. $\Gamma'$ relates
  to $\Gamma$ as in theorem \ref{thm:subred}
  \label{thm:onestep-subred}
\end{thm}
\begin{proof}
  By induction on the form of the reduction context.
  \begin{description}
    \item[Case \bracket{~} ] This corresponds to the theorem
      \ref{thm:subred}.

    \item[Case $E\ \bracket{\xi}$ ] There is nothing to prove: The rule
      \appgenrule{} applies as well after the reduction.

    \item[Case $E\ e$] We know that $e = e_1\ e_2$ and $s,e_1 \onestep
      s',e_1'$. The typing derivation looks as follows :
      \begin{equation*}
        { \inferrule* [left=\apprule]
          { \typerel{\Gamma_1}{e_1}{\tau' \effarr{\varepsilon} \tau}{\varepsilon_1} \and
          \typerel{\Gamma_2}{e_2}{\tau'}{\varepsilon_2}}
          { \typerel{\Gamma_1 \oplus \Gamma_2}{e_1 \oplus e_2}{\tau}{\varepsilon\cup\varepsilon_1\cup\varepsilon_2 } }
        }
      \end{equation*}
      We can apply the induction hypothesis to obtain $\typerel{\Gamma_1'}{e_1'}{\tau,
      \tyarr \tau}{\varepsilon_1}$.  To reconstruct a typing for $e_1'\ e_2$, we need to
      show that we can extend $\Gamma_2$ to $\Gamma_2'$ such that $\Gamma_1'\oplus \Gamma_2'$ is
      defined and $e_2$ is still typable. Let $\Gamma_2'$ be obtained in the same
      way from $\Gamma_2$ as $\Gamma_1'$ from $\Gamma$, but without adding (or removing) any
      capabilities. Using proposition \ref{prop:strongweak}, we see that $e_2$
      is still typable using $\Gamma_2'$, and it is easy to see that $\Gamma_1' \oplus
      \Gamma_2'$ is defined.

    \item[Case $v\ E$] Similar to the previous case.

    \item[Case $\rapp{E}{p}$]  Similar, the typing rule \appregrule{} does not
      pose any problems.
  \end{description}
\end{proof}

\begin{thm}[General Subject Reduction]
  If $\Gamma,s,e, \tau,\varepsilon$ fulfill the properties of definition \ref{def:invariants},
  and if $s,e \reduction s',e'$, there is $\Gamma'$  such that $\Gamma',s,e,\tau,\varepsilon$ do
  fulfill these properties as well.
  \label{thm:generalsubred}
\end{thm}
\begin{proof}
  Simple induction over the transitive closure of $\onestep$ (over the length
  of the reduction sequence). In the reflexive
  case (0 steps), we have $s = s', e = e'$ and the claim is trivial. In the
  case of a single $\onestep$ step, theorem \ref{thm:onestep-subred} says
  exactly what we want. In the transitive case, we just apply the induction
  hypothesis twice.
\end{proof}

In particular, the properties of definition \ref{def:invariants} are true for
$\Gamma = \emptyset$, and $s = \emptyset$, and any well-typed closed
($\Gamma = \emptyset$) program $e$ of type $\tau$ and effect $\varepsilon$.

\subsection{Other properties}

\begin{lem}[Values]
  The only Values of type $\refreg{\tau}{p}$ are variables and locations.
  \label{lem:valueprops}
\end{lem}
\begin{proof}
  Simply by looking at the typing rules.
\end{proof}

\section{The Weakest Precondition Calculus}

In this section we prove the correctness of the weakest precondition calculus,
including the instantiation restriction. Instead of the dealing directly with
region calculus, we take for granted that a region corresponds to a reference
and use globally created references instead. While the weakest precondition
calculus is insensible to aliasing questions, it is not presented as such
here; we give a direct interpretation of the the store types $ε$ (written
$\langle ε \rangle$ in the paper), by defining a logic called {\em effect
logic}, which in turn can be translated to standard higher-order logic. In
effect logic, all operations on stores (the combine operator $\oplus$ and the
restriction operator $|_ε$) are carried out directly and this needs the
aliasing restriction. Unfortunately, this makes the presented proof
unnecessarily dependent on the aliasing restriction.  Our proof thus only
applies in the context of alias-free programs. However, it would be easy to
adapt it to the more general context. There are some other superficial
differences with language in the paper. Here we separate more clearly between
the logic and programs.  Other small differences are that here the {\bf let}
and {\bf if}-typing rules do effect union, while in the paper the subeffecting
rule takes care of that.

\subsection{The Language}

\subsubsection*{Syntax}

  \begin{eqnarray*}
    c &::=& n \sep () \sep \true \sep \false\\
    v &::=& c \sep x~\alist{κ} \sep \lambda x\colonτ.e \sep
    R \sep \rec~(y\colonτ')~(x\colonτ).~e\\
    e &::=& v \sep v~v\sep \letst~x~\alist{χ}=e~\inst~e \sep v := v\sep
    !\,v\\
    & & \ifst~v~\thenst~e~\elsest~e\\
    r &::=& R \sep ρ \\
    b &::=& r \sep \ev\\
    ε  &::=& b,\cdots, b\\
    χ &::=& ρ \sep \ev \\
    κ &::=& τ \sep r \\
    \iota &::& \inttype \sep \unit \sep \bool \\
    τ  &::=& \iota \sep τ  ->^ε  τ \sep \reftype_r \\
    S  &::=& \forall \alist{χ}.τ \\
    E &::=& [~]\sep \letst~x~\alist{χ}=E~\inst~e\\
    Γ &::=& \emptyset \sep Γ, χ \sep Γ, x : S
  \end{eqnarray*}

\subsubsection*{Semantics}
%TODO lookup and assignment to type
  \framebox{$ s,e--\ s',e'$}
  \begin{mathpar}
    {\inferrule* [left=Beta] {~} {s, (\lambda x\colonτ.e)~v --\ s, e[x|->v]} }
    \and
    {\inferrule* [left=Rec] {~} {s,
        (\rec~(y\colonτ ')~(x\colonτ ).~e)~v --\ s,
        e[x|->v,y|->\rec~(y\colonτ ')~(x\colonτ ).~e]} }
    \and
    {\inferrule* [left=Let] {~} {s, \letst~x~\alist{χ}=v~\inst~e --\ s,
        e[\Lambda \alist{χ}.x|->v]} }
    \and {\inferrule* [left=Lookup] {~} {s, !\,r --\ s, s(r)} }
    \and {\inferrule* [left=Assign] {~} {s, r:=n --\ s[r:=n], \void} }
    \and {\inferrule* [left=IfTrue] {~}
      {s,\ifst~\true~\thenst~e_1~\elsest~e_2 --\ s,e_1} }
    \and {\inferrule* [left=IfFalse] {~}
      {s,\ifst~\false~\thenst~e_1~\elsest~e_2 --\ s,e_2} }
  \end{mathpar}
  \framebox{$ s,e--> s',e'$}
  \begin{mathpar}
    {\inferrule* [left=Context] {s,e --\ s',e'} 
      {s, E[e] --> s',E[e']} }
  \end{mathpar}
  \framebox{$ s,e->> s',e'$}
  \begin{mathpar}
    {\inferrule* [left=Refl] {~} {s,e ->> s,e} }
    \and {\inferrule* [left=Step] {s,e->>s',e' \\ s',e' ->> s'',e''} 
      {s,e ->> s'',e''} }
  \end{mathpar}

\subsubsection*{Typing}
%TODO define compatibility for ref substitution 
\framebox{$Γ |-_vv:τ$}\\
  \begin{mathpar}
      {\inferrule* [left=Var] {Γ(x) = \forall\alist{χ}.τ \\
          \sigma = [\alist{χ}|->\alist{κ}]\\ \sigma\simτ }
        {Γ|-_v x~\alist{κ} :τ\sigma} } \and
    { \inferrule* [left=Const] {~} {Γ |-_v c : \mathit{TypeOf}(c) } } \and
    {\inferrule* [left=Ref] {~} {Γ|-R : \reftype_R}} \and
    { 
\inferrule* [left=Abs] 
      {Γ ,x\colon τ '|- e : τ ;ε  \\ \ceil{Γ
          ,x:τ '}|- p : ε -> \proptype;ε \\
      \ceil{Γ ,x\colon τ '},|-q: ε -> ε
      ->\ceil{τ }->\proptype;ε} 
      {Γ |-_v \lambda x\colon τ '.\{p\}e\{q\}:τ '->^ε τ } }
    \and
    { \inferrule* [left=Rec]
%{Γ|-_v \rec~(y:\colonτ'->^ε} }
      {Γ ,y\colon τ '->^ετ ,~x\colon τ '|- e : τ ;ε  \\
        \ceil{Γ ,x\colon τ '}|- p : ε
        ->\proptype;ε \\
      \ceil{Γ ,x\colon τ '}|-q:ε ->ε ->\ceil{τ }->\proptype;ε} 
      {Γ |-_v \rec~(y\colon τ '->^{ε}τ )~(x\colon τ) 
        ').\{p\}e\{q\}:τ '->^ε τ
      } }
    \end{mathpar}
    \framebox{$Γ |-e:τ;ε$}\\
    \begin{mathpar}
    {\inferrule* [left=Val] {Γ |-_v v : τ } {Γ |-v : τ ;\emptyset} }
    \and
    {\inferrule* [left=Lookup] {Γ|-_v v : \reftype_ρ } {Γ |-!\,v : \inttype;ρ} }
    \and
    {\inferrule* [left=Assign] 
      {Γ|-_v v_1 : \reftype_ρ \\Γ |-_v v_2 : \inttype} 
      {Γ |-v_1:=v_2 : \unit;ρ} 
    }
    \and
    {\inferrule* [left=App] {Γ |-_v v_1 : τ '->^ε τ  \\ Γ |-_v v_2 : τ '} 
      {Γ |-v_1~v_2 : τ ;ε } }
    \and
    {\inferrule* [left=Let] {Γ |- e_1 : τ ';ε _1 \\ Γ ,x:τ '|- e_2 : τ ;ε _2} 
      {Γ |-\letst~x = e_1~\inst~e_2 : τ ;ε _1ε _2} }
    \and
    {\inferrule* [left=Letv] 
      {Γ ,\alist{χ}|-_v e : τ ' \\ Γ ,x:\forall \alist{χ}.τ '|- e : τ ;ε } 
      {Γ |-\letst~x~\alist{χ} = v~\inst~e : τ ;ε } }
    \and
    {\inferrule* [left=If]
      {Γ |- v : \bool \\ Γ |- e_1 : τ \\ Γ |- e_2 :
        τ}
      {Γ |- \ifst~v~\thenst~e_1~\elsest~e_2 : τ}
    }
  \end{mathpar}

\subsection{SHOL}

\subsubsection*{Syntax}
  \begin{eqnarray*}
    f &::=& c \sep x \sep \lambda x\colon\theta  .f \sep
     \\
    & & f~f\sep f => f \sep f = f \sep (f,f) \sep\\
    & & \forall x\colon\theta  .f \sep \forall \alist{χ}.f \sep \pi_1 \sep
    \pi_2 \\
    \theta   &::=& \ev\sep \inttype \sep \unit \sep \proptype \sep \\
    & & \theta   -> \theta   \sep \theta *\theta \\
    \Delta  &::=& \emptyset  \sep \Delta , x \colon \theta   \sep \Delta ,\alist{χ}
  \end{eqnarray*}
\subsubsection*{Typing rules}
    \framebox{$\Delta|-f:\theta$}\\
  \begin{mathpar}
   { \inferrule* [left=Var] {\Delta (x) = \theta  } {\Delta |- x : \theta  } } 
   \and
   { \inferrule* [left=Const] { } {\Delta |- c : \mathit{TypeOf}(c)  } } 
   \and
   { \inferrule* [left=Abs] {\Delta ,x\colon\theta  |-f:\theta  '} {\Delta |- \lambda x\colon\theta  .f : \theta  ->\theta  '} }
   \and
   { \inferrule* [left=App] {\Delta |-f:\theta  '->\theta   \\ \Delta |-f':\theta  '} {\Delta |- f~f' : \theta  } }
   \and
   { \inferrule* [left=Eq] {\Delta |-f,f':\theta   } {\Delta |- f=f' : \proptype} }
   \and
   { \inferrule* [left=Impl] {\Delta |-f,f':\proptype } {\Delta |- f=> f' : \proptype} }
   \and
   { \inferrule* [left=Ty] {\Delta ,\alist{χ}|-f:\proptype } {\Delta |-
       \forall \alist{χ}.f : \proptype} }
   \and
   { \inferrule* [left=FA] {\Delta ,x\colon\theta  |-f:\proptype } {\Delta |- \forall x\colon\theta  .f : \proptype} }
   \and
   { \inferrule* [left=Pair] {\Delta |-f_i:\theta  _i\quad i =1,2} {\Delta |- (f_1,f_2): \theta  _1*\theta  _2} }
   \and
   { \inferrule* [left=Proj,right={~\textnormal i=1,2}] {\Delta |-f:\theta  _1*\theta  _2} {\Delta |- \pi _i~f: \theta  _i}}
  \end{mathpar}

\subsection{Effect logic}

\begin{figure*}[tpb]
  \begin{minipage}[t]{0.5\linewidth}
  \end{minipage}
  \begin{minipage}[t]{0.5\linewidth}
  \end{minipage}
  \caption{Syntax and typing of the new constructs}
  \label{fig:pholeff}
\end{figure*}
\subsubsection*{Syntax}
  \begin{eqnarray*}
    l & & \text{Labels} \\
    f &::=& ... \sep !\,r^l \sep x~\alist{ε }
    \sep 
    \lambda l\colonε.f \sep \\
    & & \forall l\colonε.f \sep f~s\\
    s &::=& \{r_1 = f_1 \cdots r_n = f_n, \ev_1 = l_1, \cdots, \ev_m
      = l_m \}\\
    \theta &::=& \iota \sep \proptype \sep \theta -> \theta \sep
    \theta*\theta \sep ε ->\theta \\
    \zeta &::=& \forall \alist{χ}.\theta \\
    \Delta  &::=& \emptyset \sep \Delta ,x\colon\zeta \sep \Delta ,\alist{χ}
    \sep \Delta , l\colonε 
  \end{eqnarray*}
\subsubsection*{Typing (extension of the previous rules)}
  \begin{mathpar}
    {\inferrule* [left=Var] {\Delta (x) = \forall \alist{χ}.\theta } {\Delta |-x~\alist{ε } :
        \theta [\alist{χ}|->\alist{κ }]}} 
    \and
    {\inferrule* [left=Rvar] {r\in\Delta (l)} {\Delta |-!\,r^l : \inttype} }
    \and
    {\inferrule* [left=EffAbs] {\Delta ,l\colonε |-f:\theta }
      {\Delta |-\lambda l \colon  ε .f : ε  -> \theta } }
    \and
    {\inferrule* [left=EffFA] {\Delta ,l\colonε |-f:\proptype}
      {\Delta |-\forall l\colon ε .f : \proptype} }
    \and
    {\inferrule* [left=EffRec] 
      {\Delta|-f:ε ->\theta  \\ \Delta |-f_i :
        \inttype \\ \ev_j\in \Delta (l_j) \\ ε =
        \alist{r_i}\alist{\ev_j} } 
      {\Delta |-f~\{r_i = f_i,\ev_j = l_j\}: \theta } }
  \end{mathpar}

In the following, we denote $σ$ an atomic substitution either of the
form $[\ev|->ε] $ or of the form $[ρ|->r]$.
\begin{definition}
  Let $σ$ be a substitution. Then the application of this substitution
  is defined differently for effect substitutions and reference
  substitutions. If $σ = [\ev|->ε_0]$, then
  \begin{equation*}
    s\sigma= \left(\bigoplus_{b\in
                  ε_s,b\neq\ev}s(b)\sigma~\oplus
                ε_0^{s(\ev)}\right).
  \end{equation*}
  In the other case of $σ=[ρ|->r]$, 
  \begin{equation*}
    %TODO introduce notation for that
    s\sigma= \bigoplus_{b\in ε} bσ =  s(b)σ.
  \end{equation*}

\end{definition}
\begin{definition}
  The application of a substitution $\sigma$ to a formula is defined in the
  following way (we only show the cases for the additional constructs,
  otherwise the definition is straightforward, descending recursively
  into subterms):
  \begin{eqnarray*}
    (\lambda l\colonε.f)\sigma &=& \lambda l\colon(ε\sigma).(f\sigma)\\
    (\forall l\colonε.f)\sigma &=& \forall l\colon(ε\sigma).(f\sigma)\\
    (f~s)σ &=& fσ~sσ\\
    (!\,r)σ &=& !\,(rσ)
  \end{eqnarray*}
\end{definition}

\begin{definition}
  %TODO polish
  We define the application of a label substitution $[l|->s_l]$.
  \begin{description}
  \item[for records:] $s[l|->s_l] = \bigoplus_{b\inε}s(b)$
  \end{description}
\end{definition}

\begin{prop}
  For two effect records $s_1,s_2$ of domain $ε_1$ and $ε_2$,
  respectively, and atomic substitution $σ$, we have
  \begin{equation*}
    (s_1\oplus s_2)σ \equiv s_1σ \oplus s_2σ.
  \end{equation*}
  \label{prop:recsubst}
\end{prop}
\begin{proof}
  The proof is trivial for the reference substitution case. In the
  case of an effect substitution, let $σ=[\ev|->ε_0]$. Then we write
  the following chain of equations.
  \begin{align*}
  s_1 σ\oplus s_2σ &= 
  \bigoplus_{b\inε_1σε_2σ}\left\{
    \begin{array}[]{ll}
      s_2σ(b) & \text{if}~ b\inε_2σ\\
      s_1σ(b) & \text{otherwise}
    \end{array}
\right.\\
  &=\bigoplus_{b\inε_1σε_2σ}\left\{
    \begin{array}[]{ll}
      (\bigoplus_{b\inε_1} s_2(b)σ \oplus ε^{s_2(\ev)})(b) & \text{if}~ b\inε_2\\
      (\bigoplus_{b\inε_2} s_1(b)σ \oplus ε^{s_1(\ev)})(b) &
      \text{otherwise}\
    \end{array}
\right.\\
  &=\left(\bigoplus_{b\inε_1σε_2σ}\left\{
    \begin{array}[]{ll}
      (\bigoplus_{b\inε_1} s_2(b)σ )(b) & \text{if}~ b\inε_2\\
      (\bigoplus_{b\inε_2} s_1(b)σ )(b) &
      \text{otherwise}\
    \end{array}
\right.\right)
\oplus ε^{(s_1\oplus s_2)(\ev)}\\
  &=\left(\bigoplus_{b\inε_1σε_2σ}\left\{
    \begin{array}[]{ll}
      s_2(b)σ & \text{if}~ b\inε_2\\
      s_1(b)σ &
      \text{otherwise}\
    \end{array}
\right.\right)
\oplus ε^{(s_1\oplus s_2)(\ev)}\\
  &=\left(\bigoplus_{b\inε_1σε_2σ}\left\{
    \begin{array}[]{ll}
      s_2(b) & \text{if}~ b\inε_2\\
      s_1(b) &
      \text{otherwise}\
    \end{array}
\right.\right)σ\\
&=(s_1\oplus s_2)σ
  \end{align*}
  This proof silently supposes that both effect records contain the
  field $\ev$. If one defines $ε^{s(\ev)}$ to be the empty record if
  $s$ does not contain $\ev$, then the same proof is true in the other
  three cases.
\end{proof}

\subsubsection*{Interpretation of the new constructs}

\begin{eqnarray*}
  [|\lambda l\colon ε.f|] & = & \lambda _{b\inε }(b^l\colon \theta _b).[|f|]\\
  ~[|\forall l\colon ε.f|] & = & \forall _{b\in ε
  }(b^l\colon \theta _b). [|f|]\\
  ~[|!\,r^l |] & = &r^l \\
  ~[|f~s~|] & = &
          [|f|]~[|f_1|]~\cdots~[|f_n|]~\ev_1^{l_1}~\cdots~\ev_m^{l_m}
\end{eqnarray*}
where
\begin{equation*}
s = \{r_1 = f_1,\cdots,r_n = f_n,\ev_1 = l_1,\cdots ,\ev_m
  = l_m \}
\end{equation*}

\subsubsection*{Equational Theory}

In addition to the equation schemes present in SHOL, we add the
following equation schema:
\begin{equation}
  (λl:ε.f)~s = fμ,\qquad μ = [ b^l|->s(b) \mid b\inε].
  \label{eq:eqschema}
\end{equation}

\subsubsection*{Effect substitutions}

\begin{definition}
  An effect substitution $σ = [\ev |-> ε]$ is {\em compatible} with an
  effect $ε'$ if $\ev\inε$ implies $ε$ disjoint from
  $ε'$. Furthermore, a reference substitution $[ρ|->r]$ is {\em
    compatible} with $ε'$ if $ρ\inε'$ implies $r\notinε'$.
\end{definition}
\begin{definition}
  A substitution $σ$ is {\em compatible } with a formula $f$
  if $σ$ is compatible with every effect expression and effect record
  in $f$.
\end{definition}
\begin{definition}
  Let $μ = [Λ\alist{χ}.x |-> f_x]$. Then $μ$ is {\em compatible} with $f$
  iff for any $x~\alist{κ}$ in $f$, $[\alist{χ} |-> κ ]$ is
  compatible with $f_x$.
\end{definition}

\subsubsection*{Soundness of the translation to SHOL}

\begin{prop}
  We have $s(r)σ = sσ(rσ)$ for any compatible substitution
  $σ$ and any effect record $s$ whose domain contains $r$.
  \label{prop:effrecsubstswap}
\end{prop}
\begin{proof}
  Let $σ=[ρ|->r_0]$. Then
    \begin{equation*}
      s(r)σ = \left(\bigoplus_{b\inε}bσ = s(b)σ\right)(rσ) = sσ(rσ)
    \end{equation*}
  On the other hand, if $σ=[\ev|->ε_0]$, then
  \begin{equation*}
  s(r)σ = (\bigoplus_{b\inε}s(b)σ \oplus ε_0^{s(\ev)})(r) = sσ(rσ)
  \end{equation*}
  We have used the fact that $ε_0$ does not contain $r$ (because it is
  compatible with the domain of $s$, and that $rσ=r$ for effect
  substitutions.
\end{proof}

\begin{definition}
  Let $f$ be an effect formula and $μ$ be a compatible
  value substitution. Then write $ μ||=f$ for $||= fμ$, or, equivalently, $
  |=[|fμ|]$. Thus, $μ$ contains the only free variables of $f$.
\end{definition}

\begin{lem}
  Let $s_0$ be an effect record of domain $ε_0$ such that 
  $Δ|-_r s_0 : ε_0$. Let $Δ,l:ε_0|-f:θ$ and $Δ,l:ε_0|-_rs : ε$. Also,
  let 
  \begin{eqnarray*}
μ &=& [b^l|->s_0(b)\mid b\in ε_0] \\
μ'&=&[b^l|->s_0σ(b)\mid b\inε_0σ]
  \end{eqnarray*}
  where $σ$ is some substitution compatible with $ε,ε_0$ and $f$. We have
  \begin{enumerate}
    \item $fμσ == fσμ'$,
    \item $sμσ == sμ'σ$.
  \end{enumerate}
  \label{lem:substequal}
\end{lem}
\begin{proof}
  We have to prove both statements using mutual induction over the
  structure of formulas and effect records. Let us start with the
  first one We start by proving the first statement. The only
  interesting cases are the cases where the substitutions come into
  play. This concerns effect abstraction, quantification, effect
  records and references to variables.
  \begin{description}
  \item[Case $!\,r^l$.]
    We can write the following equation:
    \begin{align*}
      (!\,r^l)μσ &== s(r)σ ==sσ(rσ) == (!\,(rσ)^l)μ' == (!\,r^l)σμ'
    \end{align*}
    We have used proposition~\ref{prop:effrecsubstswap}.
  \item[Case $λl:ε_1.f_1$.] We write the following chain of equations.
    \begin{align*}
      (λl_1:ε_1.f_1)μσ&== (λl_1:ε_1.f_1μ)σ\\
      &== λl_1:ε_1σ.f_1μσ\\
      &==λl_1:ε_1σ.f_1σμ'\\
      &==(λl_1:ε_1σ.f_1σ)μ'\\
      &==(λl_1:ε_1.f_1)σμ'
    \end{align*}
    We have used the induction hypothesis, and the fact that the
    variable substitutions $μ,μ'$ simply descend to the body of the
    abstraction.
  \item[Case $∀l:ε_1.f_1$]. Analogous.
  \item[Case $f_1~s_1$.]
    We have $(f~s)μσ = fμσ~sμσ = fσμ'~sσμ' =(f~s)σμ' $ by the induction
    hypotheses.
  \end{description}
  Now, we need to prove the second part of the claim. We distinguish
  the two different kinds of substitutions: for effects and for references.
  \begin{description}
  \item[Case $σ=\lbracket\ev|->ε_σ\rbracket$.] 
    We write
    \begin{align*}
       sμσ &== \left(\bigoplus_{b\in ε} s(b)μ\right)σ\\
      &==\bigoplus_{b\in ε} s(b)μσ \oplusε_σ^{s(\ev)μ} \\
      &==\bigoplus_{b\in ε}s(b)σμ'\oplus \bigoplus_{b\inε_σ}b^{s(\ev)}μ'\\
      &==\left(\bigoplus_{b\inε_1σ}s(b)σ\oplusε_0^{s(\ev)}\right)μ' \\
      &== sσμ'
    \end{align*}
    This step from the second to the third line requires some
    explanations. First, we have used the induction hypothesis for the
    references part of s. Second, let us give the proof that the right
    hand sides of $\oplus$ are indeed equal. For this, first suppose
    that $s(\ev) = l$. 
    \begin{description}
    \item[Subcase $\ev\inε_0$.] Then 
    \end{description}

Then we obtain the equality
    \begin{equation*}
      ε_σ^{s_0(\ev)} = \bigoplus_{b\inεσ}s_0(b)σ.
    \end{equation*}
    However, $σ$ does never change the term on the right, as
    $\ev\notin εσ$. Thus, the equation is true by definition. In the
    other case $s(\ev) \neq l$, we have

    However, the
    induction hypothesis only accounts for the concrete reference
    fields of $s_1$. For the others (effect variable fields), consider
    some effect variable $\ev_1\in dom(s_1) = ε_1$. We know that
    $\ev_1\notin ε_0$, because σ is compatible with $f$.  First
    suppose $\ev_1\inε$. Then the identity becomes $s(\ev_1)== 
    s(\ev_1)$, which is trivially true.  On the other hand, if
    $\ev_1\notin ε$, the substitutions do not apply, and we obtain
    $s_1(\ev_1) == s_1(\ev_1)$.
  \end{description}
\end{proof}

\begin{lem}[Preservation of Equality under Substitution] In effect
  logic, $μ||= f_1 = f_2$ implies $μ||=f_1σ = f_2σ$, for any (atomic)
  substitution which is compatible with $f_1,f_2$.
\end{lem}
\begin{proof}
  By induction over the last rule applied to derive the equality.
  The only case to consider is the case when the last rule applied is
  the equation scheme~(\ref{eq:eqschema}); all other
  cases do not involve effect abstraction. This means that the
  equation is of the form
\begin{equation*}
  (λl:ε.f)~s = fμ,\qquad μ = [ b^l|->s(b) \mid b\inε].
\end{equation*}
We now want to prove
\begin{equation*}
  ((λl:ε.f)~s)σ = (fμ)σ
\end{equation*}
for some substitution $σ=[\ev|->ε_0]$. We can write
\begin{align*}
  ((λl:ε.f)~s)σ &= (λl:εσ.fσ)~sσ \\
      &= fσμ'\quad\text{where}~μ' = [b^l|->sσ(b)\mid b\in εσ]\\
      &== fμσ\quad\text{by Lemma~\ref{lem:substequal}}
\end{align*}
\end{proof}

\begin{lem}[Preservation of Equality under Translation]
  $f = f'$ in effect logic implies $[|f|] = [|f'|]$ in SHOL.
\end{lem}
\begin{proof}
  A straightforward induction over the derivation of the proof of $f =
  f'$. In the case where the schema (\ref{eq:eqschema}) has been
  applied, one can obtain the same equality in SHOL by applying
  $β$-equality $n$ times, where $n$ is the number of fields of the
  effect record involved.
\end{proof}

In summary, we have shown that equation schema~(\ref{eq:eqschema}) is
actually {\em admissible} in effect logic, if one defines validity in
effect logic by the validity of the translation to SHOL, as we do. No
additional equation can be proved. Using the lemma about preservation
of equality under substitution, we can also state the following
theorem:
\begin{thm}
  For any $σ = [\ev |-> ε]$ and $f$ such that $σ$ is compatible with
  $fμ$ and $fμ$ monomorphic, we have $μ||=∀\ev.f$ implies $μ||=fσ$.
  \label{thm:formulasubstvalid}
\end{thm}

This theorem is essential to our calculus, as it directly applies to
the proof obligations for correctness of effect polymorphic
functions. It states that the correctness remains valid if one
instantiates the function with a (compatible) effect substitution. Our
typing constraints guarantee that no other effect instantiation can
occur.

\subsection{Lifting Values and Types}

\begin{definition}
  We define a function $\lceil \cdot \rceil$ from program values to logic
  expressions as follows:
  \begin{eqnarray*}
    \ceil{c} &=& c \\
    \ceil{x~\alist{ε }} &=& x~\alist{ε } \\
    \ceil{\lambda x\colon τ .\{p\}e\{q\}}  &=& (\lambda x:\ceil{τ }.p,\lambda x:\ceil{τ }.q)\\
    \ceil{\rec~f~(x:τ).\{p\}e\{q\}} &=&
    (\lambda x:\ceil{τ }.p,\lambda x:\ceil{τ }.q)
  \end{eqnarray*}
\end{definition}
\begin{definition}
  We define a function $\lceil \cdot \rceil$ from program types to logic types
  as follows:
  \begin{eqnarray*}
    \ceil{\iota } &=& \iota  \\
    \ceil{τ ->^ε τ '} &=& (\ceil{τ } -> ε
    -> \proptype) \times\\
    & & (\ceil{τ } -> ε  -> ε  -> \ceil{τ '}
    -> \proptype)\\
  \ceil{\forall \alist{χ}.τ } & =& \forall \alist{χ}.\ceil{τ }
  \end{eqnarray*}
\end{definition}
\begin{prop}
The following two simple properties are trivially verified.
  \begin{enumerate}
  \item 
  $\ceil{τ} [\alist{χ} |-> \alist{ε}] == \ceil{τ[\alist{χ} |-> \alist{ε}]}$
  \item
  $\ceil{v} [\alist{χ} |-> \alist{ε}] == \ceil{v[\alist{χ} |-> \alist{ε}]}$
  \end{enumerate}
  \label{prop:typevaleffsub}
\end{prop}

\subsection{Weakest preconditions}

\subsection*{Definition}
\begin{eqnarray*}
  c(x~\alist{ε } ) &==& \True \\
  c(c) &==& \True \\
  c(\lambda x\colonτ .\{p\}e\{q\}) &==& \forall x\colon\ceil{τ }.\forall
  l\colon ε.p~ε^l => \wpre_{ε ^l}(e,
q~ε^l)\\
  c(v = \rec~(y\colonτ ->^ε τ ')~(x \colon τ ).\{p\}e\{q\})
  &==& \forall y\colon\ceil{τ ->^ε τ '}.\forall
  x\colon\ceil{τ }.\forall l\colonε.p~ε^l => \\
  & & y = \ceil{v}  => \wpre_{ε ^l}(e,q~ε^l)\\
  & & \\
  \wpre_{\emptyset}(v,q) &==& q~\{\}~\ceil{v} /\ c(v) \\
  \wpre_{r^l}(!\,r,q) &==& q~\{r = r^l\}~!\,r^l \\
  \wpre_{\emptyset}(r:=v,q) &==& c(v) /\ q~\{r = \ceil{v}\}~() \\
  \wpre_{ε ^l}( v_{τ '->^ε  τ }~v', q) &==& 
  c(v) /\ c(v') /\ \pre~ \ceil{v}~ \ceil{v'}~ ε ^l /\
  \\
  & & \forall m\colonε.\forall x\colon\ceil{τ }. 
  \post~\ceil{v}~\ceil{v'}~ε ^l~ε ^m~x => q~ε ^m~x \\
  \wpre_{ε ^l}( \letst~x~\alist{χ} = v~\inst~e , q) &==& 
  \forall \alist{\ev}.c(v) /\ \wpre_{ε ^l}(e,q)[\Lambda
  \alist{\ev}. x |-> \ceil{v} ] \\
  \wpre_{(ε _1ε _2)^l}( \letst~x = {e_1}_{τ
    '}~\inst~e_2, q) &==&
  \wpre_{ε _1^l}(e_1,\lambda m:ε _1.\lambda x\colon\ceil{τ '}.
  \wpre_{ε _2\subseteq (ε _2^l\oplusε _1^m)}(e_2,q))\\
  \wpre_{ε^l}(\ifst~v~\thenst~e_1~\elsest~e_2,q) &==& \ceil{v} = \true =>
  \wpre_{ε_1\subseteqε^l}(e_1,q) /\ \\
  & & \ceil{v} = false => \wpre_{ε_2\subseteqε^l}(e_2,q) \\
  & & \\
  \wpre_{ε _1\subseteqε ^l}(e,q) &==&
  \wpre_{ε ^l|ε _1}(e,\lambda m\colonε _1.q~(ε ^l\oplus ε _1^m))
\end{eqnarray*}

\subsection*{Soundness of the \wpre-calculus}

\begin{prop}
  If $v$ is well-typed of type $τ$ using $\alist{χ}$ and
  $σ=[\alist{χ}|->\alist{ε}]$ such that $σ\simτ$, then $σ$ is
  compatible with $c(v)$.
  \label{prop:compatvcv}
\end{prop}
\begin{proof}
  This is trivial for $v \equiv x~\alist{ε}$ and $v \equiv c$, because
  $c(v) = \True$ in this case.  Now if $v == λx:τ.\{p\}e\{q\}$, then
  $τ= τ_1->^ετ_2$ and because of $σ\simτ$, $σ$ is compatible with
  $ε$. The formula
  \begin{equation*}
        c(v)== ∀x:\ceil{τ}.∀l:ε.p => \wpre_{ε^l}(e,λm:ε.q)
  \end{equation*}
  can only contain effects not larger than $ε$ (because of the side
  conditions when typing the $λ$-abstraction), and thus the claim is
  true. The case for $\rec$ is analogous.
\end{proof}

\begin{prop}
  For any (not necessarily atomic nor idempotent) effect substitution
  $σ$, we have
  \begin{equation*}
    \wpre(e,q)σ \equiv \wpre(eσ, qσ)
  \end{equation*}
  and
  \begin{equation*}
    c(v)σ \equiv c(vσ) 
  \end{equation*}
  \label{prop:wpceffsubeq}
\end{prop}
\begin{proof}
  Straightforward using prop.~\ref{prop:recsubst}.
\end{proof}

\begin{rem}
  Proposition \ref{prop:wpceffsubeq} together with Theorem
  \ref{thm:formulasubstvalid} gives us the following fact: If
  $∀\alist{χ}.c(v)$ is valid and $σ$ is compatible with
  $c(v)$, then $c(vσ)$ is valid as well.
  \label{rem:correcteffsub}
\end{rem}

The following is the main lemma. It plays a similar rôle as the
substitution lemma in subject reduction proofs. Let us introduce the
following convention: if we use effect substitution or generalized
value substitution, as in $fσ$ or $fμ$, we assume (or have to prove)
that $σ$, respectively $μ$, is compatible with $f$. We assume a
similar statement for $μ||=f $, where $μ$ must be compatible with $f$
and $[|fμ|]$ be monomorphic.

\begin{lem}
  Let $e$ be an expression, and let $ε$ and $τ$ be its effect and
  type.  Let $μ = [Λ\alist{χ}.y |-> v']$, $q,q'$ be formulas of type
  $ε->\ceil{τ}->\proptype$, and set $θ = \ceil{τ}$. Suppose for some
  $μ'$ we have
  the following premises:
  \begin{itemize}
  \item $μ'||=∀\alist{χ}.c(v') $;
  \item $y$ does not occur in $q'$;
  \item $μ'||=∀x:θ.∀l:ε. qμ~ε^l~x => q'~ε^l~x$
  \end{itemize}
  Then we have:
  \begin{itemize}
    \item $μ'||=∀l:ε.\wpre_{ε^l}(e,q)μ => \wpre_{ε^l}(eμ,q')$
    \item $μ'||=c(v)μ => c(vμ)$ 
    \item $μ'||=∀l:ε.\wpre_{ε'\subseteqε^l}(e,q)μ => \wpre_{ε'\subseteqε^l}(eμ,q')$
  \end{itemize}
  \label{lem:effsubweakeningwp}
\end{lem}
\begin{rem}
  \label{rem:effsubweakeningwp}
  Actually, we want to prove two simpler properties (with identical
  assumptions on $μ,v',q$ and $q'$):
  \begin{enumerate}
  \item $μ'||=∀x:θ.∀l:ε.q~ε^l~x => q'~ε^l~x$ implies $μ'||=∀l:ε.\wpre_{ε^l}(e,q) =>
    \wpre_{ε^l}(e,q')$ ;
    \item $μ'||=∀l:ε.\wpre_{ε^l}(e,q)μ => \wpre_{ε^l}(eμ,qμ)$.
  \end{enumerate}
  The first states that the weakest precondition of a stronger
  statement implies the one of a weaker statement; the second one
  states that the weakest precondition calculus stays valid if one
  substitutes a (correct) value for a variable. Both properties are
  necessary for the $\wpre$ calculus to be modular.

  The formulation of the $\wpre$ calculus, however, forces us to prove
  the two properties at the same time, hence Lemma
  \ref{lem:effsubweakeningwp}. It is easy to see that the two desired
  properties are implied by this lemma: Setting $μ = id$ we obtain the
  first property, and setting $q' = qμ$, we obtain the second. 
\end{rem}

\begin{proof}
  The third part of the lemma follows easily from the first, by
  expanding definitions. We will use this property, however, in some
  cases for the induction hypotheses of the first property.

  Let's prove the second part about correctness of values. In the case of
  constants, there is nothing to prove.
  \begin{description}
    \item[Case $λ$] 
      \begin{align*}
        μ'||=c(λx:τ.\{p\}e\{q\})μ ~&<=>~μ'||= (∀x:\ceil{τ}.∀l:ε. p~ε^l =>
        \wpre_{ε^l}(e,q~ε^l))μ \\
        & <=>~ μ'||= ∀x:\ceil{τ}.∀l:ε.pμ~ε^l => \wpre_{ε^l}(e,q~ε^l)μ \\
        & =>~μ'||= ∀x:\ceil{τ}.∀l:ε.pμ~ε^l => \wpre_{ε^l}(eμ,qμ~ε^l) \\
        &<=>~μ'||= c(λx:τ.\{pμ\}eμ\{qμ\})
      \end{align*}
      The step from line 2 to line 3 is possible because of the induction
      hypothesis on $\wpre(e,q)$ (see remark \ref{rem:effsubweakeningwp}). 
    \item[Case $\rec$] This case is entirely analogous.

    \item[Case $x~\alist{ε}$] If $x$ is different from $y$, there is
      nothing to prove. Otherwise, set $σ = [\ev|->ε]$. As
      $x~\alist{ε}$ is well-typed, we know that $x$ is of type
      $∀\ev.τ'$ such that $σ\simτ$. Now $v$, under the assumption
      $\alist{χ}$, must be of type $τ$. We thus know by proposition~
      \ref{prop:compatvcv} that $c(v)$ is compatible to $σ$. We have
      $c((x~\alist{ε}))= c(vσ) = c(v)σ$ (proposition
      \ref{prop:wpceffsubeq}) and we can apply theorem
      \ref{thm:formulasubstvalid} to conclude.
  \end{description}

  Now let's prove the part about \wpre.
  \begin{description}
    \item[Case $v$]
      \begin{align*}
        μ'||= \wpre_{ε^l}(v,q)μ~ &<=>~μ'||= (q~\{\}~\ceil{v})μ /\ c(v)μ\\
        & <=>~μ'||= qμ~\{\}~\ceil{v}μ /\ c(v)μ\\
        & => ~μ'||=q'~\{\}~\ceil{vμ} /\ c(vμ) \\
        &<=>~ μ'||=\wpre_{ε^l}(vμ,q')
      \end{align*}
      We can go from the second to the third line by using proposition
      \ref{prop:typevaleffsub}, the induction hypothesis on values and the
      hypothesis on the implication from $q$ to $q'$.

    \item[Case $!\,r$]  
      \begin{align*}
        μ'||= \wpre_{ε^l}(!\,r,q)μ &<=>~μ'||= qμ~\{\}~r^l\\
        & =>~μ'||= q'~\{\}~r^l\\
        & <=>~μ'||= \wpre_{ε^l}(!\,r,q')
      \end{align*}
      We have only used the condition on $q$ and $q'$.

    \item[Case $r:=v$ ] 
      \begin{align*}
        μ'||= \wpre_{ε^l}(r:=v,q)μ &<=>~μ'||= c(v)μ /\ qμ~\{r = \ceil{v}μ\}~\void\\
        & =>~μ'||= c(vμ) /\ q'~\{r = \ceil{vμ}\}~\void\\
        & <=>~μ'||= \wpre_{ε^l}(r:=vμ,q')
      \end{align*}
      We used the induction hypothesis on $c(v)$ and the implication
      between $q$ and $q'$.

   \item[Case app] This case is very similar to the cases before (but
     the manipulated formula is more complex), and the induction
     hypothesis on $c(v)$ is used twice.
   \item[Case monomorphic $let$ ] 
     Here we reason inside out: 
     \begin{multline*}
       μ'||= ∀x:θ.∀l:(ε_1ε_2). qμ~(ε_1ε_2)^l~x => q'(ε_1ε_2)^l~x\\
       =>μ'||=
       ∀l:(ε_1ε_2).∀ε_1^m.∀ε_2^o.∀z:θ.(λo:ε_2.qμ~(ε_1^m\oplusε_2^o))~ε_2^l~z
       => (λo:ε_2.q'~(ε_1^m\oplusε_2^o))~ε_2^l~z\
     \end{multline*}
     Now set $f(q) = λo:ε_2.q~(ε_1^m\oplusε_2^o)$ (note that $f(qμ) =
     f(q)μ$), so we can continue:
     \begin{equation*}
       μ'||=∀l:(ε_1ε_2).∀m:ε_1.\wpre_{(ε_2^l\oplusε_1^m)|_{ε_2}}(e_2,f(q))μ
       =>\wpre_{(ε_2^l\oplusε_1^m)|_{ε_2}}(e_2μ,f(q'))
      \end{equation*}
      Now, set
      \begin{equation*}
        g(e,q) =
        λm:ε_1.λx:\ceil{τ'}.\wpre_{(ε_2^l\oplusε_1^m)|_{ε_2}}(e,f(q))
        = \wpre_{ε_2\subseteq (ε_2^l\oplusε_1^m)}(e,q)
      \end{equation*}
      and we have
     \begin{align*}
       &μ'||=∀l:(ε_1ε_2).∀z:\ceil{τ'}.g(e_2,q)μ~ε_1^l~z => g(e_2μ,q')~ε_1^l~z\\
       =>~&μ'||=∀l:(ε_1ε_2).\wpre_{ε_1^l}(e_1,g(e_2,q))μ => \wpre_{ε_1^l}(e_1μ,g(e_2μ,q'))\\
       =>~&μ'||=∀l:(ε_1ε_2).\wpre_{ε_1^l}(\letst~x=e_1~\inst~e_2,q)μ => 
         \wpre_{ε_1^l}(\letst~x=e_1μ~\inst~e_2μ,q')
     \end{align*}
     We have used the induction hypothesis twice here, for $e_1$ and
     $e_2$ and for different formulas.

   \item[Case polymorphic $let$ ] 
     Let us write $μ_0$ for $[Λ\alist{χ}.x |-> v] $, $μ_0'$ for
     $[Λ\alist{χ}.x |-> vμ] $ and $μ_0μ$ for the combination of the two
     substitutions. Now,
     \begin{align*}
       μ'||= \wpre_{ε^l}(\letst~x~\alist{χ} = v : τ'~\inst~e,q)μ
       &<=>~μ'||= ∀\alist{χ}.c(v)μ /\ \wpre_{ε^l}(e,q)μ_0μ\\
         &<=>~μ'||= ∀\alist{χ}.c(v)μ /\ \wpre_{ε^l}(e,q)μμ_0'\\
         &=>~μ'||= ∀\alist{χ}.c(vμ) /\ \wpre_{ε^l}(eμ,q')μ_0'\\
         &<=>~μ'||= \wpre_{ε^l}(\letst~x~\alist{χ} = vμ : τ'~\inst~eμ,q')
     \end{align*}
     In the third line we can exchange the substitutions in this way because
     $v'$ does not contain $x$. In the next line, we apply the
     induction hypothesis, but on $μ'μ_0'||= \wpre_{ε_0}(e,q)μ$.
     \item[Case $if$] We have
       \begin{align*}
         &μ'||=\wpre_{ε^l}(\ifst~v~\thenst~e_1~\elsest~e_2,q)μ\\
         <=>& μ'||= \ceil{v}μ = true =>
         \wpre_{ε_2\subseteqε^l}(e_2,q)μ/\ 
         \ceil{v}μ = false => 
         \wpre_{ε_2\subseteqε^l}(e_2,q)μ\\
         =>& μ'||= \ceil{vμ} = true
         =>\wpre_{ε_2\subseteqε^l}(e_2μ,q')/\
         \ceil{vμ}
         = false =>
         \wpre_{ε_2\subseteqε^l}(e_2μ,q')\\
         <=>& μ'||= \wpre_{ε^l}((\ifst~v~\thenst~e_1~\elsest~e_2)μ,q')
       \end{align*}
  \end{description}
\end{proof}

Let us now proceed to the subject reduction theorem. This kind of
theorem always states that some property (typically the type of an
expression in type safety proofs) remains valid when reducing a term
using the reduction relation. Here, it's the validity of the
preconditions that is preserved; if the weakest precondition $p$ of $e$
and $q$ is valid at some point in the reduction process, the weakest
precondition $p'$ of $e'$ and $q$ after reduction is valid as well.

Let us define $ε_s$ (where $ε$ contains only references) as the effect
record of type $ε$ in which every reference of $ε$ is set to its
value according to $s$. Now define $μ,s||=_{ε} q$, where $q$ is of type
$ε->\proptype$, as $μ||=q~ε_s$. In the particular case of
$q=\wpre(e,q)$, this gives us: 
\begin{equation*}
μ,s||=_{ε}\wpre(e,q)~ <=>~μ||=(λl:ε.\wpre_{ε^l}(e,q))~ε_s~<=>~μ||=\wpre_{ε_s}(e,q). 
\end{equation*}
Therefore, in the proofs concerning the $\wpre$ calculus, we will use
the names of the references to represent their value in s.

In the following, we also use the notation $\wpre_{ε_1\subseteq
  ε_2}(e,q)$ (note the subtle difference with $\wpre_{ε_1\subseteq
  ε_2^l}(e,q)$, the difference is that the former contains two
effects, while the latter contains one effect and one effect
record). It is defined as
\begin{equation*}
  \wpre_{ε_1\subseteq ε_2}(e,q) = λl:ε_2.\wpre_{ε_1\subseteq
    ε_2^l}(e,q).
\end{equation*}

\begin{thm}[Subject Reduction]
  For any well-typed $e$ of effect $ε$, such that $s,e --\ s',e'$ and
  $μ,s||={_ε} \wpre(e,q)$, we have $μ,s'||=_{ε} \wpre_{ε'\subseteqε}(e',q)$,
  where $ε'$ is the effect of $e'$.
  \label{thm:basered}
\end{thm}
\begin{proof}
  By induction over the reduction of $s,e$.
  \begin{description}

  \item[Case $!\,r$] Let $s(r) = n$. In this case, $μ,s||=_r\wpre(!\,r,q)
    <=> μ||=q~\{r = n\}~n <=> μ,s||=_r\wpre_{∅\subseteq r}(n,q)$.

  \item[Case $r:=n$ ] $s,r:= n --\ s[r|-> n], \void $. We have
    $μ,s||=_r\wpre(r:=n,q) <=> μ||= c(v) /\ q~\{r=n\}~\void$ and
    $\wpre_{∅\subseteq r}(\void,q) = q~\{r=n\}~\void $ is identical.

    \item[Case $let$ without generalization] 
      Suppose first that no effect generalization takes place.
      Then $s, \letst~x = v~\inst~e --\ s, e[x|->v]$. We also have 
      \begin{align*}
      \wpre_{ε^l}(\letst~x = v : τ~\inst~ e , q) &=
  \wpre_{∅}(v,λm:().λx:\ceil{τ'}.\wpre_{ε^l}(e_2, q))\\
            &= c(v) /\  \wpre_{ε^l}(e_2,q)[x|->\ceil{v}]
      \end{align*}
      Using Lemma \ref{lem:effsubweakeningwp} and the fact that
      $q[x|->[v] ] => q $ (because $q$ does not contain $x$), this can
      also be written as $\wpre(e[x|->v],q)$.

    \item[Case $let$ with generalization] 
      We have 
      \begin{equation*}
      s, \letst~x~\alist{χ} = v~ \inst~ e --\ s, e[Λ\alist{χ}.x|->\ceil{v}] . 
      \end{equation*}
      We also have 
      \begin{equation*}
        \wpre_{ε^l}(\letst~x~\alist{χ} = v~\inst~e,q) = ∀\alist{χ}.c(v) /\ 
        \wpre_{ε^l}(e,q)[Λ\alist{χ}.x|->v ].
      \end{equation*}
      For the same reason as above, we have $\wpre(e[Λ\alist{χ}.x|->v],q)$,
      which is the desired formula. The substitution is compatible by
      definition of the $\wpre$ calculus.

    \item[Case $β$] In the following, let $v' = λy:τ.\{p'\}e\{q'\}$ and
      suppose that the result of application is of type $τ'$.
      We have $s, v'~v --\ s, e[y|-> v] $, and $s||= \wpre(v'~v, q)$.
      We want to obtain $s||=\wpre(e[y|->v],q)$.
       First, we have 
       \begin{equation}
         ||= p'[y|-> \ceil{v} ]~ε_s,
         \label{eq:beta}
       \end{equation}
       as the precondition of $v'$ applied to the argument. Now, the
       correctness of $v'$ is written
       \begin{equation*}
         c(v') = ∀l:ε.∀y:\ceil{τ}. p'~ε^l => \wpre_{ε^l}(e,q'~ε^l)
       \end{equation*}
       instantiating with $ε_s$ and $\ceil{v}$ and using
       \ref{eq:beta}, we derive 
       \begin{equation}
         ||= \wpre_{ε_s}(e,q'~ε_s)[y|->\ceil{v}]
         \label{eq:subst}
       \end{equation}
       We also have
       \begin{equation*}
         ∀m:ε.∀x:\ceil{τ'}.q'[y|->~\ceil{v}]~ε_s~ε^m~x => q~ε^m~x
       \end{equation*}
       By Lemma \ref{lem:effsubweakeningwp} and \ref{eq:subst}, we can conclude
       \begin{equation*}
         \wpre_{ε_s}(e[y|->\ceil{v}],q).
       \end{equation*}

     \item[Case $\rec$] Let $v' = \rec~(z:τ'->τ)~(y:τ').\{p\}e\{q\}$.
       The result of the application is of type $τ$. As in the $β$
       case, we have 
       \begin{equation}
         ||= p'[y|-> \ceil{v} ]~ε_s,
         \label{eq:recbeta}
       \end{equation}
       and we also have the correctness of both values, in particular
       for the recursive function:
       \begin{equation*}
         c(v') = ∀l:ε.∀z:\ceil{τ'->τ}.∀y:\ceil{τ'}. p'~ε^l => z = \ceil{v'} => \wpre_{ε^l}(e,q'~ε^l)
       \end{equation*}
       Instantiating with $ε_s$ for $ε_0$, $v'$ for $z$ and $v$ for
       $y$, we obtain again
       \begin{equation*}
         ||= \wpre_{ε_s}(e,q'~ε_s)[y|->\ceil{v}]
       \end{equation*}
       and we can proceed as in the case for $β$.
     \item[Case $\mathit{if}$] One of the assertions $\ceil{v} = \true$ and
       $\ceil{v} = \false$ is trivially true in both cases. In either
       case, we obtain precisely the desired statement by applying
       modus ponens.
  \end{description}
\end{proof}

\begin{thm}
  For any well-typed $e$ of effect $ε$, such that $s,e --> s',e'$ and
  $s||={_ε} \wpre(e,q)$, we have $s'||=_{ε} \wpre_{ε'\subseteqε}(e',q)$,
  where $ε'$ is the effect of $e'$.
  \label{thm:onestep}
\end{thm}
\begin{proof}
  There are only two cases here: the empty context and the
  $\letst$-context. The statement about the empty context is precisely
  Theorem~\ref{thm:basered}.

  In the remaining case, we have $s,E[e]-->s',E[e']$ and thus also
  \begin{equation*}
    s,\letst~x=E[e]~\inst~e_1 --> s', \letst~x=E[e']~\inst~e_1.
  \end{equation*}
  Our induction hypothesis is that $s||=_ε\wpre(E[e],q)$ implies
  $s'||=_ε\wpre_{ε'\subseteqε}(E[e'],q)$.  We also have
  $s||=_ε\wpre(\letst~x=E[e]~\inst~e_1,q)$, which amounts to
  \begin{equation*}
    s||=_ε \wpre(E[e],λm:ε.λx:\ceil{τ'}.\wpre_{ε_2\subseteq ε_s}(e_2,q))
  \end{equation*}
  By the induction hypothesis, we have 
  \begin{equation*}
    s'||=_ε \wpre_{ε'\subseteq
      ε}(E[e'],λm:ε.λx:\ceil{τ'}.\wpre_{ε_2\subseteq ε_s}(e_2,q))
  \end{equation*}.
\end{proof}

\begin{thm}
  For any well-typed $e$ of effect $ε$, such that $s,e->>s',e'$, we
  have $s||=_ε\wpre(e,q)$ implies $s'||=_ε\wpre_{ε'\subseteq ε}(e',q)$
\end{thm}
\begin{proof}
  There are two cases. In the case of reflexivity, this is
  obvious. For transitivity, assume $s,e->>s',e'$ and
  $s',e'->>s'',e''$. Also assume that the effects of $e'$ and $e''$
  are $ε_1$ and $ε_2$, respectively. It is easy to obtain
  $s'||=_ε\wpre_{ε'\subseteq ε}(e',q)$ by one of the induction
  hypotheses. By unrolling definitions, this implies
  \begin{equation*}
s'||=_{ε_1} \wpre(e',λm:ε_1.q~(ε_{s'}\oplusε_1^m))
  \end{equation*}
  and again, by applying the induction hypothesis, we obtain
  \begin{equation*}
    s''||=_{ε_1} \wpre_{ε_2\subseteq ε_1}(e'',λm:ε_1.q~(ε_{s'}\oplusε_1^m)).
  \end{equation*}
  This transforms into the desired formula by the following chain of
  equations:
  \begin{align*}
    &s''||=_{ε_1} \wpre_{ε_2\subseteq
      ε_1}(e'',λm:ε_1.q~(ε_{s'}\oplusε_1^m))\\
    <=>&
    ||=\wpre_{ε_2\subseteq {ε_1}_{s''}}(e'',λm:ε_1.q~(ε_{s'}\oplusε_1^m)) \\
    <=>&
    ||=\wpre_{{ε_1}_{s''}|ε_2}(e'',λo:ε_2.(λm:ε_1.q~(ε_{s'}\oplusε_1^m))
    ({ε_1}_{s''} \oplus ε_2^o)) \\
    <=>&
    ||=\wpre_{{ε_1}_{s''}|ε_2}(e'',λo:ε_2.q~(ε_{s'}\oplus({ε_1}_{s''} \oplus ε_2^o))) \\
    <=> &||=\wpre_{ε_{s''}|ε_2}(e'',λo:ε_2.q~(ε_{s''}\oplus ε_2^o)) \\
    <=> &||=\wpre_{ε_2\subseteq ε_{s''}}(e'',q) \\
    <=> &s''||=_{ε}\wpre_{ε_2\subseteq ε}(e'',q) 
  \end{align*}
  From line 3 to line 4, we first exploit the fact that $ε_{s''}$
  restricted to $ε_2$ is the same as ${ε_1}_{s''}$, because of
  $ε_2\subseteq ε_1\subseteq ε$. For the same reason, and the
  additional fact that the only difference between $ε_{s'}$ and
  ${ε_1}_{s''}$ lies in the domain of $ε_2$, we can simplify the
  $\oplus$ chain to the right form.
\end{proof}

\begin{cor}
  For any $e$ and $q$ and $s,e->>s',v$, we have $s||=_ε\wpre(e,q)$
  implies $||=q~\ceil{v}~ε_{s'}$.
\end{cor}
\begin{proof}
  Direct consequence from the previous theorem and the definition of
  the $\wpre$ for values.
\end{proof}

\end{document}

### 2009-10-06 papers/esop2010/main.tex
go

--- 2009-10-06 papers/esop2010/main.tex
@@
 \institute{
-INRIA Saclay - Île-de-France, ProVal, Orsay, F-91893 \and
+INRIA Saclay - Île-de-France, ProVal, Orsay, F-91894 \and
  LRI, Univ Paris-Sud, CNRS, Orsay, F-91405
@@
 %<-> domain, region <-> key
-%TODO the constants dealing with maps
-
-%TODO clearly state possibilities/limitations of the system
-% * one can store everything in refs
-% * but aliasing restriction (refs in refs, lists of refs etc)
 
@@
   higher-order logic.  To our knowledge, this is the first time that effect
-  polymorphism is put to use in program verification. Additionally, we propose
-  a region-based analysis to exclude aliasing, in order to obtain much simpler
-  proof obligations. Several examples demonstrate the practicality of the
-  system. A prototype implementation exists and has been used to prove the
-  correctness of these and more examples.  \end{abstract}
+  polymorphism is put to use in program verification.  A simple mechanism
+  based on regions permits to restrict the analysis to alias-free programs,
+  thus obtaining much simpler proof obligations. Several examples demonstrate
+  the practicality of the system. A prototype implementation exists and has
+  been used to prove the correctness of these and more examples.
+\end{abstract}
 
@@
 
-The purpose of the present paper is to propose a language suited for reasoning
-about effectful higher-order programs.  Higher-order functions are a key
-feature, not only of purely functional languages like Haskell and the pure
+The purpose of the present paper is to introduce a simple weakest precondition
+calculus for imperative higher-order programs. Higher-order functions are a
+key feature, not only of purely functional languages like Haskell and the pure
 fragment of ML, but also in combination with imperative features, like
-references or arrays. However, reasoning about higher-order functions is
-difficult, and especially so in the presence of effectful computations. Any
-function's specification depends on its arguments. In a higher-order setting,
-arguments may be functions that have their own specification. In consequence,
-to be modular, the specification of a higher-order function can only be
-formulated {\em depending on the specifications of its functional arguments}.
-In a setting where side effects are possible, one also needs to formulate the
+references. However, reasoning about higher-order functions is difficult, and
+especially so in the presence of effectful computations. Any function's
+specification depends on its arguments. In a higher-order setting, arguments
+may be functions that have their own specification. In consequence, to be
+modular, the specification of a higher-order function can only be formulated
+{\em depending on the specifications of its functional arguments}. In a
+setting where side effects are possible, one also needs to formulate the
 specification {\em depending on the side effects of the functional arguments}.
@@
 specifications can be expressed in higher-order logic, to be able to state
-properties of functional arguments.  We then proceed to define a weakest
-precondition calculus for this language, transforming an annotated
-program into a logic formula, which has to be valid for the program to be
-correct. This calculus is very straight-forward and syntax-directed.
-
-Programs in our language may contain aliases. However, we obtain much simpler
-proof obligations when aliases are excluded. Therefore, we also propose a
-static region-based analysis rejecting programs with aliasing. The weakest
-precondition calculus on the alias-free fragment has been implemented in a
-prototype tool called \who~\cite{KanigFilliatre09wml}.  We show the usefulness
-of this tool on several examples, which have all been proved correct.
+properties of functional arguments. We show how two very simple modifications
+to the system let us obtain alias-free programs, with the aim of obtaining
+much simpler proof obligations.
+
+We then proceed to define a weakest precondition calculus based on this
+language, transforming an annotated program into a logic formula, which has to
+be valid for the program to be correct. This calculus is very straight-forward
+and syntax-directed.
+
+Finally, we show the usefulness of the system by giving several examples of
+annotated programs. All the programs have been proved correct using our
+prototype implementation called \who.
 
 \paragraph{Outline.}
-Section~\ref{sec:genpic} introduces our language using a motivating example.
-Sections~\ref{sec:syntax} and \ref{sec:wp} give the details about the
-programming language and the weakest precondition calculus.  In
-section~\ref{sec:examples} we present our implementation and show some
+Section~\ref{sec:genpic} gives an introduction to the described language as
+well as an example. Sections~\ref{sec:syntax} and \ref{sec:wp} give the
+details about the programming language and the weakest precondition calculus.
+In section~\ref{sec:examples} we speak about the implementation and show some
 examples. Finally, section~\ref{sec:related} discusses related work.
@@
 Consider the following first-order program\footnote{Throughout the document,
-we use \ocaml~syntax for programs.} that accumulates the sum of the contents
-of some array {\tt a} in some global reference {\tt c}, using an integer {\tt
-i} as a counter:
+we use \ocaml~Syntax for programs.} that accumulates the sum of the contents
+of some array $a$ in some global reference $c$, using an integer $i$ as
+a counter:
 \begin{ocaml}
@@
       c := !c + a.(i)
-    done
 \end{ocaml}
-This simple program can be annotated in a standard Hoare logic (for
+This simple  can be annotated in a standard Hoare logic (for
 example the system of \cite{Filliatre00a}) as follows:
@@
   let sum () =
-    { !c = 0 }
-    for i = 0 to Array.length a - 1 do
-      { !c = {{{ $\sum_0^{i-1}$ }}}  a.(i) }
-      c := !c + a.(i)
-    done
-    { !c = {{{ $\sum_0^{ {\tt Array.length~a} - 1}$ }}}  a.(i) }
+  {!!c = 0}
+    for i = 0 to length !a - 1 do
+    { !!c = {{{ $\sum_0^{i-1}$ get !!a i }}} }
+
+      c := !c + get !a i
+
+      {!!c = {{{ $\sum_0^{ {\tt length~!a} - 1}$ get !!a i }}} }
 \end{who}
-Our program is a {\tt for} loop, so it requires an {\em invariant
-annotation}, indicating which relation is preserved by the loop body. The
-equalities that need to be proved are very simple. Also, it is obvious that
-the program only modifies the reference {\tt c}, and this fact can be
-formally confirmed by an effect analysis.
+
+Here, assuming we model the array in the programming language by a reference
+to a functional array, with functions {\tt length} and {\tt get} with the
+obvious meaning. Also, in the logic we use a derefence operator {\tt !!}
+instead of the usual {\tt !}. In this case, our program is a {\tt for} loop,
+so it needs an {\em invariant annotation}, indicating which relation is
+preserved by the loop body. The equalities that need to be proved are very
+simple. Also, it is obvious that the program modifies exactly the references
+$r$ and $c$, and this fact can be formally confirmed by an effect analysis.
 
 A common reaction of a functional programmer to a program like
-{\tt sum} is to factor out the structure of the recursion, which
-may be useful in other situations, in a higher-order function. In this case,
-the more general function is the following iterator over arrays:
+$\mathit{sum}$ is to factor out the structure of the recursion, which
+may be useful in other situations, in a higher-order function. The
+more specific program is then recovered by function application. In this case,
+the more general function is of course simply {\tt Array.iter}:
 \begin{ocaml}
@@
       f a.(i)
-    done
-\end{ocaml}
-The more specific program is then recovered by function application. 
-\begin{ocaml}
+
   let sum () =
@@
 thing. However, how do we prove the correctness of the second one? One
-possibility is to impose the future usage of ${\tt iter}$ by
-specifying a particular specification for its argument {\tt f}:
+possibility is to fix the future usage of ${\tt iter}$ by
+specifying what exactly is the specification of its argument {\tt f}:
 \begin{who}
@@
     (f : (x : int) ->{c} 
-        { !c = {{{ $\sum_0^{i-1}$ }}} a.(i) } 
+        { !!c = {{{ $\sum_0^{i-1}$ get !!a i }}} } 
         unit 
-        { !c = {{{ $\sum_0^{ {\tt Array.length~a} - 1}$ }}} a.(i) } ) 
+        { !!c = {{{ $\sum_0^{length !a - 1}$ get !!a i }}} } ) 
     (a : int array) =
@@
 Here we impose precisely the specification of {\tt f}, and thus we can
-derive that {\tt sum} computes the expected value,
+derive that {\tt sum} computes the right value,
 using the same method as for the first-order case before. Note
 that, to be able to apply an effect analysis, we have to specify that
-$f$ modifies $c$, by giving it the type $\intml->^{\{c\}}\unitml$.
+$f$ modifies $c$, by giving it the type $\intml->^c\unitml$.
 
@@
 and can only compute a certain sum. If one wants to use {\tt iter} in another
-context, one has to copy the code, give it a new specification adapted to the
+context, one has to copy the code, give a new specification adapted to the
 application and prove this new specification again. In addition, in order to
@@
 
-There are two obstacles related to higher-order functions to be overcome in
-order to achieve modularity on the level of specification in the presence of
-side effects. We must be able to refer to the effect of a function parameter,
-and we must be able to refer to its specification (pre- and postconditions),
-in a generic way.
+There are two obstacles related to higher-order functions to be
+overcome in order to achieve modularity on the level of specification
+in the presence of side effects. We must be able to refer to the
+effect of a function parameter, and we must be able to refer to the
+specification (pre- and postconditions) of a function parameter, in a
+generic way.
 
 The first obstacle has a well known solution using a type and effect
-system~\cite{talpin94taed} with effect variables: effect polymorphism does
-exactly what we want here. Using effect polymorphism, we can say that {\tt f}
-(the argument of {\tt iter}) has type $α->^e\unitml$, whera $α$ is the type of
-the array elements and where $e$ is an {\em effect variable}, which amounts to
-say that {\tt f} has a {\em certain} effect $e$. The complete type of {\tt
-iter} is thus:
-\begin{displaymath}
-  ∀α∀e. (α ->^e \unitml) ->^\emptyset α~\mathtt{array} ->^e \unitml
-\end{displaymath}
-The total effect of {\tt iter} is then $e$, meaning {\tt iter}
+system~\cite{talpin94taed} with effect variables: effect polymorphism
+does exactly what we want here. Using effect polymorphism, we can say
+that {\tt f} (the argument of {\tt iter}) has type
+$\intml->^e\unitml$, where $e$ is an {\em effect variable}, which
+amounts to say that {\tt f} has a {\em certain} effect $e$. The total
+effect of {\tt iter} is then $e$, meaning {\tt iter}
 modifies whatever {\tt f} modifies.
@@
 of the function. The postcondition takes the argument and return value, but
-also the states before {\em and} after the function call. Two-state
-postconditions are a standard way to get rid of most auxiliary variables.
-However, the pre- and postcondition of {\tt f} may not refer to arbitrary
-parts of the state, but only the part {\tt f} reads or modifies.  With these
-conventions, one can now write a generic implementation and specification for
-the {\tt iter} function. It is given~Fig.~\ref{fig:iter}, in the input syntax
-of our prototype \who. There are a few syntactic differences with \ocaml, in
-particular for arrays.
-
-%Here, assuming we model the array in the programming language by a reference
-%to a functional array, with functions {\tt length} and {\tt get} with the
-%obvious meaning. Also, in the logic we use a derefence operator {\tt !!}
-%instead of the usual {\tt !}.
+also the states before {\em and} after the function call. Th so-called
+two-state postconditions are a way to get rid of most uses of auxiliary
+variables. However, the pre- and postcondition of {\tt f} may not refer to
+arbitrary parts of the state, but only the part {\tt f} reads or modifies.
+With these conventions, one can now write a generic specification for the {\tt
+iter} function, see ~Fig.~\ref{fig:iter}.
 
@@
 
-There are several new elements here. First of all, we use effect polymorphism.
-An effect variable {\tt e}, standing for the effect of {\tt f}, is introduced
-by the notation ${\mathtt [e]}$.  {\tt iter} also has an additional argument
-{\tt inv}, the {\em invariant}, taking as argument the portion of the state
-modified by {\tt f} and the integer indicating the current iteration count.
-This is a logical argument, which means that it cannot be used in the program.
-Also, {\tt inv} is a {\em pure} function, which is indicated by the arrows
-without effect annotation. The postcondition is now quite clear: we simply
-want to establish the invariant for the final state (the keyword {\tt cur}
-stands for the current state of the postcondition) and the length of the
-array, indicating that we have iterated over the entire array. It is also easy
-to see that the invariant {\tt inv} is precisely the invariant of the {\tt
-for} loop.  However, the precondition is a bit complicated: it states that we
-need the invariant to hold initially (for the current state of the
-precondition and index $0$) and we need the invariant to be preserved by the
-function call.  This is expressed by quantifying over two states, saying that
-the invariant has to imply the precondition of {\tt f} and the postcondition
-has to imply the invariant for the next array index.  With a bit of syntactic
-sugar, this could be expressed much more concisely using a Hoare triple:
+There are several new elements here. First of all, we use effect polymorphism
+to introduce an effect variable {\tt e} standing for the effect of {\tt f}.
+Also, {\tt iter} has an additional argument {\tt inv}, the {\em invariant},
+taking as argument the portion of the state modified by f and the integer
+indicating the current iteration count. This is a logical argument, which
+means that it cannot be used in the program. Also, {\tt inv} is a {\em pure}
+function, which is indicated by the arrows without effect annotation. The
+postcondition is now quite clear: we simply want to establish the invariant
+for the final state (the keyword {\tt cur} for the current state of the
+postcondition) and the length of the array, indicating that we have iterated
+over the entire array. It is also easy to see that the invariant {\tt inv} is
+precisely the invariant of the {\tt for} loop. However, the precondition is a
+bit complicated: it states that we need the invariant to hold initially (for
+the current state of the precondition and index $0$) and we need the invariant
+to be preserved by the function call. This is expressed by quantifying over
+two states, saying that the invariant has to imply the precondition of {\tt f}
+and the postcondition has to imply the invariant for the next array index.
+With a bit of syntactic sugar, this could be expressed much more concisely
+using a Hoare triple:
 \begin{who}
-  { inv cur i } f (get i !!a) { inv cur (i+1) } 
+  { inv cur i} f (get i !!a) {inv cur (i+1)}
 \end{who}
 
-% TODO rewrite
-%The {\tt iter} function and its specification raise a question: is the
-%specification correct if {\tt f} modifies the array {\tt a}, which is required
-%for the good behavior of {\tt iter}? The answer is clearly no; if {\tt f}
-%messes up the contents of {\tt a}, {\tt iter} may not actually maintain the
-%invariant.  This is an instance of the {\em aliasing problem} (see, for
-%example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
-%does not support aliasing. In this paper we make the same choice and disallow
-%aliasing in the programming language we introduce. Under this assumption, the
-%specification we gave is correct. 
-%
-%Entirely excluding aliasing renders our system inappropriate for
-%reasoning about shared mutable structures. However, it is interesting
-%to see that our alias exclusion restriction, in many cases,
-%corresponds to a natural condition imposed on higher-order functions
-%that operate on some data structure. It is a common
-%requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration
-%functions, for example, that the functional argument must not modify
-%the structure to be iterated on. In our system, this fact follows from
-%the restriction on effect instantiations, as will be seen in
-%section~\ref{sec:syntax}. In the example, the array {\tt a}
-%of the iterating function {\tt iter} must not be modified by its
-%argument.
+The {\tt iter} function and its specification raise a question: is the
+specification correct if {\tt f} modifies the array {\tt a}, which is required
+for the good behavior of {\tt iter}? The answer is clearly no; if {\tt f}
+messes up the contents of {\tt a}, {\tt iter} may not actually maintain the
+invariant.  This is an instance of the {\em aliasing problem} (see, for
+example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
+does not support aliasing. In this paper we make the same choice and disallow
+aliasing in the programming language we introduce. Under this assumption, the
+specification we gave is correct.
+
+Entirely excluding aliasing renders our system inappropriate for
+reasoning about shared mutable structures. However, it is interesting
+to see that our alias exclusion restriction, in many cases,
+corresponds to a natural condition imposed on higher-order functions
+that operate on some data structure. It is a common
+requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration
+functions, for example, that the functional argument must not modify
+the structure to be iterated on. In our system, this fact follows from
+the restriction on effect instantiations, as will be seen in
+section~\ref{sec:syntax}. In the example, the array {\tt a}
+of the iterating function {\tt iter} must not be modified by its
+argument.
 
@@
 variable stands for an unknown effect. Effects are used to specify
-the memory locations an expression can write. They are also used in function
+of memory locations an expression can write. They are also used in function
 types to describe the memory locations which may be written if the function is
@@
 \begin{figure}[tbp]
+  \framebox{$Γ|-_v t : τ$}
   \begin{mathpar}
@@
 wrt. the typing relation $|-_l$ described later on. Finally, pure applications
-between values are considered to be values as well, as they don't have any
+between values are considered as values as well, as they don't have any
 effect.
@@
 \begin{figure}[tpb]
+  \framebox{$Γ|- t : τ,ε$}
 \begin{mathpar}
@@
 are \letml-introduced. This is not a restriction, as any program can be
-transformed into A-normal form.
+transformed an A-normal form.
 
@@
 effects. In the other case (rule {\sc Let}), no generalization takes place,
-but both terms can have effects. The {\sc Let} rule actually requires both
+but both terms can have effects. The {\sc Let} rule actually obliges both
 terms to have the same effect, but together with the {\sc Sub} rule, which can
@@
 \begin{figure}[tbp]
+  \framebox{$Γ|-_l t : τ$}
   \begin{mathpar}
@@
     { Γ |-_l t : \ceil{τ} }
-  } \qquad \and
+  } \and
   { \inferrule*[Left=LetLogic]
@@
 any logical connectives. These can of course be declared independently, for
-example the conjunction is just a function: 
-\begin{equation*} 
-  /\ :  \propml -> \propml -> \propml 
-\end{equation*} 
-Actually, when equality is given, all logical operations (even quantification)
-can be defined in the logic~\cite{andrews86}.  In the remainder of the paper,
-we suppose predefined the usual logical connectives conjunction, disjunction,
+example the conjunction is just a function: \begin{equation*} /\ :  \propml ->
+\propml -> \propml \end{equation*} Actually, when equality is given, all
+logical operations (even quantification) can be defined in the
+logic~\cite{andrews86}.  In the remainder of the paper, we suppose
+predefined the usual logical connectives conjunction, disjunction,
 implication and equivalence, as well as universal and existential
 quantification. We also assume all the arithmetic operations as well as the
-constructor $(t,t)$  and the accessors $\mathit{fst}$ and $\mathit{snd}$ for
-pairs. We will also use the names \preml\ and \postml\ for $\mathit{fst}$ and
-$\mathit{snd}$, respectively.
+constructor $(t,t)$  and the accessors $fst$ and $snd$ for pairs. We will also
+use the names \preml\ and \postml\ for $fst$ and $snd$, respectively.
 
@@
 We haven't mentioned references yet, but all we need to support them is to add
-the three usual functions $\mathit{ref}, :=, !$ for reference creation, assignment and
+the three usual functions $ref, :=, !$ for reference creation, assignment and
 lookup, respectively, to the environment. In our system, these functions have
@@
 \begin{eqnarray*}
-  \mathit{ref} &:& ∀αρ.α ->^ρ α~\refml_ρ\\
+  ref &:& ∀αρ.α ->^ρ α~\refml_ρ\\
   := &:& ∀αρ.α~\refml_ρ ->^\emptyset α ->^ρ \unitml\\
@@
 A way out is to restrict programs to alias free programs. Fortunately, it is
-easy to modify the typing rules to render all programs with aliasing
+easy to modify the presented typing rules to render all programs with aliasing
 ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
-in such a way that there is exactly one memory location in each region.
-Starting from this assumption, a program can actually
+to identify memory locations entirely; each region shall describe precisely
+one memory location. Starting from this assumption, a program can actually
 exhibit {\em two} forms of aliasing. In the first one, it may use several
@@
 this, we add a side condition to the instantiation: the substitution $σ$ must
-be {\em compatible} with the type $τ$, written $σ\sim τ$. To understand what
-compatibility means, consider an {\em atomic} region substitution (whose
-domain contains only one variable) $[ρ|->ρ']$. It is compatible with any type
-which does not contain $ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2
-e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
-Expressed more generally, an atomic substitution is compatible with a type if
-the type does not contain any region or effect variables which are in the
-range of the substitution. Finally, a {\em composite} substitution (whose
-domain contains more than one variable) is compatible with a type if each
-atomic substitution is compatible with the result of the application of the
-previous substitution. In this way, it becomes impossible to have different
-names for a single reference cell.
+be {\em compatible} with the type $τ$. To understand what compatibility means,
+consider an {\em atomic} region substitution (whose domain contains only one
+variable) $[ρ|->ρ']$. It is compatible with any type which does not contain
+$ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2 e'$] is compatible with
+any type which does not contain $ ρ1 , ρ2 $ nor $e'$.  Expressed more
+generally, an atomic substitution is compatible with a type if the type does
+not contain any region or effect variables which are in the range of the
+substitution. Finally, a {\em composite} substitution (whose domain contains more
+than one variable) is compatible
+with a type if each atomic substitution is compatible with the result of the
+application of the previous substitution. In this way, it becomes impossible
+to have different names for a single reference cell.
 
 The second form of aliasing is when we use the same name for {\em different}
-memory locations. This can happen when we allow the user to call
-$\mathit{ref}$ several times with the same region instantiation. A simple
-possibility is to  suppress the $\mathit{ref}$ function entirely and to
-replace it by a \letml\refml\ construct which combines name creation and
-memory allocation. In this way, there can of course be only one reference per
-name. On the other hand, because of the side condition $ρ\notin τ$, the
-reference cannot escape the scope of its \letml\refml\ construct, and so, a
-function cannot return a reference which has been created in its body. This
-restriction could be leveraged a bit if one separates name creation and
-allocation, while still maintaining that one can only allocate a single
-reference in each region. This is how it is done in the correctness proof of
-this restriction. The presented \letml\refml\ construct is of course a special
-case.
+memory locations. This can happen when we allow the user to call $ref$ several
+times with the same region instantiation. A simple possibility is to attach
+the suppress the $ref$ function entirely and to replace it by a \letml\refml\ 
+construct which combines name creation and memory allocation. In this way,
+there can of course be only one reference per name. On the other hand, because
+of the side condition $ρ\notin τ$, the reference cannot escape the scope of
+its \letml\refml\ construct, and so, a function cannot return a reference
+which has been created in its body. This restriction could be leveraged a bit if
+one separates name creation and allocation, while still maintaining that one
+can only allocate a single reference in each region. This is how it is done in
+the correctness proof of this restriction. The presented \letml\refml\
+construct is of course a special case.
+%TODO clearly state possibilities/limitations of the system
+% * one can store everything in refs
+% * but aliasing restriction (refs in refs, lists of refs etc)
+
 %TODO now, all operations on maps can be statically resolved.
@@
 the region calculus of Tofte/Talpin~\cite{tofte97ic}.
-%TODO ref vers l'annexe -  remarque sur la difference
-
-\section{A Weakest Preconditions Calculus}
+
+\section{A Simple Weakest Preconditions Calculus}
 \label{sec:wp}
@@
 The aim of the weakest precondition calculus is to obtain proof obligations
-from annotated programs. To do this, the calculus starts from a formula
+from the annotated program. To do this, the calculus starts from a formula
 which holds, or is supposed to hold, at the end of the program, and walks
@@
 hold at the beginning of the program (hence the name {\em weakest
-precondition}).
+precondition}.
 
@@
 \letml-binding in the language. In the other case, the \wpre\ of $ t2 $ becomes
-the argument of the \wpre\ of $ t1 $ (we denote $ε$ the effect of $ t1 $ and $
+the argument of the \wpre of $ t1 $ (we denote $ε$ the effect of $ t1 $ and $
 t2 $). The rule for \ifml-expressions is straightforward. The rule for

### 2009-10-06 papers/esop2010/main.tex

--- 2009-10-06 papers/esop2010/main.tex
@@
 \end{eqnarray*}
-
-\subsection{Alias restriction}
-
-The presented programming language, along with the specifications, can, in
-fact be used to prove programs. The weakest precondition calculus presented in
-the next section can be used to obtain proof obligations which, when proved,
-guarantee the correctness of the program with respect to its specification. On
-the other hand, in its presented form, and because of the presence of
-aliasing, one would constantly need equalities and disequalities between
-reference variables to be able to reason about a program.
-
-\begin{figure}[tpb]
-  \begin{mathpar}
-    {\inferrule*[Left=Var]
-      {Γ(x) = ∀\alist{χ}.τ \\ σ = [\alist{χ}|->\alist{κ}] \\ σ \sim τ }
-      {Γ |- x : τσ }
-    } \\
-    { \inferrule*[Left=LetRef, Right={ρ \textup{fresh}, $ρ\notin τ$}]
-      {Γ |-_v t1 : τ' \\ Γ, x : τ'~\refml_ρ |- t2 : τ, ε }
-      {Γ |- \letml~x = ref~ t1 ~\inml~ t2 : τ, ε \setminus ρ}
-    }
-  \end{mathpar}
-  \caption{A modified and a new rule for aliasing}
-  \label{fig:aliasing}
-\end{figure}
-
-A way out is to restrict programs to alias free programs. Fortunately, it is
-easy to modify the typing rules to render all programs with aliasing
-ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
-in such a way that there is exactly one memory location in each region.
-Starting from this assumption, a program can actually
-exhibit {\em two} forms of aliasing. In the first one, it may use several
-names (region variables) for the same memory location. This can happen when a
-function is generalized over, say, two region variables, assuming they are
-different, and is then instantiated with the same region variable twice. The
-classical example is the following program
-\begin{who}
-  let f [ρ ρ'] (x : int ref(ρ)) (y : int ref(ρ')) =
-  { }
-  x := !x + 1;
-  !y
-  { r : r = !!y old }
-\end{who}
-which is only correct if $ ρ $ and $ ρ' $ are indeed different. To enforce
-this, we add a side condition to the instantiation: the substitution $σ$ must
-be {\em compatible} with the type $τ$, written $σ\sim τ$. To understand what
-compatibility means, consider an {\em atomic} region substitution (whose
-domain contains only one variable) $[ρ|->ρ']$. It is compatible with any type
-which does not contain $ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2
-e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
-Expressed more generally, an atomic substitution is compatible with a type if
-the type does not contain any region or effect variables which are in the
-range of the substitution. Finally, a {\em composite} substitution (whose
-domain contains more than one variable) is compatible with a type if each
-atomic substitution is compatible with the result of the application of the
-previous substitution. In this way, it becomes impossible to have different
-names for a single reference cell.
-
-The second form of aliasing is when we use the same name for {\em different}
-memory locations. This can happen when we allow the user to call
-$\mathit{ref}$ several times with the same region instantiation. A simple
-possibility is to  suppress the $\mathit{ref}$ function entirely and to
-replace it by a \letml\refml\ construct which combines name creation and
-memory allocation. In this way, there can of course be only one reference per
-name. On the other hand, because of the side condition $ρ\notin τ$, the
-reference cannot escape the scope of its \letml\refml\ construct, and so, a
-function cannot return a reference which has been created in its body. This
-restriction could be leveraged a bit if one separates name creation and
-allocation, while still maintaining that one can only allocate a single
-reference in each region. This is how it is done in the correctness proof of
-this restriction. The presented \letml\refml\ construct is of course a special
-case.
-
-In alias-free programs, the operations $\oplus$ and $|_ε$ on stores we
-previously introduced can be executed {\em statically}, i.e. just by looking
-at the domains of the stores. Also, we statically know which value (or which
-named component) the accessor {\tt !!} will get from a store.
-
-\subsection{Some properties}
-
-%TODO
-$Γ|-_v v : τ$ implies $Γ |-_l \ceil{v} : \ceil{τ}$
-
-Subject reduction of our type system is a consequence of subject reduction of
-the region calculus of Tofte/Talpin~\cite{tofte97ic}.
-%TODO ref vers l'annexe -  remarque sur la difference
 
@@
 
+\subsection{Alias restriction}
+
+The presented programming language, together with the weakest precondition
+calculus, form a basis for program verification and can already be used to
+prove programs. On the other hand, in its presented form, and because of the
+presence of aliasing, one would constantly need equalities and disequalities
+between reference variables to be able to reason about a program.
+
+\begin{figure}[tpb]
+  \begin{mathpar}
+    {\inferrule*[Left=Var]
+      {Γ(x) = ∀\alist{χ}.τ \\ σ = [\alist{χ}|->\alist{κ}] \\ σ \sim τ }
+      {Γ |- x : τσ }
+    } \\
+    { \inferrule*[Left=LetRef, Right={ρ \textup{fresh}, $ρ\notin τ$}]
+      {Γ |-_v t1 : τ' \\ Γ, x : τ'~\refml_ρ |- t2 : τ, ε }
+      {Γ |- \letml~x = ref~ t1 ~\inml~ t2 : τ, ε \setminus ρ}
+    }
+  \end{mathpar}
+  \caption{A modified and a new rule for aliasing}
+  \label{fig:aliasing}
+\end{figure}
+
+A way out is to restrict the attention to alias free programs. Fortunately, it
+is easy to modify the typing rules to render all programs with aliasing
+ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
+in such a way that there is exactly one memory location in each region.
+Starting from this assumption, a program can actually exhibit {\em two} forms
+of aliasing. In the first one, it may use several names (region variables) for
+the same memory location. This can happen when a function is generalized over,
+say, two region variables, assuming they are different, and is then
+instantiated with the same region variable twice. The classical example is the
+following program
+\begin{who}
+  let f [ρ ρ'] (x : int ref(ρ)) (y : int ref(ρ')) =
+  { }
+  x := !x + 1;
+  !y
+  { r : r = !!y old }
+\end{who}
+which is only correct if $ ρ $ and $ ρ' $ are indeed different. To enforce
+this, we add a side condition to the instantiation: the substitution $σ$ must
+be {\em compatible} with the type $τ$, written $σ\sim τ$. To understand what
+compatibility means, consider an {\em atomic} region substitution (whose
+domain contains only one variable) $[ρ|->ρ']$. It is compatible with any type
+which does not contain $ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2
+e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
+Expressed more generally, an atomic substitution is compatible with a type if
+the type does not contain any region or effect variables which are in the
+range of the substitution. Finally, a {\em composite} substitution (whose
+domain contains more than one variable) is compatible with a type if each
+atomic substitution is compatible with the result of the application of the
+previous substitution. In this way, it becomes impossible to have different
+names for a single reference cell.
+
+The second form of aliasing is when we use the same name for {\em different}
+memory locations. This can happen when we allow the user to call
+$\mathit{ref}$ several times with the same region instantiation. A simple
+possibility is to  suppress the $\mathit{ref}$ function entirely and to
+replace it by a \letml\refml\ construct which combines name creation and
+memory allocation. In this way, there can of course be only one reference per
+name. On the other hand, because of the side condition $ρ\notin τ$, the
+reference cannot escape the scope of its \letml\refml\ construct, and so, a
+function cannot return a reference which has been created in its body. This
+restriction could be leveraged a bit if one separates name creation and
+allocation, while still maintaining that one can only allocate a single
+reference in each region. This is how it is done in the correctness proof of
+this restriction. The presented \letml\refml\ construct is of course a special
+case.
+
+In alias-free programs, the operations $\oplus$ and $|_ε$ on stores we
+previously introduced can be executed {\em statically}, i.e. just by looking
+at the domains of the stores. Also, we statically know which value (or which
+named component) the accessor {\tt !!} will get from a store.
+
+With the aliasing restriction, complex programs using shared mutable data
+structures are impossible to prove. Programs with references in lists,
+references in references and so on can be written, but only with the severe
+limitation that a list can only contain a single reference (potentially
+several times) and that a reference can only contain always the same
+(statically known) reference. However, references still are first-class
+values, i.e. they can be passed to functions, returned by functions and stored
+in data structures, with the above limitations.
+
+\paragraph{Correctness}
+
+In the 
 %TODO lemmas about correctness

### 2009-10-07 papers/esop2010/main.tex
go
section headings

@@
 \usepackage{mathpartir}
+\usepackage{url}
+
+\newtheorem{thm}{Theorem}[section]
+\newtheorem{lem}[thm]{Lemma}
 
@@
 %<-> domain, region <-> key
-
-%external TODO : proofs of system
-%
 
@@
 
-\paragraph{Correctness}
-
-In the 
-%TODO lemmas about correctness
-\section{Examples}
-\label{sec:examples}
-
-In this section, we will show two simple examples which use higher-order
-features.
-
-\subsection{A simple memoization function}
+\subsection{Correctness}
+
+We have proved the correctness of our system. All the properties are proved in
+subject reduction style, maintaining an invariant during the execution of the
+program, modeled with small-step semantics. We only consider well-typed terms.
+
+For the \wpre-calculus, we want to obtain that if we can prove $\wpre_s(t,q)$
+where $s$ is the current state of the store, $t$ a program, and $q$ the
+postcondition, and if $t$ reduces to the (syntactic) value $v$ and the store
+$s'$, then $q~s'~v$ is valid. The subject reduction way of proving this
+proceeds by proving that the validity of the formula $\wpre_s(t,q)$ is in fact
+maintained during the execution of the program. However, in the presence of
+effect typing, the effect of $t$ may {\em decrease} during program execution.
+Thus, the main lemma for the \wpre-calculus is 
+\begin{lemma} 
+  $\wpre_s(t,q)$ and $s,t ->> s',t'$ implies $\wpre_{s'}( (t : ε'<:ε), q)$.
+\end{lemma}
+Here, $->>$ means ``reduces in zero or more steps to''. The main theorem
+is then an easy consequence.
+
+For the alias restriction, or more precisely the restriction on reference
+creation, we formally prove that each region contains exactly one memory
+location, i.e. is actually a {\em singleton region}. Here we have to
+instrument the semantics and mark each reference cell in the store with the
+region it belongs to; therefore the exact statement is the following:
+\begin{theorem}
+  If $s,t ->>s',t$ and if $s$ contains only singleton regions (each reference
+  cell maps to a different region), then this is also the case for $s'$.
+\end{theorem}
+
+Both theorems are proved in the technical appendix~\cite{appendixesop2010}
+accompanying this paper. The aliasing restriction is proved there in a more
+general setting, while the correctness of the \wpre-calculus is proved in a
+slightly different version.
 
@@
 \end{figure}
+
+\section{Examples}
+\label{sec:examples}
+
+In this section, we will show two simple examples which use higher-order
+features.
+
+\subsection{A simple memoization function}
 

### 2009-10-07 papers/esop2010/main.tex
go

@@
 }
-%TODO explain somewhere the correspondence store type <-> map, effect variable
-%<-> domain, region <-> key
-
 \maketitle
@@
 
-\section{The general picture and a motivating example}
+\section{The General Picture and a Motivating Example}
 \label{sec:genpic}
@@
 
-\section{A programming language for specification}
+\section{A Programming Language for Specification}
 \label{sec:syntax}
 
-%TODO  notion of substitution - correspondence χ κ
 \subsection{Syntax}
@@
 
-\subsection{Alias restriction}
+\subsection{Alias Restriction}
 
@@
 
+\subsection{A Simple Memoization Function}
+
 \begin{figure}[tpb]
@@
 
-\subsection{A more powerful memoization function}
+\subsection{A More Powerful Memoization Function}
 
@@
 
-\subsection{All examples - an overview}
-
-A prototype implementation of the presented system, called ``\who'',
- is freely available~\cite{KanigFilliatre09wml}%TODO \footnote
-. It implements the
-described language with type, region and effect polymorphism, higher-order
-logic and store types in the logic. It takes as an input a fully annotated
-program and outputs proof obligations which have to be proved to guarantee the
-correctness of the program with respect to its specification. Currently, only
-output to the interactive proof assistant Coq has been implemented. Our logic
-is a subset of Coq's, so the translation is particularly simple.
-The implementation consists of less than 3000 lines of \ocaml\ code, including
-a limited form of type and effect inference.
+\subsection{All Examples - An Overview}
+
+A prototype implementation of the presented system, called ``\who'', is freely
+available~\cite{KanigFilliatre09wml}\footnote{It is available at
+\url{http://www.lri.fr/~kanig/who.html}.}. It implements the described
+language with type, region and effect polymorphism, higher-order logic and
+store types in the logic. It takes as an input a fully annotated program and
+outputs proof obligations which have to be proved to guarantee the correctness
+of the program with respect to its specification. Currently, only output to
+the interactive proof assistant Coq has been implemented. Our logic is a
+subset of Coq's, so the translation is particularly simple.  The
+implementation consists of less than 3000 lines of \ocaml\ code, including a
+limited form of type and effect inference.
 

### 2009-10-08 papers/esop2010/main.tex
go

@@
 function's specification depends on its arguments. In a higher-order setting,
-arguments may be functions that have their own specification. In consequence,
+arguments may be functions that have their own specification. As a consequence,
 to be modular, the specification of a higher-order function can only be
@@
 programming language and the weakest precondition calculus.  In
-section~\ref{sec:examples} we present our implementation and show some
+section~\ref{sec:examples} we present our implementation and some
 examples. Finally, section~\ref{sec:related} discusses related work.
@@
 precondition}, $q$ to the {\em postcondition} of the function. The analogy to
-Hoare triples is of course obvious. As usual, there are also variables,
+Hoare triples is obvious. As usual, there are also variables,
 \letml-bindings and an \ifml-\thenml-\elseml-construct. Variables on the type
@@
 
-In alias-free programs, the operations $\oplus$ and $|_ε$ on stores we
-previously introduced can be executed {\em statically}, i.e. just by looking
-at the domains of the stores. Also, we statically know which value (or which
-named component) the accessor {\tt !!} will get from a store.
+In alias-free programs, the operations $\oplus$ and $|_ε$ on stores can be
+executed {\em statically}, i.e. just by looking at the domains of the stores.
+Also, we statically know which value (or which named component) the accessor
+{\tt !!} will get from a store.
 
@@
 structures are impossible to prove. Programs with references in lists,
-references in references and so on can be written, but only with the severe
+references in references, etc., can be written, but only with the severe
 limitation that a list can only contain a single reference (potentially
@@
 
-Let's move on to the specification of {\tt ymemo}. We first define the {\tt
+Let us move on to the specification of {\tt ymemo}. We first define the {\tt
 realizes} predicate between a pure function {\tt f0} and an impure one

### 2009-10-08 papers/esop2010/main.tex
version submitted 15h

@@
 
-\begin{abstract} We present a specification language suited for reasoning
+\begin{abstract} We introduce a specification language suited for reasoning
   about effectful higher-order programs. The language's main features are
@@
   system. A prototype implementation exists and has been used to prove the
-  correctness of these and more examples.  \end{abstract}
+  correctness of these and more examples.  
+\end{abstract}
+
+%TODO overfull boxes
 
@@
 Sections~\ref{sec:syntax} and \ref{sec:wp} give the details about the
-programming language and the weakest precondition calculus.  In
-section~\ref{sec:examples} we present our implementation and some
-examples. Finally, section~\ref{sec:related} discusses related work.
+programming language and the weakest precondition calculus.
+Section~\ref{sec:alias} shows how two modifications to the type system can
+eliminate aliasing. In section~\ref{sec:examples} we present our
+implementation and some examples. Finally, section~\ref{sec:related} discusses
+related work.
 
@@
 of our prototype \who. There are a few syntactic differences with \ocaml, in
-particular for arrays.
+particular for arrays: the type of arrays is {\tt 'a farray ref}, the length
+of the array {\tt a} is {\tt length !a} in the program and {\tt length !!a} in
+the logic, and similarily access to the {\tt i}th element is {\tt get i !a} in
+the program and {\tt get i !!a} in the logic.
 
@@
 
-To avoid this situation, we additionally introduce a simple means to exclude
-programs with aliasing. The instantiation of effect variables is restricted to
-the ones which do not create aliasing situations. Our system, with aliasing
-exclusion enabled, is inappropriate for reasoning about shared mutable
-structures. However, it is interesting to see that this alias exclusion
-restriction, in many cases, corresponds to a natural condition imposed on
-higher-order functions that operate on some data structure. It is a common
+That is why we also propose in this paper a simple technique to exclude
+programs with aliasing and to obtain simpler proof obligations. The solution
+consists in restricting the instantiation of effect variables. In the example
+of Fig.~\ref{fig:iter}, this means that {\tt e} could not be instantiated with
+an effect containing {\tt a}.  With aliasing exclusion enabled, our system is
+of course inappropriate for reasoning about shared mutable structures.
+However, it is interesting to see that this alias exclusion restriction, in
+many cases, corresponds to a natural condition imposed on higher-order
+functions that operate on some data structure. It is a common
 requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration functions,
@@
 iterated on. In our system, this fact follows from the restriction on effect
-instantiations, as will be seen in section~\ref{sec:syntax}. In the example,
-the array {\tt a} of the iterating function {\tt iter} must not be modified by
-its argument.
+instantiations, as will be seen in section~\ref{sec:alias}.
+
+It is important to note that neither our specification language nor its
+weakest precondition calculus are restricted to alias-free programs. The
+aliasing exclusion is only a means to get simpler proof obligations whenever
+possible.
 
@@
 
-Finally, we introduce three symbols
+The store types $\efft{ε}$ describe a portion of the store and are basically
+mappings from region variables to values. The domain of the mapping is the
+effect $ε$, where effect variables stand for unknown parts of the domain. To
+dea with objects of these types, we introduce three symbols
 \begin{eqnarray*}
@@
 
-\subsection{Alias Restriction}
+We do not give a correctness proof for this calculus; this proof would be very
+similar to and actually simpler than the proof given in the next section.
+
+\section{The Alias Restriction}
+\label{sec:alias}
 
@@
 Also, we statically know which value (or which named component) the accessor
-{\tt !!} will get from a store.
+{\tt !!} will get from a store. Therefore, reasoning about stores and
+equalities between regions can be completely eliminated.
 
@@
 
-\subsection{Correctness}
+\paragraph{Correctness.}
 
@@
 
-\subsection{A Simple Memoization Function}
-
 \begin{figure}[tpb]
@@
 
-\subsection{A simple memoization function}
+\subsection{A Simple Memoization Function}
 
@@
 
-\subsection{A More Powerful Memoization Function}
+\subsection{Memoization of Recursive Functions}
 
@@
 \end{who}
-\caption{The {\tt ymemo function} and its specification}
+\caption{The {\tt ymemo} function and its specification}
   \label{fig:ymemo}
@@
 
-Our second, more complicated example is the recursive version of {\tt memo},
-dubbed {\tt ymemo}. We want to be able to define a recursive function whose
-recursive calls are automatically memoized. The basic idea of {\tt ymemo} is
-analogous to the fixpoint combinator in λ-calculus. As an example, here is how
-one would implement a memoized Fibonacci function
-using ymemo (ignoring the logic argument $f0$):
+Our second, more complicated example is a version of {\tt memo} for recursive
+functions, dubbed {\tt ymemo}. We want to be able to define a recursive
+function whose recursive calls are automatically memoized. The basic idea of
+{\tt ymemo} is analogous to the fixpoint combinator in λ-calculus, hence the
+name. As an example, here is how one would implement a memoized Fibonacci
+function using {\tt ymemo}:
 \begin{ocaml}
@@
 \end{ocaml}
-The anonymous function uses its first argument {\tt f} to do recursive calls.
+The anonymous function uses its first argument {\tt g} to do recursive calls.
 
@@
 body (see Fig~\ref{fig:ymemo}). This function, when called with an integer
-argument {\tt x} first checks if the table contains {\tt x} and if not, calls
+argument {\tt x} first checks if the table contains {\tt x}, and if not, calls
 {\tt ff} {\em with {\tt f} itself} and x as argument. As before, it stores the
@@
 language with type, region and effect polymorphism, higher-order logic and
-store types in the logic. It takes as an input a fully annotated program and
-outputs proof obligations which have to be proved to guarantee the correctness
-of the program with respect to its specification. Currently, only output to
-the interactive proof assistant Coq has been implemented. Our logic is a
-subset of Coq's, so the translation is particularly simple.  The
+store types in the logic. It is currently restricted to the subset
+characterized by the alias restriction. It takes as an input a fully annotated
+program and outputs proof obligations which have to be proved to guarantee the
+correctness of the program with respect to its specification. Currently, only
+output to the interactive proof assistant Coq has been implemented. Our logic
+is a subset of Coq's, so the translation is particularly simple.  The
 implementation consists of less than 3000 lines of \ocaml\ code, including a
@@
 is syntax-directed and thus entirely automated. Their logic of proof
-obligations is non-standard, while we obtain obligations in standard higher-order
-logic.
-
-They deal with aliasing using equalities and disequalities between
-pointers, which we chose not to do.
-
-They can reason about the classic {\em
-Landin's knot}, which consists in obtaining recursion by storing a function in
-a reference which calls this same reference. Our typing restriction prohibits
-this particular kind of programs.  Honda, Yoshida and Berger prove a
-completeness result for their system. We have proved no such result.
-
-%, using a slightly more
-%expressive language than ours. However, their calculus can
-%deal easily with functions in references, even if the stored function
-%reads or modifies the reference it is stored in. While, with the
-%appropriate extensions, one can reason about functions in references
-%in our system, this particular kind of programs cannot be reasoned
-%about.  Their extended
-%system~\cite{BergerHondaYoshida05aliasing} has the same advantages and
-%drawbacks compared to ours, but additionally deals with aliasing
-%situations.
+obligations is non-standard, while we obtain obligations in standard
+higher-order logic. They deal with aliasing using equalities and disequalities
+between pointers.  They can reason about the classic {\em Landin's knot},
+which consists in obtaining recursion by storing a function in a reference
+which calls this same reference. Our typing restriction prohibits this
+particular kind of programs.  Honda, Yoshida and Berger prove a completeness
+result for their system. We have proved no such result.
 
@@
 in adopting such an effect system, but haven't worked out the details. A
-second possibility is to deal with aliasing via equalities and disequalities
-over pointers, using regions to obtain some separation automatically. On
-itself, we believe that this technique does not scale, due to the explosion of
-needed disequalities.  However, in combination with other separation
+second possibility is to use techniques of static analysis to infer disjoint
+regions, each possibly containing several pointers~\cite{hubert07hav}.
+This approach has proved useful in combination with other separation
 techniques, for example the Burstall-Bornat memory model~\cite{bornat00mpc}
-which separates automatically objects of different structural types, this
-approach has worked well applied to first-order programs.
+which separates automatically objects of different structural types.
 

### 2009-10-08 papers/esop2010/main.tex
go

@@
 \end{abstract}
-
-%TODO overfull boxes
 
@@
 The store types $\efft{ε}$ describe a portion of the store and are basically
-mappings from region variables to values. The domain of the mapping is the
+mappings from references to values. The domain of the mapping is the
 effect $ε$, where effect variables stand for unknown parts of the domain. To
-dea with objects of these types, we introduce three symbols
+deal with objects of these types, we introduce three symbols
 \begin{eqnarray*}

### 2009-10-09 papers/esop2010/main.tex
go

--- 2009-10-08 papers/esop2010/main.tex
@@
 exactly what we want here. Using effect polymorphism, we can say that {\tt f}
-(the argument of {\tt iter}) has type $α->^e\unitml$, whera $α$ is the type of
+(the argument of {\tt iter}) has type $α->^e\unitml$, where $α$ is the type of
 the array elements and where $e$ is an {\em effect variable}, which amounts to
@@
 of the array {\tt a} is {\tt length !a} in the program and {\tt length !!a} in
-the logic, and similarily access to the {\tt i}th element is {\tt get i !a} in
+the logic, and similarly access to the {\tt i}th element is {\tt get i !a} in
 the program and {\tt get i !!a} in the logic.
@@
 The specification of {\tt iter} is correct, but it will be difficult to prove
-in general, because it is not clear wether {\tt f} modifies the array or not,
+in general, because it is not clear whether {\tt f} modifies the array or not,
 or more precisely, if the effect variable {\tt e} does contain {\tt a} or not.
@@
 effect $ε$, where effect variables stand for unknown parts of the domain. To
-dea with objects of these types, we introduce three symbols
+deal with objects of these types, we introduce three symbols
 \begin{eqnarray*}
@@
 the table should only contain bindings $y |-> {\mathtt f}~y$, which is
-expressend by the {\tt stores} predicate. The postcondition states that the
+expressed by the {\tt stores} predicate. The postcondition states that the
 result is equal to {\tt f0~x} and that the table still verifies the {\tt

### 2009-10-09 papers/esop2010/main.tex
go

@@
 This simple program can be annotated in a standard Hoare logic (for
-example the system of \cite{Filliatre00a}) as follows:
+example the Why system~\cite{Filliatre00a}) as follows:
 \begin{who}
@@
 The main idea of this paper is to combine the effect-based Hoare logic of
-\cite{Filliatre00a} with the more powerful effect system of
-\cite{talpin94taed} and the more powerful annotations of
-\cite{regis-gianas-pottier-08} to obtain a modular system for proving
-higher-order programs with side effects. As in~\cite{regis-gianas-pottier-08},
-we use operators \preml\ and \postml\ to access the pre- and postcondition of
-a function. However, in our system, a precondition takes as an argument not
-only the function argument, but also the initial state, before the execution
-of the function. The postcondition takes the argument and return value, but
-also the states before {\em and} after the function call. Two-state
-postconditions are a standard way to get rid of most auxiliary variables.
-However, the pre- and postcondition of {\tt f} may not refer to arbitrary
-parts of the state, but only the part {\tt f} reads or modifies.  With these
-conventions, one can now write a generic implementation and specification for
-the {\tt iter} function. It is given~Fig.~\ref{fig:iter}, in the input syntax
-of our prototype \who. There are a few syntactic differences with \ocaml, in
-particular for arrays: the type of arrays is {\tt 'a farray ref}, the length
-of the array {\tt a} is {\tt length !a} in the program and {\tt length !!a} in
-the logic, and similarly access to the {\tt i}th element is {\tt get i !a} in
-the program and {\tt get i !!a} in the logic.
+Why~\cite{Filliatre00a} with the more powerful effect system of Talpin and
+Jouvelot~\cite{talpin94taed} and the more powerful annotations of
+Régis-Gianas/Pottier~\cite{regis-gianas-pottier-08} to obtain a modular
+system for proving higher-order programs with side effects. As
+in~\cite{regis-gianas-pottier-08}, we use operators \preml\ and \postml\ to
+access the pre- and postcondition of a function. However, in our system, a
+precondition takes as an argument not only the function argument, but also the
+initial state, before the execution of the function. The postcondition takes
+the argument and return value, but also the states before {\em and} after the
+function call. Two-state postconditions are a standard way to get rid of most
+auxiliary variables.  However, the pre- and postcondition of {\tt f} may not
+refer to arbitrary parts of the state, but only the part {\tt f} reads or
+modifies.  With these conventions, one can now write a generic implementation
+and specification for the {\tt iter} function. It is
+given~Fig.~\ref{fig:iter}, in the input syntax of our prototype \who. There
+are a few syntactic differences with \ocaml, in particular for arrays: the
+type of arrays is {\tt 'a farray ref}, the length of the array {\tt a} is {\tt
+length !a} in the program and {\tt length !!a} in the logic, and similarly
+access to the {\tt i}th element is {\tt get i !a} in the program and {\tt get
+i !!a} in the logic.
 
@@
 The store types $\efft{ε}$ describe a portion of the store and are basically
-mappings from region variables to values. The domain of the mapping is the
+mappings from references to values. The domain of the mapping is the
 effect $ε$, where effect variables stand for unknown parts of the domain. To

### 2009-11-22 papers/esop2010/rebuttal.txt
rebuttal for ESOP 2010

Reviewer 1
==========

Lemma 1 of section 5 states the soundness theorem of the WP calculus. The
soundness of the type system itself is included in the precise formulation of
Theorem 1 (in the technical appendix). Despite the superficial differences,
the soundness theorem (and proof) is very similar to the classical
presentation of effect systems, so we judged it to be of less importance.

Concerning the related works, this paper aims at achieving (almost) as much as
others (such as YNot) without using dependent types or separation logic. Our
aim, stated in the section of future works, is to be able to discharge the
majority of the proof obligations by  off-the-shelve first-order provers. This
would bring a higher degree of automation than existing tools, but we are not
there yet.

We do not share the reviewers' aversion against regions. Regions are employed
in many recent papers (Charguéraud/Pottier ICFP 2008), and once an effect
system is used, they are a very simple way to improve the precision of the
system (compared to using program names).

We don't understand the "suspicion" expressed by the reviewer. We believe that
there are basically the following possibilities to deal with aliasing:
  1) separation logic (the systems cited by the reviewer use it)
  2) a very powerful effect system
     (again Charguéraud/Pottier ICFP 2008, for example)
  3) separation analyses (see for example Hubert/Marché [11] and Bornat [4])
  4) pointer (dis-)equalities
We do not want to use separation logic, and we need time to integrate ideas
from the domains 2) and 3) (stated in the future works). The remaining
solution is 4), which is possible in our system, but we consider it to be not
very practical for the difficult case of shared mutable data structures. We
therefore decided to deal with alias-free programs first. But there is no
inherent incapacity of effect systems to deal with aliasing programs.

We think that the paragraph "It would have been a somewhat different
matter ..." does not do justice to our paper. Sections 3 and 4 present a type
system and a wp calculus which *allow* programs with aliasing. The current
prototype implementation treats alias-free programs by default, but to enable
aliasing there is just a simple check in the type system to be removed
(Section 5) and the simplifications enabled by the absence of aliasing
omitted.

Reviewer 2
==========

Yes, distinguishing read/write effects would be a very valuable extension, as
it greatly improve the practical usage of such a system. We are aware of the
importance of such an extension, but haven't worked out the details.

p12 - "The second form of aliasing" - No, this is not about alpha-conversion.
In the usual region calculus and in [24] (Talpin/Jouvelot), a region may
contain more than one reference. You seem to understand the notion of region
in the same way, see your comment about "effects" and "regions" before.
Different references which lay in the same region are aliased in this way -
they cannot be distinguished by the type system. The second restriction of
section 5 disallows allocating twice in the same region, thus eliminating this
form of aliasing. Maybe calling this aliasing is confusing because this notion
usually refers only to the first form in the literature. However, both are
really dual.

Our contribution with respect to Why / Spec# is of cours the support for
higher-order programs; both tools only address first-order programs.

Reviewer 3
==========

Detail a bit more what our concrete contributions are

Concerning aliasing: While reasoning about programs containing aliasing is
possible in our system, we do not claim to present a practical solution for
this problem in this paper. Our system is practical for alias-free programs
and we believe that the examples show this. However, just as separating
analyses other than separation logic are possible in a first-order setting, we
believe this to be true for higher-order programs. These analyses could be
done on top of our system. However, we don't have any experimental data to
support this claim.

Yes, our ML Workshop 2009 paper is mainly about an extended case study using
our tool Who. The citation always appears in connection with the tool our the
case study.

Reviewer 4
===========

Detail a bit more what our concrete contributions are

### 2010-02-18 doc/manual/manual.tex
a beginning of a manual

\documentclass[a4paper]{scrreprt}

\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[usenames,dvipsnames]{xcolor}

\newcommand{\cat}[1]{$<$\textcolor{Aquamarine}{#1}$>$}

\begin{document}

\chapter{Concrete Syntax}

In this section, we describe formally the syntax of the input language.

We describe the input language by a grammar, separated in several categories.
By necessity, we have to mix concrete and abstract syntax. We use the syntax
of the form \cat{Category} to refer to categories. Every other word or symbol
is part of the concrete syntax, with the exception of the following symbols:
\texttt{( ) + ? * |}. These symbols have the sense they usually have when
talking about grammars and regular expressions: parentheses group expressions,
{\tt *} stands for zero or more occurrences, {\tt +} for at least one
occurrence, and {\tt ?} for an optional occurrence. Finally, {\tt |}
represents a choice.  When we want to refer to the actual symbol, we write
{\tt '('}, {\tt ')'} and so on.

We write {\tt list*(expr,symbol)} or {\tt list+(expr,symbol)} to describe
lists, possibly empty or containing at least one element, of expressions
separated by a symbol. Finally, we abbreviate {\tt list+(expr, ',')} by {\tt
commasep+(expr)}.

An \cat{Id} is program identifier or a global type and is defined as: a word
composed of {\tt [A-Za-z0-9']*}, which starts with a letter out ot {\tt
[A-Za-z]}.

A \cat{Region} is the same category.

An \cat{EffectVar} or \cat{Tyvar} is defined as the same category, 
but starting with a {\tt '}.

A file is a list of \cat{Declaration}s.

A \cat{Declaration} is one of the following: 
\begin{itemize}
  \item {\tt axiom} \cat{Id} \cat{Gen} {\tt :} \cat{Term}
  \item {\tt logic} \cat{Id} \cat{Gen} {\tt :} \cat{Type}
  \item {\tt type} \cat{Id} \cat{Gen}  ( = \cat{Type}){\tt ?}
  \item {\tt letregion} \cat{Id}
  \item {\tt parameter} \cat{Id} \cat{Gen} \cat{Argslist} {\tt :} \cat{Type}
  \item {\tt parameter} \cat{Id} \cat{Gen} \cat{Argslist} {\tt :} \cat{Type}
    {\tt ,} \cat{Effect} {\tt =} \cat{Allocdef} {\cat{Pre}} {\cat{Post}}
  \item \cat{Letform}
\end{itemize}

A \cat{Letform} is:
\begin{itemize}
  \item {\tt let} \cat{Id} \cat{Gen} \cat{Argslist} {\tt =} \cat{Term}
  \item {\tt let} \cat{Id} \cat{Gen} \cat{Argslist} {\tt =} \cat{Allocdef} {\cat{Pre}} \cat{Term} {\cat{Post}}
  \item {\tt let rec} \cat{Gen} '{\tt (}' \cat{Id} {\tt :} \cat{ty} '{\tt )}'
    \cat{Argslist} {\tt =} \cat{Allocdef} {\cat{Pre}} \cat{Term} {\cat{Post}}
\end{itemize}

A generalisation \cat{Gen} is:
\begin{itemize}
  \item {\tt [} \cat{Tyvar}* '{\tt |}' \cat{Region} '{\tt |}' \cat{EffectVar}*
    {\tt ]}
\end{itemize}

An argument list \cat{Argslist} is:
\begin{itemize}
  \item ( '{\tt (}' \cat{Id} : \cat{Type} '{\tt )}' )*
\end{itemize}

An \cat{Allocdef} is:
\begin{itemize}
  \item ( allocates \cat{Region} ){\tt ?}
\end{itemize}

A precondition \cat{Pre} is: 
\begin{itemize}
  \item \cat{Term}{\tt ?}
\end{itemize}

A postcondition \cat{Post} is: 
\begin{itemize}
  \item \cat{Term}{\tt ?}? 
  \item (\cat{Id} {\tt :} )? \cat{Term}
\end{itemize}

A \cat{Type} is one of the following
\begin{itemize}
  \item {\tt int}
  \item {\tt prop}
  \item \cat{Tyvar}
  \item '{\tt (}' \cat{Type} '{\tt )}'
  \item \cat{Id} \cat{inst}?
  \item \cat{Type} \cat{Id}
  \item '{\tt (}' commasep(\cat{Type}) '{\tt )}' \cat{Id}
  \item \cat{Type} {\tt ->} \cat{Type}
  \item \cat{Type} '{\tt *}' \cat{Type}
  \item \cat{Type -}\cat{Effect} \cat{Type}
  \item '{\tt <}' \cat{Effect} '{\tt >}'
  \item {\tt [[} \cat{Type} {\tt ]]}
  \item {\tt ref} '{\tt (}'\cat{Region} \cat{Type}'{\tt )}'
\end{itemize}

An \cat{Effect} is: 
\begin{itemize}
  \item {\tt \{} ( \cat{Id} {\tt |} \cat{EffectVar} )* {\tt \}}
\end{itemize}

An instantiation \cat{inst} is:
\begin{itemize}
  \item {\tt [} commasep*(\cat{Type}) '{\tt |}' \cat{Region} '{\tt |}' \cat{Effect}* {\tt
    ]}
\end{itemize}

A \cat{Term} is one of the following:
\begin{itemize}
  \item '{\tt (}' '{\tt )}'
  \item {\tt !} \cat{Id}
  \item {\tt !!} \cat{Id} ( '{\tt |}' \cat{Term} )
  \item \cat{Id} \cat{inst}?
  \item '{\tt (}' \cat{Term} ( {\tt :} \cat{Type})? '{\tt )}'
  \item \cat{Term} ({\tt \{\{ } \cat{Region} {\tt \}\}} )? \cat{Term} 
  \item \cat{Term} {\tt ;} \cat{Term} 
  \item \cat{Term} \cat{Infix} \cat{Term}
  \item fun \cat{Argslist} {\tt ->} \cat{Term}
  \item fun \cat{Argslist} {\tt ->} \cat{Allocdef} {\cat{Pre}} \cat{Term} {\cat{Post}}
  \item ( {\tt forall} {\tt |} {\tt exists} ) \cat{Argslist} {\tt .} \cat{Term}
  \item {\tt letregion} \cat{Region} {\tt in} \cat{Term}
  \item {\tt if} \cat{Term} {\tt then} \cat{Term} {\tt else} \cat{Term}
  \item \cat{Letform} {\tt in} \cat{Term}
  \item {\tt for} \cat{Id} = \cat{Term} ( {\tt to | downto} ) \cat{Term} {\tt
    do} {\cat{Pre}} \cat{Term} {\tt done}
  \item {\tt [[} \cat{Pre} {\tt ]]} \cat{Term} {\tt [[} \cat{Post} {\tt ]]}
\end{itemize}

\end{document}

### 2010-02-18 doc/manual/manual.tex
links in manual

@@
 \usepackage[usenames,dvipsnames]{xcolor}
-\usepackage[usenames,dvipsnames]{xcolor}
+\usepackage{xcolor}
+\definecolor{dablue}{rgb}{0.1,0.1,0.4}
 
-\newcommand{\cat}[1]{$<$\textcolor{Aquamarine}{#1}$>$}
+\usepackage[colorlinks=true,pdfborder={0,0,0,0},urlcolor=dablue,
+linkcolor=dablue, pdftex]{hyperref}
+
+%\newcommand{\cat}[1]{$<$\textcolor{Aquamarine}{#1}$>$}
+\newcommand{\cat}[1]{$<$\hyperref[#1]{#1}$>$}
+\newcommand{\catdef}[1]{\label{#1}$<$\textcolor{Aquamarine}{#1}$>$}
 
@@
 
-An \cat{Id} is program identifier or a global type and is defined as: a word
+\paragraph{}
+An \catdef{Id} is program identifier or a global type and is defined as: a word
 composed of {\tt [A-Za-z0-9']*}, which starts with a letter out ot {\tt
@@
 
-A \cat{Region} is the same category.
+\paragraph{}
+A \catdef{Region} is the same category.
 
-An \cat{EffectVar} or \cat{Tyvar} is defined as the same category, 
+\paragraph{}
+An \catdef{EffectVar} or \catdef{Tyvar} is defined as the same category, 
 but starting with a {\tt '}.
@@
 
-A \cat{Declaration} is one of the following: 
+\paragraph{}
+A \catdef{Declaration} is one of the following: 
 \begin{itemize}
@@
 
-A \cat{Letform} is:
+\paragraph{}
+A \catdef{Letform} is:
 \begin{itemize}
@@
 
-A generalisation \cat{Gen} is:
+\paragraph{}
+A generalisation \catdef{Gen} is:
 \begin{itemize}
@@
 
-An argument list \cat{Argslist} is:
+\paragraph{}
+An argument list \catdef{Argslist} is:
 \begin{itemize}
@@
 
-An \cat{Allocdef} is:
+\paragraph{}
+An \catdef{Allocdef} is:
 \begin{itemize}
@@
 
-A precondition \cat{Pre} is: 
+\paragraph{}
+A precondition \catdef{Pre} is: 
 \begin{itemize}
@@
 
-A postcondition \cat{Post} is: 
+\paragraph{}
+A postcondition \catdef{Post} is: 
 \begin{itemize}
@@
 
-A \cat{Type} is one of the following
+\paragraph{}
+A \catdef{Type} is one of the following
 \begin{itemize}
@@
   \item '{\tt (}' \cat{Type} '{\tt )}'
-  \item \cat{Id} \cat{inst}?
+  \item \cat{Id} \cat{Inst}?
   \item \cat{Type} \cat{Id}
@@
 
-An \cat{Effect} is: 
+\paragraph{}
+An \catdef{Effect} is: 
 \begin{itemize}
@@
 
-An instantiation \cat{inst} is:
+\paragraph{}
+An instantiation \catdef{Inst} is:
 \begin{itemize}
@@
 
-A \cat{Term} is one of the following:
+\paragraph{}
+A \catdef{Term} is one of the following:
 \begin{itemize}
@@
   \item {\tt !!} \cat{Id} ( '{\tt |}' \cat{Term} )
-  \item \cat{Id} \cat{inst}?
+  \item \cat{Id} \cat{Inst}?
   \item '{\tt (}' \cat{Term} ( {\tt :} \cat{Type})? '{\tt )}'

### 2010-02-23 doc/manual/manual.tex
typo
fixed manual
manual - a first presentable version

@@
 \usepackage{xcolor}
+\usepackage{xspace}
 \definecolor{dablue}{rgb}{0.1,0.1,0.4}
@@
 \newcommand{\cat}[1]{$<$\hyperref[#1]{#1}$>$}
-\newcommand{\catdef}[1]{\label{#1}$<$\textcolor{Aquamarine}{#1}$>$}
+\newcommand{\likecat}[1]{$<$\textcolor{Aquamarine}{#1}$>$}
+\newcommand{\catdef}[1]{\label{#1}\likecat{#1}}
+\newcommand{\who}{Who\xspace}
+
+\usepackage{alltt}
 
@@
 
+\title{Who - The Manual}
+\author{Johannes Kanig}
+
+\maketitle
+\tableofcontents
+
+\chapter{A walkthrough using Arrays}
+
+In this chapter, we aim to describe almost all syntactic elements of the
+\who~language.
+
+\section{Types, Axioms and Pure Functions}
+
+A type definition of a polymorphic type looks as follows:
+\begin{alltt}
+  type array ['a||]
+\end{alltt}
+Notice the two bars at the end to indicate empty region and effect parameters.
+Also note that despite the name, {\tt array} is {\em not} a mutable type.
+
+We can declare functions over this type in the following way:
+\begin{alltt}
+  logic create ['a||] :  int -> 'a -> 'a array
+  logic get ['a||] : int -> 'a array -> 'a
+  logic set ['a||]: int -> 'a -> 'a array -> 'a array 
+  logic len ['a||] :  'a array -> int
+\end{alltt}
+Again, notice that these are pure functions which act on {\em functional}
+arrays, for example {\tt set} returns a new, modified array.
+
+Now, we can declare properties of these functions in form of axioms:
+\begin{alltt}
+  axiom update_length ['a||] : 
+    forall (t : 'a array) (i : int) (z : 'a).
+    0 <= i \verb|/\| i < len t -> len t = len (set i z t)
+  axiom get_set_neq ['a||] : 
+    forall (t : 'a array ) (i : int) (j : int) (z : 'a).
+      (0 <= i \verb|/\| i < len t) -> 
+      (0 <= j \verb|/\| j < len t) -> 
+        i <> j -> get i (set j z t) = get i t
+  axiom get_set_eq ['a||] : 
+    forall (t : 'a array) (i : int) (z : 'a).
+      0 <= i \verb|/\| i < len t -> get i (set i z t) = z
+\end{alltt}
+
+Now, we can also {\em define} functions over those functional arrays:
+\begin{alltt}
+  let zeroset ['a||] (v : 'a) (a : 'a array) = set 0 v a
+\end{alltt}
+Here, we have defined a {\em pure} function {\tt zeroset} of type {\tt ['a||].
+int -> 'a array -> 'a array}. No effects take place, and nothing has to be
+proved.
+
+\section{Functions with Annotations}
+
+If we want to define a {\em calculatory function} which does the same, here we
+go:
+\begin{alltt}
+  let zerosetc ['a||] (v : 'a) (a : 'a array) = 
+    \{ \}
+    set 0 v a
+    \{ \}
+\end{alltt}
+Notice the additional curly brackets at the beginning and the end of the
+function body. The function {\tt zerosetc} has the type {\tt ['a||]. int -> 'a
+array ->{} 'a array}. We see that the last (and only the last) arrow has an
+effect annotation. In this case, this effect annotation is empty; no effect
+takes place, because {\tt set} returns a new function.
+
+Now, if we want to state anything about the return value of {\tt zerosetc}, we
+need to write a {\em postcondition}. For example, we would like to state that
+the returned array contains zero at position $i$:
+\begin{alltt}
+  let zerosetc ['a||] (v : 'a) (a : 'a array) = 
+    \{ \}
+    set 0 v a
+    \{ r : get 0 r = v \}
+\end{alltt}
+We use the syntax {\tt r : } to speak about the return value; {\tt r} is a
+name chosen by the user. This definition produces a proof obligation. In this
+case, this proof obligation is actually unprovable, because the array {\tt a}
+might be empty, in which case the axioms do not say anything of {\tt set 0 v a}.
+We have to give a {\em precondition}:
+\begin{alltt}
+  let zerosetc ['a||] (v : 'a) (a : 'a array) = 
+    \{ 0 < len a \}
+    set 0 v a
+    \{ r : get 0 r = v \}
+\end{alltt}
+And now we can prove this function.
+
+\section{References}
+
+To write an effectful function, we need references. A reference type on an
+array is written as follows:
+\begin{alltt}
+  ref(r,'a array)
+\end{alltt}
+where {\tt r} is a region name. Regions, as types, can be concrete or
+parameters of a function. Let us start with the case where it is a parameter.
+We want to define the same function as before, but in an imperative way. Here
+we go:
+\begin{alltt}
+  let zerosetc ['a|r|] (v : 'a) (a : ref(r,'a array)) = 
+    \{ 0 < len !!a \}
+    a := set 0 v !a
+    \{ get 0 !!a = v \}
+\end{alltt}
+There are several remarks here. First, we have to introduce the region name
+used with the reference {\tt a}; this is occurrence of {\tt r} between {\tt |}
+in the brackets. In programs, references are treated with {\tt :=} and {\tt !}
+as usual (there is currently no special syntax for arrays). In annotations, to
+access the contents of a reference, we write {\tt !!} (this is subject to
+change). The return type of this function is {\tt unit} and the overall type
+is:
+\begin{alltt}
+  ['a|r|]. 'a -> ref(r,'a array) ->{r} unit
+\end{alltt}
+We see that this function has an effect on {\tt r}.
+
+Let us write a function which increments the contents of the first cell of an
+array.
+\begin{alltt}
+  let zeroincr [|r|] (a : ref(r,int array)) = 
+    \{ 0 < len !!a \}
+    a := set 0 (get 0 a +1) a
+    \{ get 0 !!a = get 0 !!a|old + 1 \}
+\end{alltt}
+Here, we have used the syntax {\tt !!a|old} to speak of the {\em old} value of
+{\tt a}, before the execution of the function {\tt zeroincr}.
+
+To create references, we need an additional mechanism.
+\begin{alltt}
+  let zero_array [|r|] (l : int) = allocates r
+    \{ 0 <= l \}
+    if l == 0 then ref\{\{r\}\} ar_empty
+    else ref\{\{r\}\} (create l 0)
+    \{ x : forall i. 0 <= i \verb|/\| i < l -> get i !!x = 0 \}
+\end{alltt}
+To be able to create a reference inside this function, we need an {\tt
+allocates} declaration. This enables us to use a {\em allocation capacity}
+when calling {\tt ref}, using the syntax {\tt ref\{\{r\}\}}. Also notice the
+use {\tt ==} instead of {\tt =} in the test, to obtain a boolean return value.
+
+\section{Effect polymorphism}
+The last missing bit is effect polymorphism. Suppose we want to implement {\tt
+Array.iter} in \who. This can be easily done, at first without logical
+annotations:
+\begin{alltt}
+let iter ['a|t|'e] (ar : ref(t, 'a array)) (f : 'a ->\{t 'e\} unit)  =
+  \{  \}
+  for i = 0 to len !ar - 1 do 
+    \{ \}
+    f (get i !ar)
+  done
+  \{ \}
+\end{alltt}
+We use the syntax for {\tt for} loops here; one has to give the loop
+invariant, but for the moment we leave that empty. The only other new aspect
+is the use of {\em effect polymorphism}, for which is reserved the third field
+of the generalisation declaration {\tt ['a|out|'e]}. We generalize over a
+certain effect {\tt 'e} and say that the function argument {\tt f} has
+precisely this effect {\tt 'e} (see the type of {\tt f}).
+
+The specification of {\tt iter} is relatively complex; the reason is that any
+iterative higher-order function is basically a fold, because of the unknown
+effect of the function in argument.
+
+The entire code of {\tt iter} looks like this:
+\begin{alltt}
+let iter ['a|t|'e] 
+  (inv : <t 'e> -> int ->  prop) 
+  (ar : ref(t, 'a array)) (f : 'a ->\{t 'e\} unit)  =
+  \{ inv cur 0 \verb|/\| 
+    forall (i:int). 0 <= i \verb|/\| i < len (!!ar) ->
+    [[ inv cur i ]] f (get i !ar) [[inv cur (i+1)]]
+  \}
+  for i = 0 to len !ar - 1 do 
+    \{ inv cur i \}
+    f (get i !ar)
+  done
+  \{inv cur (len !!ar|old) \}
+\end{alltt}
+There are three new elements in this code, as well as a trick to specify this
+function. First keep in mind that {\tt iter} is basically a {\tt for} loop,
+this means that we need an invariant to specify it. Now, the trick is to take
+this invariant as an additional argument of {\tt iter}, called {\tt inv}. The
+first new element is hidden in the type of {\tt inv}, namely the type {\tt
+<'e>} of its first argument. Types in angle brackets {\tt < >} represent parts
+of the {\em state}, here we say that the invariant may talk about the state
+represented by {\tt 'e}. Just as {\tt old} is a keyword for the iter state of
+the function call, available in the postcondition, {\tt cur} is a keyword for
+the current state, and it is available in any annotation.  Now the
+postcondition simply states that we obtain the invariant of the final state
+over the entire array. The invariant of the {\tt for} loop states that we keep
+the invariant during the execution. Finally, the precondition states that the
+invariant holds for the index zero, and that the invariant is preserved by the
+function argument. This is expressed by the syntax
+\begin{alltt}
+    [[ inv cur i ]] f (get i !ar) [[inv cur (i+1)]]
+\end{alltt}
+which is basically a Hoare triple.
+
+There are a few other syntactic points. First, Hoare Triples using the double
+bracket syntax are syntactic sugar for a more complex predicate. Assume that {\tt
+f} has the type {\tt t1 ->{e} t2} where {\tt e} is some effect, then the Hoare
+triple
+\begin{alltt}
+    [[ p cur x ]] f x [[r : q old cur x r]]
+\end{alltt}
+is a shortcut for
+\begin{alltt}
+  forall (m n : <e>) (r : t2). 
+    p cur x -> pre f x m /\
+    post f x m n r -> q m n x r
+\end{alltt}
+Here we have used the ability to quantify over the portion of a state
+described by an effect expression.
+
+Concerning state expressions, one can restrict a state expression {\tt s} to a
+smaller domain described by an effect using the syntax:
+\begin{alltt}
+s|\{e\}
+\end{alltt}
+
+\appendix
+
 \chapter{Concrete Syntax}
 
-In this section, we describe formally the syntax of the input language.
+In this appendix, we describe formally the syntax of the input language.
 
@@
 By necessity, we have to mix concrete and abstract syntax. We use the syntax
-of the form \cat{Category} to refer to categories. Every other word or symbol
+of the form \likecat{Category} to refer to categories. Every other word or symbol
 is part of the concrete syntax, with the exception of the following symbols:
@@
   \item {\tt type} \cat{Id} \cat{Gen}  ( = \cat{Type}){\tt ?}
-  \item {\tt letregion} \cat{Id}
+  \item {\tt letregion} \cat{Id}*
   \item {\tt parameter} \cat{Id} \cat{Gen} \cat{Argslist} {\tt :} \cat{Type}
@@
   \item {\tt let} \cat{Id} \cat{Gen} \cat{Argslist} {\tt =} \cat{Allocdef} {\cat{Pre}} \cat{Term} {\cat{Post}}
-  \item {\tt let rec} \cat{Gen} '{\tt (}' \cat{Id} {\tt :} \cat{ty} '{\tt )}'
+  \item {\tt let rec} \cat{Gen} '{\tt (}' \cat{Id} {\tt :} \cat{Type} '{\tt )}'
     \cat{Argslist} {\tt =} \cat{Allocdef} {\cat{Pre}} \cat{Term} {\cat{Post}}
@@
 \begin{itemize}
-  \item {\tt [} \cat{Tyvar}* '{\tt |}' \cat{Region} '{\tt |}' \cat{EffectVar}*
+  \item {\tt [} \cat{Tyvar}* '{\tt |}' \cat{Region}* '{\tt |}' \cat{EffectVar}*
     {\tt ]}
@@
 \begin{itemize}
-  \item ( '{\tt (}' \cat{Id} : \cat{Type} '{\tt )}' )*
+  \item ( '{\tt (}' \cat{Id} + : \cat{Type} '{\tt )}' )*
 \end{itemize}
@@
 \begin{itemize}
-  \item ( allocates \cat{Region} ){\tt ?}
+  \item ( allocates \cat{Region}* ){\tt ?}
 \end{itemize}
@@
   \item \cat{Type} '{\tt *}' \cat{Type}
-  \item \cat{Type -}\cat{Effect} \cat{Type}
-  \item '{\tt <}' \cat{Effect} '{\tt >}'
+  \item \cat{Type} {\tt ->} \cat{Effect} \cat{Type}
+  \item '{\tt <}' \cat{EffectWithoutBraces} '{\tt >}'
   \item {\tt [[} \cat{Type} {\tt ]]}
@@
 \paragraph{}
+An \catdef{EffectWithoutBraces} is: 
+\begin{itemize}
+  \item ( \cat{Id} {\tt |} \cat{EffectVar} )*
+\end{itemize}
+
+\paragraph{}
 An \catdef{Effect} is: 
 \begin{itemize}
-  \item {\tt \{} ( \cat{Id} {\tt |} \cat{EffectVar} )* {\tt \}}
+  \item {\tt \{} \cat{EffectWithoutBraces} {\tt \}}
 \end{itemize}
@@
 \begin{itemize}
-  \item {\tt [} commasep*(\cat{Type}) '{\tt |}' \cat{Region} '{\tt |}' \cat{Effect}* {\tt
+  \item {\tt [} commasep*(\cat{Type}) '{\tt |}' \cat{Region}* '{\tt |}' \cat{Effect}* {\tt
     ]}
@@
   \item '{\tt (}' \cat{Term} ( {\tt :} \cat{Type})? '{\tt )}'
-  \item \cat{Term} ({\tt \{\{ } \cat{Region} {\tt \}\}} )? \cat{Term} 
+  \item \cat{Term} ({\tt \{\{ } \cat{Region}* {\tt \}\}} )? \cat{Term} 
   \item \cat{Term} {\tt ;} \cat{Term} 
@@
   \item ( {\tt forall} {\tt |} {\tt exists} ) \cat{Argslist} {\tt .} \cat{Term}
-  \item {\tt letregion} \cat{Region} {\tt in} \cat{Term}
+  \item {\tt letregion} \cat{Region}* {\tt in} \cat{Term}
   \item {\tt if} \cat{Term} {\tt then} \cat{Term} {\tt else} \cat{Term}
@@
 
+\paragraph{}
+The following \catdef{Infix} symbols are predefined: 
+  \begin{itemize}
+  \item the comparison operators of return type prop: {\tt <}, {\tt >}, {\tt
+  <=}, {\tt =>}, {\tt =}, {\tt <>}
+  \item the comparison operators of return type bool: {\tt <<}, {\tt >>}, {\tt
+  <<=}, {\tt >>=}, {\tt ==}, {\tt !=}
+  \item the logical connectives: \verb|/\|, \verb|\/|, {\tt ->}
+  \item arithmetic operations: {\tt +}, {\tt -}, {\tt *}
+  \item tuple construction: {\tt ,}
+  \end{itemize}
+
 \end{document}

### 2010-04-03 doc/manual/manual.tex
an option for versions
version 0.3

@@
 
+\chapter{The command line tool}
+
 \chapter{A walkthrough using Arrays}
@@
 \begin{alltt}
-  type array ['a||]
+  type array ['a]
 \end{alltt}
-Notice the two bars at the end to indicate empty region and effect parameters.
 Also note that despite the name, {\tt array} is {\em not} a mutable type.
@@
 \begin{alltt}
-  logic create ['a||] :  int -> 'a -> 'a array
-  logic get ['a||] : int -> 'a array -> 'a
-  logic set ['a||]: int -> 'a -> 'a array -> 'a array 
-  logic len ['a||] :  'a array -> int
+  logic create ['a] :  int -> 'a -> 'a array
+  logic get ['a] : int -> 'a array -> 'a
+  logic set ['a]: int -> 'a -> 'a array -> 'a array
+  logic len ['a] :  'a array -> int
 \end{alltt}
@@
 \begin{alltt}
-  axiom update_length ['a||] : 
+  axiom update_length ['a] :
     forall (t : 'a array) (i : int) (z : 'a).
     0 <= i \verb|/\| i < len t -> len t = len (set i z t)
-  axiom get_set_neq ['a||] : 
+  axiom get_set_neq ['a] :
     forall (t : 'a array ) (i : int) (j : int) (z : 'a).
@@
         i <> j -> get i (set j z t) = get i t
-  axiom get_set_eq ['a||] : 
+  axiom get_set_eq ['a] :
     forall (t : 'a array) (i : int) (z : 'a).
@@
 \begin{alltt}
-  let zeroset ['a||] (v : 'a) (a : 'a array) = set 0 v a
+  let zeroset ['a] (v : 'a) (a : 'a array) = set 0 v a
 \end{alltt}
-Here, we have defined a {\em pure} function {\tt zeroset} of type {\tt ['a||].
+Here, we have defined a {\em pure} function {\tt zeroset} of type {\tt ['a].
 int -> 'a array -> 'a array}. No effects take place, and nothing has to be
@@
 Notice the additional curly brackets at the beginning and the end of the
-function body. The function {\tt zerosetc} has the type {\tt ['a||]. int -> 'a
+function body. The function {\tt zerosetc} has the type {\tt ['a]. int -> 'a
 array ->{} 'a array}. We see that the last (and only the last) arrow has an
@@
 \begin{alltt}
-  let zerosetc ['a||] (v : 'a) (a : 'a array) = 
+  let zerosetc ['a] (v : 'a) (a : 'a array) =
     \{ \}
@@
 \begin{alltt}
-  let zerosetc ['a||] (v : 'a) (a : 'a array) = 
+  let zerosetc ['a] (v : 'a) (a : 'a array) =
     \{ 0 < len a \}
@@
 \end{alltt}
+
+%TODO read/write effects
 

### 2010-04-12 whytools/proveit.sh
prove it

@@ -1,0 +1,19 @@
+#!/bin/bash
+
+set -x
+TIMEOUT=5
+while getopts 't:' OPTION
+do
+  case $OPTION in
+    t) TIMEOUT=$OPTARG ;;
+    ?) exit 2;;
+  esac
+done
+
+CAT=`date +%F+%X`
+proofmgr -add-bench -c $CAT $*
+proofmgr -run -i 1 -c $CAT -cores 5 -timeout $TIMEOUT
+proofmgr -run -i 2 -c $CAT -cores 2 -timeout $TIMEOUT
+proofmgr -run -i 3 -c $CAT -cores 3 -timeout $TIMEOUT
+proofmgr -run -i 4 -c $CAT -cores 2 -timeout $TIMEOUT
+proofmgr -run -i 5 -c $CAT -cores 5 -timeout $TIMEOUT

### 2010-04-13 whytools/proveit.sh
test commit

@@ -17,3 +17,4 @@ proofmgr -run -i 2 -c $CAT -cores 2 -timeout $TIMEOUT
 proofmgr -run -i 3 -c $CAT -cores 3 -timeout $TIMEOUT
 proofmgr -run -i 4 -c $CAT -cores 2 -timeout $TIMEOUT
 proofmgr -run -i 5 -c $CAT -cores 5 -timeout $TIMEOUT
+

### 2010-04-13 whytools/proveit.sh
find list of provers automatically

@@ -2,19 +2,20 @@
 
 set -x
 TIMEOUT=5
-while getopts 't:' OPTION
+CORES=5
+while getopts 't:c:' OPTION
 do
   case $OPTION in
     t) TIMEOUT=$OPTARG ;;
+    c) CORES=$OPTARG ;;
     ?) exit 2;;
   esac
 done
 
 CAT=`date +%F+%X`
 proofmgr -add-bench -c $CAT $*
-proofmgr -run -i 1 -c $CAT -cores 5 -timeout $TIMEOUT
-proofmgr -run -i 2 -c $CAT -cores 2 -timeout $TIMEOUT
-proofmgr -run -i 3 -c $CAT -cores 3 -timeout $TIMEOUT
-proofmgr -run -i 4 -c $CAT -cores 2 -timeout $TIMEOUT
-proofmgr -run -i 5 -c $CAT -cores 5 -timeout $TIMEOUT
+PROVERS=`proofmgr -show-provers | grep "\[.*\]" | sed -e "s/.*\[.\(.*\)\].*/\1/"`
+for i in $PROVERS ; do
+  proofmgr -run -i $i -c $CAT -cores $CORES -timeout $TIMEOUT
+done
 

### 2010-04-14 whytools/proveit.sh
more functions for arrays

@@ -16,6 +16,6 @@ CAT=`date +%F+%X`
 proofmgr -add-bench -c $CAT $*
 PROVERS=`proofmgr -show-provers | grep "\[.*\]" | sed -e "s/.*\[.\(.*\)\].*/\1/"`
 for i in $PROVERS ; do
-  proofmgr -run -i $i -c $CAT -cores $CORES -timeout $TIMEOUT
+  proofmgr -run -i $i -cores $CORES -c $CAT -timeout $TIMEOUT
 done
 

### 2010-04-14 whytools/proveit.sh
fixed proof script

@@ -7,15 +7,17 @@ while getopts 't:c:' OPTION
 do
   case $OPTION in
     t) TIMEOUT=$OPTARG ;;
-    c) CORES=$OPTARG ;;
+    c) CORES=$OPTARG;;
     ?) exit 2;;
   esac
 done
+
+shift $OPTIND
 
 CAT=`date +%F+%X`
 proofmgr -add-bench -c $CAT $*
 PROVERS=`proofmgr -show-provers | grep "\[.*\]" | sed -e "s/.*\[.\(.*\)\].*/\1/"`
 for i in $PROVERS ; do
-  proofmgr -run -i $i -cores $CORES -c $CAT -timeout $TIMEOUT
+  proofmgr -run -i $i -c $CAT -timeout $TIMEOUT
 done
 

### 2010-04-14 whytools/proveit.sh
the script always forgot one file

@@ -12,7 +12,7 @@ do
   esac
 done
 
-shift $OPTIND
+shift $(( $(OPTIND) - 1 ))
 
 CAT=`date +%F+%X`
 proofmgr -add-bench -c $CAT $*

### 2010-04-14 whytools/proveit.sh
simplify proveit; getopt just doesnt work

@@ -1,23 +1,15 @@
 #!/bin/bash
-
 set -x
-TIMEOUT=5
-CORES=5
-while getopts 't:c:' OPTION
-do
-  case $OPTION in
-    t) TIMEOUT=$OPTARG ;;
-    c) CORES=$OPTARG;;
-    ?) exit 2;;
-  esac
-done
-
-shift $(( $(OPTIND) - 1 ))
-
+if [ -z "$TIMEOUT" ] ; then
+  TIMEOUT=5
+fi
+if [ -z "$CORES" ] ; then
+  CORES=5
+fi
 CAT=`date +%F+%X`
 proofmgr -add-bench -c $CAT $*
 PROVERS=`proofmgr -show-provers | grep "\[.*\]" | sed -e "s/.*\[.\(.*\)\].*/\1/"`
 for i in $PROVERS ; do
-  proofmgr -run -i $i -c $CAT -timeout $TIMEOUT
+  proofmgr -run -i $i -c $CAT -timeout $TIMEOUT -cores $CORES
 done
 
